---
title: "ARG_3.0"
format: html
---
## setup chunks
```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(DESeq2)
library(apeglm)
library(phyloseq)
library(microbiome)
library(vegan)
library(arrow)
library(VennDiagram)
library(pheatmap)
library(ggrepel)
library(here)
library(grid)
library(rstatix)
library(ggpubr)

# Set figure output directory
if (!dir.exists("../figures")) dir.create("../figures", recursive = TRUE)
if (!dir.exists("../rgiresults")) dir.create("../rgiresults", recursive = TRUE)
```
### color codes and themes
```{r - set figure template}
#| include = FALSE

# Global text size used for ggplot figures in this report
FIG_BASE_SIZE <- 18
PVAL_LABEL_SIZE <- 5

format_p_label <- function(p) {
  dplyr::case_when(
    is.na(p) ~ NA_character_,
    p < 0.001 ~ "p<0.001",
    TRUE ~ paste0("p=", sprintf("%.3f", p))
  )
}

stat_pvalue_manual <- function(data, ..., p_col = NULL, label = NULL, hide.ns = TRUE, size = PVAL_LABEL_SIZE, alpha = 0.05) {
  if (!is.data.frame(data)) {
    return(ggpubr::stat_pvalue_manual(data = data, ..., label = label, hide.ns = hide.ns, size = size))
  }

  # Prefer adjusted p-values when available.
  p_col_candidates <- unique(c(p_col, "p.adj", "p.value", "p"))
  p_col_detected <- p_col_candidates[p_col_candidates %in% names(data)][1]

  if (!is.na(p_col_detected)) {
    data <- data %>%
      dplyr::filter(!is.na(.data[[p_col_detected]]), .data[[p_col_detected]] < alpha)

    # Keep explicit label columns (for example: "label" or "p.adj.formatted").
    use_global_label <- is.null(label) || identical(label, p_col_detected) || label %in% c("p", "p.adj", "p.value")

    if (!identical(label, "p.signif") && use_global_label) {
      data <- data %>%
        dplyr::mutate(.p_label_global = format_p_label(.data[[p_col_detected]]))
      label <- ".p_label_global"
    }
  }

  if (nrow(data) == 0) {
    return(ggplot2::geom_blank())
  }

  ggpubr::stat_pvalue_manual(data = data, ..., label = label, hide.ns = hide.ns, size = size)
}

my_theme <- function() {
  theme_bw(base_family = "DejaVu Sans", 
           base_size = FIG_BASE_SIZE) +
    theme(
      plot.title = element_text(hjust = 0.5),
      axis.title = element_text(size = rel(1.05)),
      axis.text = element_text(size = rel(0.95)),
      legend.title = element_text(size = rel(0.95)),
      legend.text = element_text(size = rel(0.9)),
      strip.text = element_text(size = rel(0.95))
    )
}

library(ggplot2)
theme_set(my_theme())

# Load color codes from YAML file
color_codes <- yaml::read_yaml("../../Metadata/color_codes.yaml")

# Extract specific color palettes
sample_type_colors <- unlist(color_codes$sample_types)
pig_stage_colors <- unlist(color_codes$pig_stages)
category2_colors <- unlist(color_codes$category2)

# publication theme
theme_publication <- function() {
  my_theme()+
    theme(
      legend.position = "none"
    )}

theme_publication_with_title <- function(){
  theme_publication() +
    theme(plot.title = element_text(hjust = 0.5, size = FIG_BASE_SIZE + 2),
          plot.subtitle = element_text(hjust = 0.5, size = FIG_BASE_SIZE))
}
```

## load data

```{r}


# Load feature info and metadata

counts_raw <- read_feather("../data/rgi_combined/All_Mapped_Reads.feather") # loading raw counts

# store as csv
write.csv(counts_raw, "../data/rgi_combined/All_Mapped_Reads.csv", row.names = FALSE)


Depth_raw <- read_feather("../data/rgi_combined/Depth.feather") # loading relative abundance (Depth)

feats <- read_feather("../data/rgi_combined/feature_info.feather") # loading feature info

# Manual curation for AAC(6')-Ib-cr variants:
# 1) ensure Aminoglycosides is present
# 2) remove fluoroquinolone/quinolone antibiotic labels to avoid class splitting
aac_ib_cr_terms <- c(
  "AAC(6')-Ib-cr1",
  "AAC(6')-Ib-cr3",
  "AAC(6')-Ib-cr5",
  "AAC(6')-Ib-cr6",
  "AAC(6')-Ib-cr7"
)

feats <- feats %>%
  mutate(
    ARO_Term = as.character(ARO_Term),
    Drug_Class = as.character(Drug_Class),
    Drug_Class = case_when(
      ARO_Term %in% aac_ib_cr_terms &
        !str_detect(coalesce(Drug_Class, ""), "(^|;\\s*)Aminoglycosides($|;\\s*)") ~
        str_trim(paste0(coalesce(Drug_Class, ""), if_else(coalesce(Drug_Class, "") == "", "", "; "), "Aminoglycosides")),
      TRUE ~ Drug_Class
    ),
    Drug_Class = case_when(
      ARO_Term %in% aac_ib_cr_terms ~ Drug_Class %>%
        str_remove_all("(?i)(^|;\\s*)(fluoroquinolone antibiotic|quinolone antibiotic)(?=;|$)") %>%
        str_replace_all("\\s*;\\s*", "; ") %>%
        str_replace_all("^;\\s*|\\s*;$", "") %>%
        na_if(""),
      TRUE ~ Drug_Class
    )
  )

feats <- feats %>%
  mutate(Prevalence_Sequence_ID = as.character(Prevalence_Sequence_ID)) %>%
  distinct(Prevalence_Sequence_ID, .keep_all = TRUE) %>%
  column_to_rownames(var = "Prevalence_Sequence_ID")

meta <- read_csv("../../Metadata/metadata_silas.csv") %>% # loading metadata
        filter(Duplicated_Sample == FALSE & id != "s24H1")%>%
        column_to_rownames("id")

table(meta$category)

read_stat <- read_tsv("../data/rgi_map_results/bam_stats.tsv")%>%
  column_to_rownames("sample") # loading read stats (for mapping rate)

# transpose matrix in correct shape (genes as rows and samples as columns)

sample_names <- counts_raw[[1]]

mat <- as.matrix(counts_raw[,-1, drop = FALSE])

mode(mat) <- "numeric" 
rownames(mat) <- sample_names

counts_df <- as.data.frame(t(mat), check.names = FALSE)

# remove ppB-samples
counts_df <- counts_df[, !grepl("ppB", colnames(counts_df))]


# check difference between meta and counts samples
names_meta <- rownames(meta)
names_counts <- colnames(counts_df)

diff_names <- setdiff(names_meta, names_counts)
print(diff_names)

names_counts_raw <- colnames(counts_raw)
if ("s24H1" %in% colnames(counts_raw)) { print("yes")} else {print("no")}


# check All_Mapped_Reads columns
library(arrow)
df <- read_feather("../data/rgi_combined/All_Mapped_Reads.feather")
"s24H1" %in% colnames(df)



```
#### inspect data
```{r}
# genes
cat("Numbers of genes:", ncol(counts_raw), "\n")

cat("Numbers of ARO-Terms:", length(unique(feats$ARO_Term)), "\n")

# cat("Numbers of Drug Classes:", length(unique(cpm_drugclass_long$Drug_Class)), "\n")

res_mech <- feats %>%
  pull(Resistance_Mechanism) %>%
  unique()


print(res_mech)
cat("Numbers of Resistance Mechanisms:", length(unique(feats$Resistance_Mechanism)), "\n")
```


### library size

```{r}
sample_depths <- colSums(counts_df)
sort(sample_depths) %>% head(10)  # check the lowest samples

# visualize
data.frame(
  sample = names(sample_depths),
  depth  = sample_depths
) %>%
  left_join(meta %>% rownames_to_column("sample"), by = "sample") %>%
  ggplot(aes(x = reorder(sample, depth), y = depth, fill = category)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = min(sample_depths), color = "red", linetype = "dashed") +
  scale_fill_manual(values = sample_type_colors) +
  labs(title = "ARG library depth per sample",
       x = "Sample", y = "Total ARG reads") +
  theme_publication() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6))

#ggsave
ggsave("../figures/ARG_library_depth_per_sample.svg", width = 10, height = 6, web_fonts = "DejaVu Sans")

## boxplot
data.frame(
  sample = names(sample_depths),
  depth  = sample_depths
) %>%
  left_join(meta %>% rownames_to_column("sample"), by = "sample") %>%
  ggplot(aes(x = category, y = depth, fill = category)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_fill_manual(values = sample_type_colors) +
  labs(title = "ARG library depth by category",
       x = "Category", y = "Total ARG reads") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))+
  theme_publication()

## ggsave
ggsave("../figures/ARG_library_depth_by_category.svg", width = 8, height = 6, web_fonts = "DejaVu Sans")
```

### Normalize counts to relative abundance (Depth) for diversity analyses

We use the following formula: $$ \text{norm coverage} = \frac{\text{N mapped reads}}{L - r + 1} $$ - N = number of reads mapped to the gene (from counts) - L = gene length (from feats) - r = average read length (from read_stat)

```{r}
# make reference length numeric (handle list columns from feather)
feats$Reference_Length <- as.numeric(feats$Reference_Length)

# Align to counts
feats_aligned <- feats[rownames(counts_df), , drop = FALSE]

# Aling read_stats
read_stat_aligned <- read_stat[colnames(counts_df), , drop = FALSE]

# Directly pull the numeric vector
L <- as.numeric(feats_aligned$Reference_Length)


counts_mat <- as.matrix(counts_df)

length(L); sum(is.na(L)) # should be n_genes and 0

# read lengths aligned to the count matrix columns
r <- as.numeric(read_stat_aligned$avg_mapped_length)

bad_r_idx <- which(is.na(r))
length(bad_r_idx)
colnames(counts_df)[bad_r_idx]
setdiff(colnames(counts_df), rownames(read_stat))
read_stat_aligned[bad_r_idx, "avg_mapped_length", drop = FALSE]

# fail early if lengths are missing
if (anyNA(L) || anyNA(r)) {
  stop("Missing gene or read lengths after alignment. Check read_stat and feats rownames.")
}

# build denominator matrix: genes x samples
denom <- outer(L, r, FUN = function(Li, rj) Li - rj + 1)
denom[denom < 1] <- 1   # avoid 0 or negative

# Sanity check
cat("Denom range: [", min(denom), ",", max(denom), "]\n")
cat("Any zeros:", any(denom == 0), "\n")


# normalized depth: reads per possible start site effective-length corrected
counts_norm <- counts_mat / denom


### --- normalize against ARG-mapping reads ---
# counts per millions
cpm <- sweep(counts_norm, MARGIN = 2, STATS = colSums(counts_norm), FUN = "/") * 1e6


# sanity check if col sums sum up to 1e6 before filtering
stopifnot(all(abs(colSums(cpm) - 1e6) < 1e-6))
cat("✓ cpm columns sum to 1e6 before filter: OK\n")

# Global CPM filter: set detections below 1 CPM to zero
cpm_unfiltered <- cpm
n_below_1 <- sum(cpm < 1, na.rm = TRUE)
cpm[cpm < 1] <- 0
cat("Global CPM filter applied (threshold = 1). Entries set to 0:", n_below_1, "\n")


head(cpm)

```

sanity check

```{r}
boxplot(r ~ meta$category)
```

### Phyloseq-Object

```{r}
# Phyloseq with RAW COUNTS
otu_raw <- otu_table(as.matrix(counts_df), taxa_are_rows = TRUE)
sam_data <- sample_data(meta)
physeq_raw <- phyloseq(otu_raw, sam_data)


# Phyloseq with normalized Depth
otu_norm <- otu_table(as.matrix(cpm), taxa_are_rows = TRUE)
physeq_norm <- phyloseq(otu_norm, sam_data)

# save phyesq_norm
saveRDS(physeq_norm, "../results/physeq_normalized.rds")
```

### Calculate mapping rate to CARD database

wie viel Reads mappend zu der card database im vergleich zu den total reads per sample?

```{r}
mapping_stats <- read_stat %>%
  rownames_to_column("sample") %>%
  right_join(meta %>% rownames_to_column("sample"), by = "sample") %>%
  mutate(mapping_rate = mapped / total_filtered)

# Summary by category
mapping_summary <- mapping_stats %>%
  group_by(category) %>%
  summarise(
    mean_mapping_rate = mean(mapping_rate, na.rm = TRUE),
    sd_mapping_rate = sd(mapping_rate, na.rm = TRUE),
    median_mapping_rate = median(mapping_rate, na.rm = TRUE),
    .groups = "drop"
  )

cat("=== Mapping Statistics ===\n")
cat("Reads mapped to CARD database by sample category:\n")
print(mapping_summary) 

# add statistics: Kruskal-Wallis test
mapping_stats$category <- as.factor(mapping_stats$category)
mapping_kw <- mapping_stats %>%
  rstatix::kruskal_test(mapping_rate ~ category)
cat("Kruskal-Wallis test for mapping rate differences across categories:\n")
print(mapping_kw)

# pairwise Wilcoxon tests with BH correction
mapping_pairwise <- mapping_stats %>%
  pairwise_wilcox_test(mapping_rate ~ category, p.adjust.method = "BH") %>%
  add_significance()


# add y-position for plotting
mapping_pairwise <- mapping_pairwise %>%
  mutate(
    y.position = seq(
      from = max(mapping_stats$mapping_rate, na.rm = TRUE) * 1.05,
      to   = max(mapping_stats$mapping_rate, na.rm = TRUE) * 1.25,
      length.out = n()
    )
  )%>%
  ungroup()

# Plot mapping rate by category
p_mapping <- ggplot(mapping_stats, aes(x = category, y = mapping_rate, fill = category)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_fill_manual(values = sample_type_colors) +
  my_theme() +
  labs(
    title = "CARD Database Mapping Rate",
    subtitle = "Percentage of reads mapped to CARD antibiotic resistance genes",
    x = "",
    y = "Mapping Rate"
  ) +
  theme(legend.position = "none")+
  stat_pvalue_manual(
    mapping_pairwise,
    label = "p.adj",
    tip.length = 0.01,
    size = 5,
    family = "DejaVu Sans"
  )+
  theme_publication()+
  scale_x_discrete(labels = color_codes$label_mapping)

print(p_mapping)

ggsave("../figures/mapping_rate_to_CARD.svg", p_mapping, width = 8, height = 6, web_fonts = "DejaVu Sans")
```

## Richness & Prevlance Analysis

### Richness

```{r}
# ARG Richness (cpm > 0: presence/absence)
richness_data <- cpm
richness_data[richness_data > 0] <- 1
richness_df <- data.frame(
  sample = colnames(richness_data),
  argrichness = colSums(richness_data)
) %>% 
  right_join(meta %>% rownames_to_column("sample"), by = "sample")


# Global Kruskal-Wallis test

# richness as factor
richness_df$category <- as.factor(richness_df$category)

# global Kurskal-Wallis test
stats_richness <- richness_df %>%
  rstatix::kruskal_test(argrichness ~ category)

stats_richness
cat("=== The global test has a significance level of p =", stats_richness$p, "indicating that there are significant differences in ARG richness across the categories.\n")

# pairwise Wilcoxon tests with Benjamini-Hochberg correction
pairwise_richness <- richness_df %>%
  pairwise_wilcox_test(argrichness ~ category, p.adjust.method = "BH")%>%
  add_significance() %>%
  mutate(
    # define label position per comparison
    y.position = seq(
      from = max(richness_df$argrichness, na.rm = T) * 1.05,
      to = max(richness_df$argrichness, na.rm =T)*1.25,
      length.out = n()
      )
    )



# Plot richness
p_rich <- ggplot(richness_df, aes(x = category, y = argrichness, fill = category)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_fill_manual(values = color_codes$sample_types) +
  my_theme()+
  labs(
    title = "ARG Richness by Category")+
  ylab("Richness")+
  xlab("")+
  stat_pvalue_manual(
    pairwise_richness,
    label = "p.adj",
    tip.length = 0.01,
    size = 4,
    family = "DejaVu Sans"
  )+
  theme_publication()
print(p_rich)

```

save this plot

```{r}
ggsave("../figures/ARG_richness_by_category.svg", p_rich, width = 8, height = 6, web_fonts = "DejaVu Sans")
```

### sensitivity 

```{r}
# Sensitivity of richness to CPM abundance/prevalence filters.
# x = minimum CPM per sample for presence; y = minimum prevalence across all samples.
min_cpm_grid <- c(0.1, 1, 5)
min_prev_grid <- c(0.01, 0.05, 0.10, 0.20)

sensitivity_tbl <- purrr::map_dfr(min_cpm_grid, function(min_cpm_i) {
  # Presence/absence under current CPM threshold.
  pa_mat <- (cpm >= min_cpm_i) * 1

  purrr::map_dfr(min_prev_grid, function(min_prev_i) {
    keep_features <- rowMeans(pa_mat, na.rm = TRUE) >= min_prev_i

    pa_filt <- pa_mat[keep_features, , drop = FALSE]
    richness_i <- colSums(pa_filt, na.rm = TRUE)

    richness_df_i <- tibble(
      sample = names(richness_i),
      argrichness = as.numeric(richness_i)
    ) %>%
      left_join(meta %>% rownames_to_column("sample"), by = "sample") %>%
      left_join(
        data.frame(sample = rownames(read_stat), depth = read_stat$total_filtered),
        by = "sample"
      ) %>%
      filter(!is.na(category), !is.na(depth))

    lm_i <- lm(argrichness ~ category + depth, data = richness_df_i)
    anova_i <- anova(lm_i)

    tibble(
      min_cpm = min_cpm_i,
      min_prevalence = min_prev_i,
      n_features_kept = sum(keep_features),
      median_richness = median(richness_df_i$argrichness, na.rm = TRUE),
      p_category_depth_adjusted = anova_i["category", "Pr(>F)"]
    )
  })
}) %>%
  mutate(
    min_prevalence_pct = paste0(min_prevalence * 100, "%"),
    signif = case_when(
      is.na(p_category_depth_adjusted) ~ "NA",
      p_category_depth_adjusted < 0.001 ~ "***",
      p_category_depth_adjusted < 0.01 ~ "**",
      p_category_depth_adjusted < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

print(sensitivity_tbl)
write_csv(sensitivity_tbl, "../results/richness_sensitivity_cpm_prevalence.csv")

p_sensitivity <- ggplot(
  sensitivity_tbl,
  aes(x = factor(min_cpm), y = min_prevalence_pct, fill = median_richness)
) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = signif), size = 4, family = "DejaVu Sans") +
  scale_fill_viridis_c(option = "C", name = "Median richness") +
  labs(
    title = "Richness sensitivity to CPM/prevalence filters",
    subtitle = "Stars show depth-adjusted category effect (lm: richness ~ category + depth)",
    x = "Minimum CPM threshold (x)",
    y = "Minimum prevalence threshold (y)"
  ) +
  my_theme()

print(p_sensitivity)
ggsave(
  "../figures/richness_sensitivity_cpm_prevalence.svg",
  p_sensitivity,
  width = 8,
  height = 5,
  web_fonts = "DejaVu Sans"
)
```

#### zero-check in controls

```{r}
# Check whether controls are dominated by zeros in CPM profiles.
sample_zero_tbl <- tibble(
  sample = colnames(cpm),
  zero_fraction = colMeans(cpm == 0, na.rm = TRUE),
  nonzero_fraction = colMeans(cpm > 0, na.rm = TRUE)
) %>%
  left_join(meta %>% rownames_to_column("sample") %>% select(sample, category), by = "sample")

control_level <- "Non_farmer_controls"

controls_zero_tbl <- sample_zero_tbl %>%
  filter(category == control_level)

cat("Control samples:", nrow(controls_zero_tbl), "\n")
cat("Mean zero fraction in controls:", round(mean(controls_zero_tbl$zero_fraction, na.rm = TRUE), 4), "\n")
cat("Median zero fraction in controls:", round(median(controls_zero_tbl$zero_fraction, na.rm = TRUE), 4), "\n")
cat("Range zero fraction in controls:",
    round(min(controls_zero_tbl$zero_fraction, na.rm = TRUE), 4), "-",
    round(max(controls_zero_tbl$zero_fraction, na.rm = TRUE), 4), "\n")

# Optional context: controls vs other categories.
zero_by_category <- sample_zero_tbl %>%
  group_by(category) %>%
  summarise(
    n_samples = n(),
    mean_zero_fraction = mean(zero_fraction, na.rm = TRUE),
    median_zero_fraction = median(zero_fraction, na.rm = TRUE),
    .groups = "drop"
  )

print(zero_by_category)

# Pairwise Wilcoxon tests on sample-level zero fractions (BH-adjusted p-values).
zero_wilcox <- sample_zero_tbl %>%
  filter(!is.na(category)) %>%
  pairwise_wilcox_test(zero_fraction ~ category, p.adjust.method = "BH") %>%
  add_significance() %>%
  mutate(
    y.position = seq(
      from = max(sample_zero_tbl$zero_fraction, na.rm = TRUE) * 1.03,
      to = max(sample_zero_tbl$zero_fraction, na.rm = TRUE) * 1.12,
      length.out = n()
    )
  )

print(zero_wilcox)

zero_plot <- ggplot(sample_zero_tbl, aes(x = category, y = zero_fraction, fill = category)) +
  geom_boxplot(alpha = 0.75, outlier.shape = NA) +
  geom_jitter(width = 0.12, alpha = 0.45, size = 1.4) +
  scale_fill_manual(values = sample_type_colors) +
  labs(
    title = "Fraction of zero CPM entries per sample",
    subtitle = "Higher values indicate more sparse sample profiles",
    x = "",
    y = "Zero fraction"
  ) +
  stat_pvalue_manual(
    zero_wilcox,
    label = "p.adj",
    tip.length = 0.01,
    size = 3.5,
    family = "DejaVu Sans"
  ) +
  my_theme() +
  theme_publication()

print(zero_plot)

# Feature-level check inside controls: how often each feature is zero across control samples.
control_samples <- rownames(meta)[meta$category == control_level]
control_samples <- intersect(control_samples, colnames(cpm))

feature_zero_controls <- tibble(
  feature = rownames(cpm),
  zero_prevalence_controls = rowMeans(cpm[, control_samples, drop = FALSE] == 0, na.rm = TRUE)
)

cat("Features always zero in controls:",
    sum(feature_zero_controls$zero_prevalence_controls == 1),
    "of", nrow(feature_zero_controls), "\n")

write_csv(sample_zero_tbl, "../results/sample_zero_fraction_by_category.csv")
write_csv(zero_wilcox, "../results/sample_zero_fraction_pairwise_wilcox_BH.csv")
write_csv(feature_zero_controls, "../results/feature_zero_prevalence_in_controls.csv")

ggsave(
  "../figures/sample_zero_fraction_by_category.svg",
  zero_plot,
  width = 8,
  height = 5,
  web_fonts = "DejaVu Sans"
)
```



### is richness correlated with depth ?

```{r}
# Is richness correlated with depth?
richness_df <- richness_df %>%
  left_join(
    data.frame(sample = rownames(read_stat), depth = read_stat$total_filtered),
    by = "sample"
  )

cor.test(richness_df$argrichness, richness_df$depth, method = "spearman")

# Visualize
ggplot(richness_df, aes(x = depth, y = argrichness, color = category)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = sample_type_colors) +
  scale_x_log10() +
  my_theme() +
  labs(title = "ARG Richness vs Sequencing Depth",
       x = "Total reads (log scale)", y = "ARG Richness")

# ggsave
ggsave("../figures/ARG_richness_vs_depth.svg", width = 8, height = 6, web_fonts = "DejaVu Sans")
```

#### richness depth corrected

```{r}
library(coin)  # or use a linear model approach
library(car)
library(emmeans)

# Linear model: richness ~ category + depth
lm_richness <- lm(argrichness ~ category + depth, data = richness_df)
summary(lm_richness)

# Are category effects still significant after controlling for depth? - global F-test
anova(lm_richness)


# estimated marginal means (adjusted for depth)
emm <- emmeans(lm_richness, ~ category)
emm

# pairwise post-hoc tests
posthoc <- pairs(emm, adjust = "tukey") # Tukey-adjusted p-values

print(posthoc)


# p-values df of posthoc test
posthoc_df <- as.data.frame(posthoc) %>%
  mutate(
    p.signif = ifelse(p.value < 0.001, "***",
               ifelse(p.value < 0.01,  "**",
               ifelse(p.value < 0.05,  "*", "ns")))
  )



# Residuals nur für Visualisierung
richness_df$richness_resid <- residuals(lm(argrichness ~ depth, data = richness_df))

# Annotationen manuell erstellen
annotations <- posthoc_df %>%
  separate(contrast, into = c("group1", "group2"), sep = " - ") %>%
  mutate(
    group1 = trimws(group1),
    group2 = trimws(group2),
    y.position = c(
      max(richness_df$argrichness, na.rm = TRUE) * 1.10,
      max(richness_df$argrichness, na.rm = TRUE) * 1.20,
      max(richness_df$argrichness, na.rm = TRUE) * 1.30
    )
  )

annotations <- annotations %>%
  mutate(label = ifelse(p.value < 0.001, 
                        "p < 0.001",
                        paste0("p = ", round(p.value, 3))))

# Plot
p_rich_corrected <- ggplot(richness_df, 
                           aes(x = category, y = argrichness, fill = category)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_fill_manual(values = sample_type_colors) +
  my_theme() +
  labs(
    title    = "ARG Richness by Category",
    subtitle = "p-values from lm(richness ~ category + depth), Tukey-adjusted",
    x = "",
    y = "ARG Richness"
  ) +
  theme(legend.position = "none") +
  stat_pvalue_manual(
    annotations,
    label      = "label",  # exact p-values
    tip.length = 0.01,
    size       = 3.5,
    family = "DejaVu Sans"
    
  )+
  theme_publication()

print(p_rich_corrected)

# ggsave
ggsave("../figures/ARG_richness_by_category_depth_corrected.svg", p_rich_corrected, width = 8, height = 6, web_fonts = "DejaVu Sans")


## richness depth corrected per growth stage 



df_richness_stage <- richness_df %>%
  filter(!is.na(growth_stage))

# add statistics
# Linear model: richness ~ growth_stage + depth
lm_richness_stage <- lm(argrichness ~ growth_stage + depth, data = df_richness_stage)
summary(lm_richness_stage)

# Are growth_stage effects still significant after controlling for depth? - global F-test
anova(lm_richness_stage)


# estimated marginal means (adjusted for depth)
emm_stage <- emmeans(lm_richness_stage, ~ growth_stage)
emm_stage
# pairwise post-hoc tests
posthoc_stage <- pairs(emm_stage, adjust = "tukey") # Tukey
posthoc_stage

# p-values df of posthoc test
posthoc_stage_df <- as.data.frame(posthoc_stage) %>%
  mutate(
    p.signif = ifelse(p.value < 0.001, "***",
               ifelse(p.value < 0.01,  "**",
               ifelse(p.value < 0.05,  "*", "ns")))
  ) %>%
  separate(contrast, into = c("group1", "group2"), sep = " - ") %>%
  mutate(
    group1 = trimws(group1),
    group2 = trimws(group2),
    y.position = c(
      max(df_richness_stage$argrichness, na.rm = TRUE) * 1.10,
      max(df_richness_stage$argrichness, na.rm = TRUE) * 1.20,
      max(df_richness_stage$argrichness, na.rm = TRUE) * 1.30
    ),
    label = ifelse(p.value < 0.001, 
                   "p < 0.001",
                   paste0("p = ", round(p.value, 3)))
  )

# order factor to suckling -> weaning -> fattening
df_richness_stage$growth_stage <- factor(df_richness_stage$growth_stage,
                                         levels = c("Suckling", "Weaning", "Fattening"))


p_rich_corrected_stage <- ggplot(df_richness_stage, 
                           aes(x = growth_stage, y = argrichness, fill = growth_stage)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_fill_manual(values = pig_stage_colors) +
  my_theme() +
  labs(
    title    = "ARG Richness by Category",
    subtitle = "p-values from lm(richness ~ category + depth), Tukey-adjusted",
    x = "",
    y = "ARG Richness"
  ) +
  theme_publication()+
  stat_pvalue_manual(
    posthoc_stage_df,
    label      = "label",  # significance stars
    tip.length = 0.01,
    size       = 3.5,
    family = "DejaVu Sans"
  )

print(p_rich_corrected_stage)

# ggsave
ggsave("../figures/ARG_richness_by_stage_depth_corrected.svg", p_rich_corrected_stage, width = 8, height = 6, web_fonts = "DejaVu Sans")


```

### Observed
```{r}
observed <- estimate_richness(physeq_raw, measures = "Observed") %>%
  rownames_to_column("sample") %>%
  left_join(meta %>% rownames_to_column("sample"), by = "sample") %>%
  mutate(depth = sample_sums(physeq_norm)[sample])

library(rstatix)
# add statistics for observed richness
observed$category <- as.factor(observed$category)
stats_observed <- observed %>%
  rstatix::kruskal_test(Observed ~ category)

stats_observed
cat("=== The global test for Observed richness has a significance level of p =", stats_observed$p)

# pairwise Wilcoxon tests with Benjamini-Hochberg correction

pairwise_observed <- observed %>%
  pairwise_wilcox_test(Observed ~ category, p.adjust.method = "BH")%>%
  add_significance() %>%
  mutate(
    y.position = seq(
      from = max(observed$Observed, na.rm = TRUE) * 1.05,
      to = max(observed$Observed, na.rm = TRUE) * 1.25,
      length.out = n()
    )
  )

p_observed <- ggplot(data = observed, aes(x = category, y = Observed, fill = category)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_fill_manual(values = sample_type_colors) +
  labs(title = "Observed Richness by Category",
       x = "", y = "Observed Richness") +
  my_theme()+
  theme(legend.position = "none") +
  stat_pvalue_manual(
    pairwise_observed,
    label = "p.adj",
    tip.length = 0.01,
    size = 4,
    family = "DejaVu Sans"
  )


print(p_observed)

```


### Prevalence

```{r}
# Prevalence Venn diagram (1% threshold)

cat("counts_df dims:", nrow(counts_df), "genes x", ncol(counts_df), "samples\n")
prevalence <- counts_df %>%
  mutate(across(where(is.numeric), ~ .x > 0)) %>%
  select(where(is.logical)) %>%
  rownames_to_column("feature") %>%                      # features (genes/taxa)
  pivot_longer(-feature, names_to = "sample", values_to = "present") %>%
  left_join(meta %>% rownames_to_column("sample"), by = "sample") %>%
  group_by(feature, category) %>%
  summarise(prevalence = mean(present, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = category, values_from = prevalence) %>%
  column_to_rownames("feature")

head(prevalence)



# Create gene sets based on 1% prevalence threshold
gene_sets <- list(
  Pigs = rownames(prevalence)[prevalence$Pigs > 0.01],
  PigFarmers = rownames(prevalence)[prevalence$Pig_farmers > 0.01],
  Non_Farmer_Controls = rownames(prevalence)[prevalence$Non_farmer_controls > 0.01 ]
)

# Create Venn diagram
venn_plot <- venn.diagram(
  x = gene_sets,
  category.names = c("Pig", "PigFarmers", "Non_Farmer_Controls"),
  filename = NULL,
  output = TRUE,
  col = "transparent",
  fill = sample_type_colors,
  alpha = 0.50,
  cex = 1.5,
  cat.cex = 1.5,
  fontfamily = "DeJaVu Sans",
  cat.fontfamily = "DejaVu Sans",
  main.fontfamily = "DejaVu Sans"
  )
grid.newpage(); grid.draw(venn_plot)

# Save the plot
svg("../figures/Venn_Unigenes_with_AR_annotation_R.svg", width = 8, height = 6)
grid.draw(venn_plot)
dev.off()

# plot with eulerr:
library(eulerr)

fit <- euler(gene_sets)

nice_labels <- c(
  color_codes$label_mapping[["Pigs"]],
  color_codes$label_mapping[["Pig_farmers"]],
  color_codes$label_mapping[["Non_farmer_controls"]]
)

p_euler <- plot(
  fit,
  quantities = list(type = "counts", fontsize = 14),
  labels = list(labels = nice_labels,font = 2, fontsize = 14),
  fills = list(
    fill = c(
      sample_type_colors[["Pigs"]],
      sample_type_colors[["Pig_farmers"]],
      sample_type_colors[["Non_farmer_controls"]]
    ),
    alpha = 0.5
  ),
  edges = list(col = "transparent")
)

print(p_euler)

ggsave(
  "../figures/Venn_Unigenes_with_AR_annotation_euler.svg",
  p_euler,
  width = 8, height = 6
)


```

## Diversity-analysis

### alpha diversity (depth corrected)

```{r}
# Estimate alpha diversity from normalized phyloseq object ────────────────
alpha_df <- estimate_richness(physeq_norm, measures = c("Shannon", "Simpson")) %>%
  rownames_to_column("sample") %>%
  left_join(meta %>% rownames_to_column("sample"), by = "sample") %>%
  mutate(depth = sample_sums(physeq_norm)[sample])

# Add depth to alpha_df
alpha_df <- alpha_df %>%
  mutate(depth = sample_sums(physeq_norm)[sample])  # or from meta directly

# Loop over both measures
library(purrr)

alpha_models <- c("Shannon", "Simpson") %>%
  set_names() %>%
  map(~ {
    formula <- as.formula(paste(.x, "~ category + depth"))
    lm(formula, data = alpha_df)
  })

# EMMs and post-hoc per measure
alpha_emm <- map(alpha_models, ~ emmeans(.x, ~ category))
alpha_posthoc <- map(alpha_emm, ~ pairs(.x, adjust = "tukey"))

# Prepare post-hoc results for annotation
posthoc_df <- map2_dfr(alpha_posthoc, names(alpha_posthoc), ~ {
  as.data.frame(.x) %>%
    mutate(
      measure  = .y,
      p.signif = ifelse(p.value < 0.001, "***",
                 ifelse(p.value < 0.01,  "**",
                 ifelse(p.value < 0.05,  "*", "ns")))
    )
})

# separate contrast into group1 / group2
posthoc_df <- posthoc_df %>%
  separate(contrast, into = c("group1", "group2"), sep = " - ") %>%
  mutate(
    group1 = trimws(group1),
    group2 = trimws(group2)
  )

# y.position based on RAW values (not residuals!)
posthoc_df <- posthoc_df %>%
  group_by(measure) %>%
  mutate(
    y.position = seq(
      from = max(alpha_df[[unique(measure)]], na.rm = TRUE) * 1.05,  # ← raw
      to   = max(alpha_df[[unique(measure)]], na.rm = TRUE) * 1.25,
      length.out = n()
    )
  ) %>%
  ungroup()

# pivot RAW values to long format
alpha_long <- alpha_df %>%
  select(sample, category, Shannon, Simpson) %>%  
  pivot_longer(
    cols      = c(Shannon, Simpson),
    names_to  = "measure",
    values_to = "value"
  )

posthoc_df$label <- scales::pvalue(posthoc_df$p.value, add_p = TRUE)

# Plot with RAW values + Tukey-adjusted p-values from lm()
p_alpha <- ggplot(alpha_long, aes(x = category, y = value, fill = category)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_fill_manual(values = sample_type_colors) +
  facet_wrap(~ measure, scales = "free_y") +
  stat_pvalue_manual(
    posthoc_df,
    label      = "label",
    tip.length = 0.01,
    size       = 4,
    family     = "DejaVu Sans"
  ) +
  labs(
    title    = "Alpha Diversity by Category",
    subtitle = "p-values from lm(diversity ~ category + depth), Tukey-adjusted",
    y        = "Diversity index",
    x = " "
  ) +
  scale_x_discrete(labels = color_codes$label_mapping) +
  my_theme() +
  theme(
    #axis.text.x    = element_text(angle = 0, hjust = 1),
    legend.position = "none"
  )+
  theme_publication()

print(p_alpha)

ggsave("../figures/alpha_diversity_ARG_counts_depth_corrected.svg", p_alpha, width = 10, height = 6, web_fonts = "DejaVu Sans")
```

### alpha diversity 5% prevalence filter

```{r}
# Sensitivity analysis: global 5% prevalence filter on physeq_norm.
# Feature is kept if CPM > 1000 in at least 5% of all samples.
otu_norm_mat <- as(otu_table(physeq_norm), "matrix")

if (!taxa_are_rows(physeq_norm)) {
  otu_norm_mat <- t(otu_norm_mat)
}

tax_prev_global <- rowMeans(otu_norm_mat > 1000, na.rm = TRUE)
keep_taxa_5pct <- names(tax_prev_global)[tax_prev_global >= 0.05]

physeq_norm_prev5 <- prune_taxa(keep_taxa_5pct, physeq_norm)

cat("Taxa kept after 5% global prevalence filter:", length(keep_taxa_5pct),
    "of", ntaxa(physeq_norm), "\n")

alpha_df_prev5 <- estimate_richness(physeq_norm_prev5, measures = c("Shannon", "Simpson")) %>%
  rownames_to_column("sample") %>%
  left_join(meta %>% rownames_to_column("sample"), by = "sample") %>%
  mutate(depth = sample_sums(physeq_norm_prev5)[sample])

alpha_models_prev5 <- c("Shannon", "Simpson") %>%
  set_names() %>%
  map(~ {
    formula <- as.formula(paste(.x, "~ category + depth"))
    lm(formula, data = alpha_df_prev5)
  })

alpha_emm_prev5 <- map(alpha_models_prev5, ~ emmeans(.x, ~ category))
alpha_posthoc_prev5 <- map(alpha_emm_prev5, ~ pairs(.x, adjust = "tukey"))

posthoc_df_prev5 <- map2_dfr(alpha_posthoc_prev5, names(alpha_posthoc_prev5), ~ {
  as.data.frame(.x) %>%
    mutate(measure = .y)
}) %>%
  separate(contrast, into = c("group1", "group2"), sep = " - ") %>%
  mutate(
    group1 = trimws(group1),
    group2 = trimws(group2)
  )

posthoc_df_prev5 <- posthoc_df_prev5 %>%
  group_by(measure) %>%
  mutate(
    y.position = seq(
      from = max(alpha_df_prev5[[unique(measure)]], na.rm = TRUE) * 1.05,
      to = max(alpha_df_prev5[[unique(measure)]], na.rm = TRUE) * 1.25,
      length.out = n()
    )
  ) %>%
  ungroup()

posthoc_df_prev5$label <- scales::pvalue(posthoc_df_prev5$p.value, add_p = TRUE)

alpha_long_prev5 <- alpha_df_prev5 %>%
  select(sample, category, Shannon, Simpson) %>%
  pivot_longer(
    cols = c(Shannon, Simpson),
    names_to = "measure",
    values_to = "value"
  )

p_alpha_prev5 <- ggplot(alpha_long_prev5, aes(x = category, y = value, fill = category)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_fill_manual(values = sample_type_colors) +
  facet_wrap(~ measure, scales = "free_y") +
  stat_pvalue_manual(
    posthoc_df_prev5,
    label = "label",
    tip.length = 0.01,
    size = 4,
    family = "DejaVu Sans"
  ) +
  labs(
    title = "Alpha Diversity by Category (5% global prevalence filter)",
    subtitle = "p-values from lm(diversity ~ category + depth), Tukey-adjusted",
    y = "Diversity index",
    x = " "
  ) +
  scale_x_discrete(labels = color_codes$label_mapping) +
  my_theme() +
  theme(legend.position = "none") +
  theme_publication()

print(p_alpha_prev5)

write_csv(alpha_df_prev5, "../results/alpha_diversity_prev5_values.csv")
write_csv(posthoc_df_prev5, "../results/alpha_diversity_prev5_posthoc_tukey.csv")

ggsave(
  "../figures/alpha_diversity_ARG_counts_depth_corrected_prev5.svg",
  p_alpha_prev5,
  width = 10,
  height = 6,
  web_fonts = "DejaVu Sans"
)
```


### beta diversity

```{r}

cpm_clr <- t(microbiome::transform(cpm, "clr"))   # samples × genes ✓

pca_res <- prcomp(cpm_clr)

df_pca <- as.data.frame(pca_res$x) %>% 
  rownames_to_column("sample") %>%  # rownames ARE samples
  left_join(meta %>% rownames_to_column("sample"), by = "sample")  # Now matches!

# === Statistics ===
dist_clr <- dist(cpm_clr, method = "euclidean")

meta_perm <- meta[rownames(cpm_clr), , drop = F]

adon <- adonis2(dist_clr ~ category2, data = meta_perm, permutations = 999)


adon

# extract statistics

R2 <- adon$R2[1]
p_perm <- adon$`Pr(>F)`[1]

# === Plot ===
# Verify join worked
cat("Samples in PCA:", nrow(df_pca), "\n")
cat("NA category:", sum(is.na(df_pca$category2)), "\n")
print(head(df_pca))


pct_var <- round(100 * summary(pca_res)$importance[2, 1:2], 1)



p_pca <- ggplot(df_pca, aes(PC1, PC2, color = category2)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(type = "norm", level = 0.95, linewidth = 0.7)+
  scale_color_manual(values = category2_colors)+
  my_theme()+
  labs(
  title = "PCA on normalized CLR transformed gene profiles",
    x = paste0("PC1: ", pct_var[1], "% variance"),
    y = paste0("PC2: ", pct_var[2], "% variance"),
    color = "Category"
  )+
  annotate("text", x = Inf, y = Inf, label = paste0("PERMANOVA R² = ", round(R2, 3), "\np = ", round(p_perm, 3)),
           hjust = 1.1, vjust = 1.1, size = 5, family = "DejaVu Sans")+
  theme_publication()+
  theme(legend.position = "right")
print(p_pca)
```

save this plot

```{r}
ggsave("../figures/beta_diversity_ARG_counts_PCA_cat2.svg", p_pca, width = 11, height = 8, web_fonts = "DejaVu Sans")
```

#### Bray-Curtis distance to pigs

Test whether pig farmers are closer to pigs than non-farmer controls are.

```{r}
# Samples x genes matrix for Bray-Curtis
cpm_t <- t(cpm)

# Bray-Curtis distance matrix among samples
bc_dist <- vegan::vegdist(cpm_t, method = "bray")
bc_mat <- as.matrix(bc_dist)

# Robust handling of possible farmer label names
category_levels <- unique(as.character(meta$category))
farmer_level_beta <- if ("Pig_farmers" %in% category_levels) {
  "Pig_farmers"
} else if ("Pig_farmer" %in% category_levels) {
  "Pig_farmer"
} else {
  stop("Neither 'Pig_farmers' nor 'Pig_farmer' found in meta$category")
}

pig_ids <- rownames(meta)[meta$category == "Pigs"]
farmer_ids <- rownames(meta)[meta$category == farmer_level_beta]
control_ids <- rownames(meta)[meta$category == "Non_farmer_controls"]

# Keep only samples present in the distance matrix
pig_ids <- intersect(pig_ids, rownames(bc_mat))
farmer_ids <- intersect(farmer_ids, rownames(bc_mat))
control_ids <- intersect(control_ids, rownames(bc_mat))

if (length(pig_ids) == 0 || length(farmer_ids) == 0 || length(control_ids) == 0) {
  stop("Missing samples for Bray-Curtis distance comparison. Check meta$category labels.")
}

# Per pig sample: mean distance to each comparison group
bc_to_pigs <- tibble::tibble(
  pig_sample = pig_ids,
  mean_to_farmers = purrr::map_dbl(pig_ids, ~ mean(bc_mat[.x, farmer_ids], na.rm = TRUE)),
  mean_to_controls = purrr::map_dbl(pig_ids, ~ mean(bc_mat[.x, control_ids], na.rm = TRUE))
) %>%
  left_join(
    meta %>%
      rownames_to_column("pig_sample") %>%
      select(pig_sample, growth_stage),
    by = "pig_sample"
  )

# Paired test across pig samples
bc_test <- wilcox.test(
  bc_to_pigs$mean_to_farmers,
  bc_to_pigs$mean_to_controls,
  paired = TRUE,
  exact = FALSE
)

p_label <- if (bc_test$p.value < 1e-3) {
  paste0("paired Wilcoxon p = ", format(bc_test$p.value, scientific = TRUE, digits = 2))
} else {
  paste0("paired Wilcoxon p = ", round(bc_test$p.value, 4))
}

plot_bc <- bc_to_pigs %>%
  select(pig_sample, mean_to_farmers, mean_to_controls, growth_stage) %>%
  pivot_longer(
    cols = c(mean_to_farmers, mean_to_controls),
    names_to = "comparison",
    values_to = "distance"
  ) %>%
  dplyr::mutate(
    comparison = dplyr::recode(
      comparison,
      "mean_to_farmers" = "Pig farmers",
      "mean_to_controls" = "Non farmer controls"
    )
  )

y_max <- max(plot_bc$distance, na.rm = TRUE)

p_bc <- ggplot(plot_bc, aes(x = comparison, y = distance, fill = comparison)) +
  geom_boxplot(alpha = 0.75, width = 0.55, outlier.shape = NA) +
  #geom_line(aes(group = pig_sample), colour = "grey70", alpha = 0.6) +
  #geom_point(aes(group = pig_sample), size = 1.6, alpha = 0.75, colour = "grey45") +
  scale_fill_manual(values = c("Pig farmers" = "#E69F00", "Non farmer controls" = "#56B4E9")) +
  geom_jitter(width = 0.1,alpha = 0.4)+
  labs(
    title = "Distance of pig samples to comparison groups",
    subtitle = p_label,
    x = "",
    y = "Mean Bray-Curtis distance to pigs"
  ) +
  annotate("text", x = 1.5, y = y_max * 1.04, label = p_label, family = "DejaVu Sans", size = 5) +
  my_theme() +
  theme_publication()

print(p_bc)

ggsave("../figures/bc_distance_to_pigs.svg", p_bc, width = 6, height = 6, web_fonts = "DejaVu Sans")

cat("N pig samples:", nrow(bc_to_pigs), "\n")
cat("Mean distance to pig farmers:", round(mean(bc_to_pigs$mean_to_farmers), 4), "\n")
cat("Mean distance to non farmer controls:", round(mean(bc_to_pigs$mean_to_controls), 4), "\n")
print(bc_test)
```


#### Bray-Curtis distance from pig farmers to pig stages and controls

For each pig farmer sample, compute mean distance to Suckling/Weaning/Fattening pigs and to controls.

```{r}
# Reuse sample x sample Bray-Curtis matrix from the previous chunk
# (bc_mat computed on t(cpm))

# Robust handling of possible farmer label names
category_levels <- unique(as.character(meta$category))
farmer_level_beta <- if ("Pig_farmers" %in% category_levels) {
  "Pig_farmers"
} else if ("Pig_farmer" %in% category_levels) {
  "Pig_farmer"
} else {
  stop("Neither 'Pig_farmers' nor 'Pig_farmer' found in meta$category")
}

ordered_stages <- c("Suckling", "Weaning", "Fattening")

pig_stage_tbl <- meta %>%
  rownames_to_column("sample") %>%
  filter(category == "Pigs", !is.na(growth_stage), growth_stage != "") %>%
  mutate(growth_stage = as.character(growth_stage))

present_stages <- ordered_stages[ordered_stages %in% unique(pig_stage_tbl$growth_stage)]
if (length(present_stages) < 3) {
  stop("Need all pig stages (Suckling, Weaning, Fattening) in metadata for this comparison.")
}

pig_stage_tbl <- pig_stage_tbl %>%
  filter(growth_stage %in% ordered_stages) %>%
  mutate(growth_stage = factor(growth_stage, levels = ordered_stages))

stage_map <- split(pig_stage_tbl$sample, pig_stage_tbl$growth_stage)
stage_map <- lapply(stage_map, intersect, y = rownames(bc_mat))

farmer_ids <- rownames(meta)[meta$category == farmer_level_beta]
control_ids <- rownames(meta)[meta$category == "Non_farmer_controls"]
pig_ids <- rownames(meta)[meta$category == "Pigs"]

farmer_ids <- intersect(farmer_ids, rownames(bc_mat))
control_ids <- intersect(control_ids, rownames(bc_mat))
pig_ids <- intersect(pig_ids, rownames(bc_mat))

if (length(farmer_ids) == 0 || length(control_ids) == 0 || length(pig_ids) == 0) {
  stop("Missing farmer/control/pig samples in Bray-Curtis matrix.")
}

# One row per pig-farmer sample with mean distance to each reference group
bc_farmer_df <- tibble::tibble(
  farmer_sample = farmer_ids,
  to_suckling = purrr::map_dbl(farmer_ids, ~ mean(bc_mat[.x, stage_map[["Suckling"]]], na.rm = TRUE)),
  to_weaning = purrr::map_dbl(farmer_ids, ~ mean(bc_mat[.x, stage_map[["Weaning"]]], na.rm = TRUE)),
  to_fattening = purrr::map_dbl(farmer_ids, ~ mean(bc_mat[.x, stage_map[["Fattening"]]], na.rm = TRUE)),
  to_controls = purrr::map_dbl(farmer_ids, ~ mean(bc_mat[.x, control_ids], na.rm = TRUE)),
  to_all_pigs = purrr::map_dbl(farmer_ids, ~ mean(bc_mat[.x, pig_ids], na.rm = TRUE))
)

plot_df <- bc_farmer_df %>%
  select(farmer_sample, to_suckling, to_weaning, to_fattening, to_controls) %>%
  pivot_longer(
    cols = c(to_suckling, to_weaning, to_fattening, to_controls),
    names_to = "comparison",
    values_to = "distance"
  ) %>%
  mutate(
    comparison = factor(
      comparison,
      levels = c("to_suckling", "to_weaning", "to_fattening", "to_controls"),
      labels = c("Suckling pigs", "Weaning pigs", "Fattening pigs", "Controls")
    )
  )

# Paired tests: each farmer sample against controls baseline
stage_vs_ctrl_tests <- tibble::tibble(
  stage = c("Suckling", "Weaning", "Fattening"),
  p.value = c(
    wilcox.test(bc_farmer_df$to_suckling, bc_farmer_df$to_controls, paired = TRUE, exact = FALSE)$p.value,
    wilcox.test(bc_farmer_df$to_weaning, bc_farmer_df$to_controls, paired = TRUE, exact = FALSE)$p.value,
    wilcox.test(bc_farmer_df$to_fattening, bc_farmer_df$to_controls, paired = TRUE, exact = FALSE)$p.value
  )
) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH"))

# Overall question: are farmers generally closer to pigs than to controls?
overall_test <- wilcox.test(
  bc_farmer_df$to_all_pigs,
  bc_farmer_df$to_controls,
  paired = TRUE,
  exact = FALSE
)

overall_label <- if (overall_test$p.value < 1e-3) {
  paste0("paired Wilcoxon (all pigs vs controls) p = ", format(overall_test$p.value, scientific = TRUE, digits = 2))
} else {
  paste0("paired Wilcoxon (all pigs vs controls) p = ", round(overall_test$p.value, 5))
}

stage_color_map <- c(
  "Suckling" = if ("Suckling" %in% names(pig_stage_colors)) pig_stage_colors[["Suckling"]] else "#66C2A5",
  "Weaning" = if ("Weaning" %in% names(pig_stage_colors)) pig_stage_colors[["Weaning"]] else "#FC8D62",
  "Fattening" = if ("Fattening" %in% names(pig_stage_colors)) pig_stage_colors[["Fattening"]] else "#8DA0CB"
)

fill_vals <- c(
  "Suckling pigs" = stage_color_map[["Suckling"]],
  "Weaning pigs" = stage_color_map[["Weaning"]],
  "Fattening pigs" = stage_color_map[["Fattening"]],
  "Controls" = "#56B4E9"
)

# Create annotation tibble for stage vs. control pairwise comparisons
bc_pvalue_anno <- tibble(
  group1 = c("Suckling pigs", "Weaning pigs", "Fattening pigs"),
  group2 = c("Controls", "Controls", "Controls"),
  xmin = c(1, 2, 3),  # Suckling, Weaning, Fattening positions
  xmax = c(4, 4, 4),  # Controls position
  p.adj = stage_vs_ctrl_tests$p.adj
) %>%
  mutate(
    p.adj_label = case_when(
      p.adj < 0.001 ~ "<0.001",
      TRUE ~ as.character(round(p.adj, 3))
    )
  )

# Calculate y positions based on data range
y_max <- max(plot_df$distance, na.rm = TRUE)
y_min <- min(plot_df$distance, na.rm = TRUE)
y_range <- y_max - y_min
y_step <- y_range * 0.05

# Add y positions for significance bars (stack vertically)
bc_pvalue_anno <- bc_pvalue_anno %>%
  mutate(
    y = y_max + y_step * (row_number())
  )

p_bc_stage <- ggplot(plot_df, aes(x = comparison, y = distance, fill = comparison)) +
  geom_boxplot(alpha = 0.75, width = 0.6, outlier.shape = NA) +
  #geom_point(aes(group = farmer_sample), size = 1.5, alpha = 0.75, colour = "grey45") +
  geom_jitter(width = 0.15, alpha = 0.4)+
  scale_fill_manual(values = fill_vals) +
  stat_pvalue_manual(
    data = bc_pvalue_anno,
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y",
    label = "p.adj_label",
    inherit.aes = FALSE,
    tip.length = 0.01,
    bracket.size = 0.3,
    size = 5,
    family = "DejaVu Sans"
  ) +
  labs(
    title = "Pig farmers: distance to pig stages and controls",
    subtitle = overall_label,
    x = "",
    y = "Avg. Bray–Curtis distance",
    fill = "Reference group"
  ) +
  theme_publication()

print(p_bc_stage)
ggsave("../figures/bc_distance_from_farmers_to_stages_and_controls.svg", p_bc_stage, width = 9, height = 6, web_fonts = "DejaVu Sans")

cat("N pig-farmer samples:", nrow(bc_farmer_df), "\n")

distance_summary <- bc_farmer_df %>%
  summarise(
    mean_to_suckling = mean(to_suckling, na.rm = TRUE),
    mean_to_weaning = mean(to_weaning, na.rm = TRUE),
    mean_to_fattening = mean(to_fattening, na.rm = TRUE),
    mean_to_controls = mean(to_controls, na.rm = TRUE),
    mean_to_all_pigs = mean(to_all_pigs, na.rm = TRUE)
  )

print(distance_summary)
print(stage_vs_ctrl_tests)
print(overall_test)
```

#### beta diversity - stages

here I want to make the same pca plot as above but use the stages separately - so that we have a cluster of the controls but then also a cluster of all pig farmers and pig stages

```{r}
df_pca_stage <- df_pca %>%
  mutate(
    pca_group = case_when(
      category %in% c("Pigs", "Pig_farmers") & !is.na(growth_stage) & growth_stage != "" ~ as.character(growth_stage),
      TRUE ~ as.character(category)
    )
  )

# Keep category colors for controls/farmers; use pig_stage_colors for pig stages present in data.
stage_levels <- df_pca_stage %>%
  filter(category %in% c("Pigs", "Pig_farmers")) %>%
  pull(growth_stage) %>%
  as.character() %>%
  unique() %>%
  na.omit()

stage_levels <- stage_levels[stage_levels != ""]

pca_group_levels <- c("Non_farmer_controls", "Pig_farmers", stage_levels)
df_pca_stage$pca_group <- factor(df_pca_stage$pca_group, levels = unique(pca_group_levels))

pca_group_colors <- c(
  "Non_farmer_controls" = sample_type_colors[["Non_farmer_controls"]],
  "Pig_farmers" = sample_type_colors[["Pig_farmers"]]
)

stage_color_map <- pig_stage_colors[intersect(names(pig_stage_colors), stage_levels)]
missing_stage_cols <- setdiff(stage_levels, names(stage_color_map))
if (length(missing_stage_cols) > 0) {
  extra_cols <- grDevices::hcl.colors(length(missing_stage_cols), palette = "Dark 3")
  names(extra_cols) <- missing_stage_cols
  stage_color_map <- c(stage_color_map, extra_cols)
}

# === Statistics ===
# PERMANOVA by growth stage (pig samples only, with non-missing stage labels)
meta_stage_perm <- meta_perm %>%
  rownames_to_column("sample") %>%
  filter(category == "Pigs", !is.na(growth_stage), growth_stage != "") %>%
  mutate(growth_stage = factor(growth_stage))

if (n_distinct(meta_stage_perm$growth_stage) >= 2) {
  stage_samples <- meta_stage_perm$sample
  dist_clr_stage <- as.dist(as.matrix(dist_clr)[stage_samples, stage_samples, drop = FALSE])

  adon_growth_stage <- adonis2(
    dist_clr_stage ~ growth_stage,
    data = meta_stage_perm %>% column_to_rownames("sample"),
    permutations = 999
  )

  r2_stage <- adon_growth_stage$R2[1]
  p_stage <- adon_growth_stage$`Pr(>F)`[1]
  permanova_stage_label <- paste0(
    "PERMANOVA (growth stage) R2 = ", round(r2_stage, 3),
    " | p = ", ifelse(p_stage < 0.001, "<0.001", sprintf("%.3f", p_stage))
  )

  write_csv(
    tibble(
      term = "growth_stage",
      R2 = r2_stage,
      p_value = p_stage,
      n_samples = nrow(meta_stage_perm),
      n_levels = n_distinct(meta_stage_perm$growth_stage)
    ),
    "../results/permanova_growth_stage_pigs.csv"
  )

  print(adon_growth_stage)
} else {
  permanova_stage_label <- "PERMANOVA (growth stage): not enough stage levels"
  warning("Not enough growth_stage levels among pig samples for PERMANOVA.")
}


pca_group_colors <- c(pca_group_colors, stage_color_map)

p_pca_2 <- ggplot(df_pca_stage, aes(PC1, PC2, color = pca_group, shape = category)) +
  geom_point(size = 3.2, alpha = 0.85) +
  stat_ellipse(
    data = df_pca_stage,
    aes(PC1, PC2, group = pca_group, color = pca_group),
    inherit.aes = FALSE,
    type = "norm",
    level = 0.95,
    linewidth = 0.7,
    alpha = 0.15
  ) +
  scale_color_manual(values = pca_group_colors, breaks = levels(df_pca_stage$pca_group), drop = FALSE) +
  scale_shape_manual(values = c("Pigs" = 16, "Pig_farmers" = 17, "Non_farmer_controls" = 15), drop = FALSE) +
  labs(
    title = "PCA with stage-aware grouping",
    subtitle = paste(
      "Pigs are colored by growth stage; farmers and controls by category",
      permanova_stage_label,
      sep = "\n"
    ),
    x = paste0("PC1: ", pct_var[1], "% variance"),
    y = paste0("PC2: ", pct_var[2], "% variance"),
    color = "Stage / category",
    shape = "Category"
  ) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = permanova_stage_label,
    hjust = 1.05,
    vjust = 1.5,
    size = 4,
    family = "DejaVu Sans"
  ) +
  my_theme()+
  theme_publication()+
  theme(legend.position = "left")
print(p_pca_2)

#ggsave
ggsave("../figures/PCA_growth_stage.svg", p_pca_2, width = 10, height = 6, web_fonts = "DejaVu Sans")
```

### eveness (Pielou's eveness)

```{r}

cpm_t <- t(cpm)
H <- diversity(cpm_t, index = "shannon") # shannon H
S <- specnumber(cpm_t) # richness S
J <- H / log(S) # Pielou's eveness J

# combine with metadata
eveness_df <- data.frame(sample = rownames(cpm_t), Pielou_J = J) %>%
  left_join(meta %>% rownames_to_column("sample"), by = "sample")

# pairwise Wilcoxon tests for eveness
stat_eveness <- eveness_df %>%
  pairwise_wilcox_test(Pielou_J ~ category, p.adjust.method = "BH") %>%
  add_significance() %>%
  mutate(
    y.position = seq(
      from = max(eveness_df$Pielou_J, na.rm = TRUE) * 1.05,
      to   = max(eveness_df$Pielou_J, na.rm = TRUE) * 1.25,
      length.out = n()
    )
  )

# Plot eveness
p_eveness <- ggplot(eveness_df, aes(x = category, y = Pielou_J, color = category)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(alpha = 0.5, width = 0.2) +
  scale_color_manual(values = sample_type_colors) +
  labs(
    title = "Pielou's Eveness (J)",
    x = "",
    y = "Eveness index"
  ) +
  stat_pvalue_manual(
    stat_eveness,
    label = "p.adj",
    tip.length = 0.01,
    size = 4,
    family = "DejaVu Sans"
  ) +
  my_theme() +
  theme(
    #axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )+
  scale_x_discrete(labels = color_codes$label_mapping)
print(p_eveness)

```

save evenessplot

```{r}
ggsave("../figures/eveness_ARG_counts.svg", p_eveness, width = 6, height =6, web_fonts = "DejaVu Sans")
```

## ARO-Term level

### DESeq2 - aggregated on ARO-Term

#### ARO_Term-level DESeq2 (aggregated counts)

```{r}
aro_map <- feats %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  select(Prevalence_Sequence_ID, ARO_Term = any_of(c("ARO_Term", "AROName", "AROAccession")))

counts_aro_df <- counts_df %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  left_join(aro_map, by = "Prevalence_Sequence_ID") %>%
  mutate(ARO_Term = ifelse(is.na(ARO_Term), Prevalence_Sequence_ID, ARO_Term)) %>%
  select(-Prevalence_Sequence_ID) %>%
  group_by(ARO_Term) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "drop")

counts_aro_mat <- as.matrix(counts_aro_df %>% select(-ARO_Term))
rownames(counts_aro_mat) <- counts_aro_df$ARO_Term
mode(counts_aro_mat) <- "integer"

common_samples <- intersect(rownames(meta), colnames(counts_aro_mat))
meta <- meta[common_samples, , drop = FALSE]
counts_aro_mat <- counts_aro_mat[, common_samples, drop = FALSE]

category_levels <- unique(as.character(meta$category))
farmer_level <- if ("Pig_farmer" %in% category_levels) {
  "Pig_farmer"
} else if ("Pig_farmers" %in% category_levels) {
  "Pig_farmers"
} else {
  stop("Neither 'Pig_farmer' nor 'Pig_farmers' found in meta$category")
}

dds_aro <- DESeqDataSetFromMatrix(countData = counts_aro_mat, colData = meta, design = ~ category)
dds_aro <- estimateSizeFactors(dds_aro, type = "poscounts")
dds_aro <- DESeq(dds_aro)

res_aro_pig_vs_ctrl <- results(dds_aro, contrast = c("category", "Pigs", "Non_farmer_controls")) %>%
  as.data.frame() %>%
  rownames_to_column("ARO_Term")

res_aro_farmer_vs_ctrl <- results(dds_aro, contrast = c("category", farmer_level, "Non_farmer_controls")) %>%
  as.data.frame() %>%
  rownames_to_column("ARO_Term")

res_aro_both <- res_aro_pig_vs_ctrl %>%
  dplyr::rename(
    log2FC_PIG_CTRL = log2FoldChange,
    pvalue_PIG_CTRL = pvalue,
    padj_PIG_CTRL = padj
  ) %>%
  full_join(
    res_aro_farmer_vs_ctrl %>%
      dplyr::rename(
        log2FC_PIGFARMER_CTRL = log2FoldChange,
        pvalue_PIGFARMER_CTRL = pvalue,
        padj_PIGFARMER_CTRL = padj
      ),
    by = "ARO_Term"
  ) %>%
  filter(!is.na(log2FC_PIG_CTRL) & !is.na(log2FC_PIGFARMER_CTRL))

# Backward-compatible normalization in case an older in-memory object still has .x/.y names.
pick_num_col <- function(df, candidates) {
  for (nm in candidates) {
    if (nm %in% names(df)) return(as.numeric(df[[nm]]))
  }
  rep(NA_real_, nrow(df))
}

res_aro_both <- res_aro_both %>%
  mutate(
    pvalue_PIG_CTRL = pick_num_col(., c("pvalue_PIG_CTRL", "pvalue.x", "pvalue")),
    padj_PIG_CTRL = pick_num_col(., c("padj_PIG_CTRL", "padj.x", "padj")),
    pvalue_PIGFARMER_CTRL = pick_num_col(., c("pvalue_PIGFARMER_CTRL", "pvalue.y")),
    padj_PIGFARMER_CTRL = pick_num_col(., c("padj_PIGFARMER_CTRL", "padj.y"))
  )

# add labels of top 20 (top right) ARO_Terms
top20_aro_terms <- res_aro_both %>%
  filter(log2FC_PIG_CTRL > 1, log2FC_PIGFARMER_CTRL > 1) %>%
  arrange(desc(log2FC_PIG_CTRL), desc(log2FC_PIGFARMER_CTRL)) %>%
  slice_head(n = 20) %>%
  pull(ARO_Term)

# Export top 20 annotated ARO terms with DESeq2 statistics from both models.
aro_annotation_tbl <- feats %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  transmute(
    ARO_Term = if_else(
      is.na(ARO_Term) | ARO_Term == "",
      as.character(Prevalence_Sequence_ID),
      as.character(ARO_Term)
    ),
    Drug_Class = as.character(Drug_Class),
    Resistance_Mechanism = as.character(Resistance_Mechanism)
  ) %>%
  group_by(ARO_Term) %>%
  summarise(
    Drug_Class = paste(sort(unique(na.omit(Drug_Class[Drug_Class != ""]))), collapse = "; "),
    Resistance_Mechanism = paste(sort(unique(na.omit(Resistance_Mechanism[Resistance_Mechanism != ""]))), collapse = "; "),
    .groups = "drop"
  )

top20_aro_export <- res_aro_both %>%
  select(
    ARO_Term,
    log2FC_PIG_CTRL,
    pvalue_PIG_CTRL,
    padj_PIG_CTRL,
    log2FC_PIGFARMER_CTRL,
    pvalue_PIGFARMER_CTRL,
    padj_PIGFARMER_CTRL
  ) %>%
  mutate(
    median_log2FC = apply(
      cbind(log2FC_PIG_CTRL, log2FC_PIGFARMER_CTRL),
      1,
      function(x) median(x, na.rm = TRUE)
    )
  ) %>%
  filter(ARO_Term %in% top20_aro_terms) %>%
  left_join(aro_annotation_tbl, by = "ARO_Term") %>%
  mutate(
    Drug_Class = if_else(is.na(Drug_Class) | Drug_Class == "", "unannotated", Drug_Class),
    Resistance_Mechanism = if_else(
      is.na(Resistance_Mechanism) | Resistance_Mechanism == "",
      "unannotated",
      Resistance_Mechanism
    )
  ) %>%
  arrange(desc(median_log2FC), desc(log2FC_PIG_CTRL), desc(log2FC_PIGFARMER_CTRL)) %>%
  select(
    ARO_Term,
    Drug_Class,
    Resistance_Mechanism,
    median_log2FC,
    log2FC_PIG_CTRL,
    pvalue_PIG_CTRL,
    padj_PIG_CTRL,
    log2FC_PIGFARMER_CTRL,
    pvalue_PIGFARMER_CTRL,
    padj_PIGFARMER_CTRL
  )

write_csv(top20_aro_export, "../results/top20_aro_terms_deseq2_annotated.csv")

scatterplot_ARO <- ggplot(res_aro_both,
                          aes(x = log2FC_PIG_CTRL, y = log2FC_PIGFARMER_CTRL)) +
  geom_point(alpha = 0.5) +
  geom_point(data = res_aro_both %>% filter(ARO_Term %in% top20_aro_terms), colour = "red") +
  geom_text_repel(
    data = res_aro_both %>% filter(ARO_Term %in% top20_aro_terms),
    aes(label = ARO_Term),
    max.overlaps = 50,
    box.padding = 0.3
  ) +
  geom_vline(xintercept = c(-1, 1), colour = "black", linetype = "dashed") +
  geom_hline(yintercept = c(-1, 1), colour = "black", linetype = "dashed") +
  theme_bw(base_family = "DejaVu Sans", base_size = FIG_BASE_SIZE) +
  xlab("Log2FC | CTRL < PIG (ARO level)") +
  ylab("Log2FC | CTRL < PIG FARMER (ARO level)")

print(scatterplot_ARO)
ggsave("../figures/ARO_Term_log2FC_scatterplot.svg", scatterplot_ARO, width = 6, height = 6, web_fonts = "DejaVu Sans")
```

```{r}
# Highlight selected ARO terms on the DESeq2 ARO-level log2FC scatterplot
highlight_terms <- c("adeF", "tet(O)")

highlight_points <- res_aro_both %>%
  filter(ARO_Term %in% highlight_terms)

missing_highlight_terms <- setdiff(highlight_terms, highlight_points$ARO_Term)
if (length(missing_highlight_terms) > 0) {
  message("Not found in DESeq2 results: ", paste(missing_highlight_terms, collapse = ", "))
}

p_highlight_adef_teto <- ggplot(res_aro_both, aes(x = log2FC_PIG_CTRL, y = log2FC_PIGFARMER_CTRL)) +
  geom_point(alpha = 0.35, color = "grey55") +
  geom_vline(xintercept = c(-1, 1), colour = "black", linetype = "dashed") +
  geom_hline(yintercept = c(-1, 1), colour = "black", linetype = "dashed") +
  geom_point(data = highlight_points, color = "#D62728", size = 3.2) +
  ggrepel::geom_text_repel(
    data = highlight_points,
    aes(label = ARO_Term),
    color = "#D62728",
    size = 4,
    box.padding = 0.35,
    point.padding = 0.25,
    max.overlaps = Inf
  ) +
  theme_bw(base_family = "DejaVu Sans", base_size = FIG_BASE_SIZE) +
  labs(
    title = "DESeq2 ARO-term log2FC scatter with highlighted terms",
    subtitle = "Highlighted: adeF and tet(O)",
    x = "Log2FC | CTRL < PIG (ARO level)",
    y = "Log2FC | CTRL < PIG FARMER (ARO level)"
  )

print(p_highlight_adef_teto)
ggsave(
  "../figures/ARO_Term_log2FC_scatterplot_highlight_adeF_tetO.svg",
  p_highlight_adef_teto,
  width = 6,
  height = 6,
  web_fonts = "DejaVu Sans"
)
```

inspect dataframes

```{r}
dim(counts_aro_mat)
dim(meta)

head(colnames(counts_aro_mat), 10)
head(rownames(meta), 10)

# boolean checks
all(colnames(counts_aro_mat) %in% rownames(meta))   # are all sample names present in meta?
all(rownames(meta) %in% colnames(counts_aro_mat))   # are there extra rows in meta?
identical(colnames(counts_aro_mat), rownames(meta)) # exact order + names match?

```

#### Volcano plots and top-25 heatmaps

```{r}
# Volcano plots for both contrasts
plot_volcano <- function(res_tbl, name, alpha = 0.05, lfc_cut = 1) {
  df <- res_tbl %>% mutate(padj = ifelse(is.na(padj), 1, padj))
  df <- df %>% mutate(sig = padj < alpha & abs(log2FoldChange) >= lfc_cut)
  p <- ggplot(df, aes(x = log2FoldChange, y = -log10(padj + 1e-300))) +
    geom_point(aes(color = sig), alpha = 0.6, size = 1.2) +
    scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "red3"), guide = FALSE) +
    theme_minimal() +
    labs(title = name, x = "log2 fold change", y = "-log10(padj)")
  print(p)
  ggsave(file.path("../figures", paste0("volcano_", name, ".svg")), p, width = 6, height = 4)
}

plot_volcano(res_aro_farmer_vs_ctrl, "Pig_farmers_vs_Non_farmer_controls")
plot_volcano(res_aro_pig_vs_ctrl, "Pigs_vs_Non_farmer_controls")

# Top-25 heatmaps from VST-normalized counts
vsd <- tryCatch(
  vst(dds_aro, blind = TRUE),
  error = function(e) {
    message("vst failed: ", conditionMessage(e), " — falling back to varianceStabilizingTransformation()")
    varianceStabilizingTransformation(dds_aro, blind = TRUE)
  }
)

plot_top_heatmap <- function(vsd_obj, res_tbl, name, n_top = 25) {
  genes <- res_tbl %>%
    filter(!is.na(padj)) %>%
    arrange(padj) %>%
    slice_head(n = n_top) %>%
    pull(ARO_Term)

  if (length(genes) < 2) {
    genes <- res_tbl %>%
      arrange(desc(abs(log2FoldChange))) %>%
      slice_head(n = min(n_top, nrow(res_tbl))) %>%
      pull(ARO_Term)
  }

  mat <- assay(vsd_obj)[genes, rownames(meta), drop = FALSE]
  mat_z <- t(scale(t(mat)))
  ann <- data.frame(category = meta$category)
  rownames(ann) <- rownames(meta)

  pheatmap(
    mat_z,
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    annotation_col = ann,
    filename = file.path("../figures", paste0("heatmap_top", length(genes), "_", name, ".png")),
    width = 8,
    height = 6
  )
}

plot_top_heatmap(vsd, res_aro_farmer_vs_ctrl, "Pig_farmers_vs_Non_farmer_controls", n_top = 25)
plot_top_heatmap(vsd, res_aro_pig_vs_ctrl, "Pigs_vs_Non_farmer_controls", n_top = 25)
```



### Richness and Drug-Classes ARO

#### richness
```{r}

# ARO-Term level richness and alpha diversity (depth-corrected)

# 1) Aggregate CPM from gene level to ARO-Term level
gene_to_aro_local <- feats %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  transmute(
    Prevalence_Sequence_ID,
    ARO_Term = if_else(
      is.na(ARO_Term) | ARO_Term == "",
      Prevalence_Sequence_ID,
      as.character(ARO_Term)
    )
  )

cpm_aro_long_local <- cpm %>%
  as.data.frame() %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  pivot_longer(
    cols = -Prevalence_Sequence_ID,
    names_to = "sample",
    values_to = "cpm"
  ) %>%
  left_join(gene_to_aro_local, by = "Prevalence_Sequence_ID") %>%
  group_by(sample, ARO_Term) %>%
  summarise(cpm_aro = sum(cpm, na.rm = TRUE), .groups = "drop")

cpm_aro_mat <- cpm_aro_long_local %>%
  pivot_wider(names_from = sample, values_from = cpm_aro, values_fill = 0) %>%
  as.data.frame()

rownames(cpm_aro_mat) <- cpm_aro_mat$ARO_Term
cpm_aro_mat$ARO_Term <- NULL
cpm_aro_mat <- as.matrix(cpm_aro_mat)

# Keep only samples present in metadata and depth vector
samples_keep <- intersect(colnames(cpm_aro_mat), rownames(meta))
samples_keep <- intersect(samples_keep, names(sample_depths))
cpm_aro_mat <- cpm_aro_mat[, samples_keep, drop = FALSE]

# 2) Richness on ARO-Term level (presence/absence of ARO terms)
richness_aro_df <- tibble(
  sample = colnames(cpm_aro_mat),
  aro_richness = colSums(cpm_aro_mat > 0, na.rm = TRUE)
) %>%
  left_join(meta %>% rownames_to_column("sample"), by = "sample") %>%
  mutate(
    category = as.factor(category),
    depth = as.numeric(sample_depths[sample])
  )

# Kruskal-Wallis test for the richness plot
kw_richness_aro <- rstatix::kruskal_test(richness_aro_df, aro_richness ~ category)

# perform a pairwise Wilcoxon-test
pairwise_richness_aro <- richness_aro_df %>%
  pairwise_wilcox_test(aro_richness ~ category, p.adjust.method = "BH") %>%
  add_significance() %>%
  mutate(
    y.position = seq(
      from = max(richness_aro_df$aro_richness, na.rm = TRUE) * 1.05,
      to   = max(richness_aro_df$aro_richness, na.rm = TRUE) * 1.25,
      length.out = n()
    )
  )

# richness plot with global Kruskal-Wallis p-value annotation
p_richness_aro <- ggplot(data = richness_aro_df, aes(x = category, y = aro_richness, fill = category)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_fill_manual(values = sample_type_colors) +
  labs(
    title = "ARO-term richness by category",
    subtitle = "Global p-value from Kruskal-Wallis test",
    x = NULL,
    y = "ARO-term richness"
  ) +
  stat_pvalue_manual(
    pairwise_richness_aro,
    label = "p.adj",
    tip.length = 0.01,
    size = 5,
    family = "DejaVu Sans"
  )+
  scale_x_discrete(labels = color_codes$label_mapping)+
  theme_publication() +
  theme(legend.position = "none")


print(p_richness_aro)

ggsave("../figures/aro_term_richness_by_category_kruskal.svg", p_richness_aro, width = 8, height = 6, web_fonts = "DejaVu Sans")


# Richness on ARO-term level by pig growth stage (Pigs only)
richness_aro_stage_df <- richness_aro_df %>%
  filter(category == "Pigs", !is.na(growth_stage), growth_stage != "") %>%
  mutate(growth_stage = as.character(growth_stage))


ordered_stages <- c("Suckling", "Weaning", "Fattening")
present_stages <- ordered_stages[ordered_stages %in% unique(richness_aro_stage_df$growth_stage)]

if (length(present_stages) >= 2) {
  richness_aro_stage_df <- richness_aro_stage_df %>%
    filter(growth_stage %in% present_stages) %>%
    mutate(growth_stage = factor(growth_stage, levels = present_stages))

  pairwise_richness_aro_stage <- richness_aro_stage_df %>%
    pairwise_wilcox_test(aro_richness ~ growth_stage, p.adjust.method = "BH") %>%
    add_significance() %>%
    mutate(
      y.position = seq(
        from = max(richness_aro_stage_df$aro_richness, na.rm = TRUE) * 1.05,
        to = max(richness_aro_stage_df$aro_richness, na.rm = TRUE) * 1.20,
        length.out = n()
      ),
      label = if_else(p.adj < 0.001, "p < 0.001", paste0("p = ", signif(p.adj, 3)))
    )

  p_richness_aro_stage <- ggplot(
    richness_aro_stage_df,
    aes(x = growth_stage, y = aro_richness, fill = growth_stage)
  ) +
    geom_boxplot(alpha = 0.8) +
    geom_jitter(width = 0.15, alpha = 0.55) +
    scale_fill_manual(values = pig_stage_colors[present_stages], drop = FALSE) +
    labs(
      title = "ARO-term richness by pig growth stage",
      subtitle = "Pairwise Wilcoxon rank-sum tests (BH-adjusted)",
      x = NULL,
      y = "ARO-term richness"
    ) +
    stat_pvalue_manual(
      pairwise_richness_aro_stage,
      label = "label",
      tip.length = 0.01,
      size = 5,
      family = "DejaVu Sans"
    ) +
    theme_publication() +
    theme(legend.position = "none")

  print(p_richness_aro_stage)

  write_csv(pairwise_richness_aro_stage, "../results/aro_term_richness_pig_growth_stage_pairwise_wilcox.csv")

  ggsave(
    "../figures/aro_term_richness_by_pig_growth_stage_wilcox.svg",
    p_richness_aro_stage,
    width = 8,
    height = 6,
    web_fonts = "DejaVu Sans"
  )
} else {
  message("Not enough pig growth stages available for pairwise Wilcoxon tests on ARO-term richness.")
}




```

#### alpha diversity ARO-Term level
```{r}
# alpha diversity 

lm_richness_aro <- lm(aro_richness ~ category + depth, data = richness_aro_df)
anova_richness_aro <- anova(lm_richness_aro)
emm_richness_aro <- emmeans::emmeans(lm_richness_aro, ~ category)
posthoc_richness_aro <- pairs(emm_richness_aro, adjust = "tukey") %>% as.data.frame()

cat("=== ARO-Term richness (depth corrected) ===\n")
print(anova_richness_aro)
cat("Category p-value:", anova_richness_aro["category", "Pr(>F)"], "\n")
print(posthoc_richness_aro)
print(kw_richness_aro)

# 3) Shannon + Simpson on ARO-Term level (depth corrected)
alpha_aro_df <- tibble(
  sample = colnames(cpm_aro_mat),
  Shannon = vegan::diversity(t(cpm_aro_mat), index = "shannon"),
  Simpson = vegan::diversity(t(cpm_aro_mat), index = "simpson")
) %>%
  left_join(meta %>% rownames_to_column("sample"), by = "sample") %>%
  mutate(
    category = as.factor(category),
    depth = as.numeric(sample_depths[sample])
  )

alpha_aro_models <- c("Shannon", "Simpson") %>%
  set_names() %>%
  purrr::map(~ lm(as.formula(paste(.x, "~ category + depth")), data = alpha_aro_df))

alpha_aro_anova <- purrr::map2_dfr(alpha_aro_models, names(alpha_aro_models), ~ {
  a <- anova(.x)
  tibble(
    measure = .y,
    p_category = a["category", "Pr(>F)"],
    p_depth = a["depth", "Pr(>F)"]
  )
})

alpha_aro_posthoc <- purrr::map2_dfr(alpha_aro_models, names(alpha_aro_models), ~ {
  emmeans::emmeans(.x, ~ category) %>%
    pairs(adjust = "tukey") %>%
    as.data.frame() %>%
    mutate(measure = .y)
})

cat("=== ARO-Term alpha diversity (depth corrected) ===\n")
print(alpha_aro_anova)
print(alpha_aro_posthoc)

# alpha diversity plot (depth-corrected statistics from linear model + Tukey)
alpha_aro_posthoc_plot <- alpha_aro_posthoc %>%
  tidyr::separate(contrast, into = c("group1", "group2"), sep = " - ") %>%
  mutate(
    group1 = trimws(group1),
    group2 = trimws(group2)
  ) %>%
  group_by(measure) %>%
  mutate(
    y.position = seq(
      from = max(alpha_aro_df[[measure[1]]], na.rm = TRUE) * 1.03,
      by = diff(range(alpha_aro_df[[measure[1]]], na.rm = TRUE)) * 0.10,
      length.out = n()
    ),
    label = scales::pvalue(p.value, add_p = TRUE)
  ) %>%
  ungroup()

alpha_aro_long <- alpha_aro_df %>%
  pivot_longer(
    cols = c("Shannon", "Simpson"),
    names_to = "measure",
    values_to = "value"
  )

p_alpha_aro <- ggplot(alpha_aro_long, aes(x = category, y = value, fill = category)) +
  geom_boxplot(alpha = 0.75, outlier.shape = NA) +
  geom_jitter(width = 0.18, alpha = 0.55, size = 1.8) +
  stat_pvalue_manual(
    alpha_aro_posthoc_plot,
    label = "label",
    y.position = "y.position",
    xmin = "group1",
    xmax = "group2",
    tip.length = 0.01,
    step.increase = 0,
    bracket.size = 0.35, 
    family = "DejaVu Sans"
  ) +
  facet_wrap(~measure, scales = "free_y") +
  scale_fill_manual(values = sample_type_colors) +
  labs(
    title = "ARO-Term alpha diversity by category",
    subtitle = "Linear model (depth-corrected) with Tukey-adjusted pairwise p-values",
    x = NULL,
    y = "Diversity index"
  ) +
  theme_publication() +
  theme(legend.position = "none")+
  scale_x_discrete(labels = color_codes$label_mapping)

print(p_alpha_aro)

write_csv(richness_aro_df, "../results/aro_term_richness_depth_corrected_values.csv")
write_csv(posthoc_richness_aro, "../results/aro_term_richness_depth_corrected_posthoc_tukey.csv")
write_csv(kw_richness_aro, "../results/aro_term_richness_kruskal_test.csv")
write_csv(alpha_aro_df, "../results/aro_term_alpha_diversity_depth_corrected_values.csv")
write_csv(alpha_aro_anova, "../results/aro_term_alpha_diversity_depth_corrected_anova.csv")
write_csv(alpha_aro_posthoc, "../results/aro_term_alpha_diversity_depth_corrected_posthoc_tukey.csv")

ggsave("../figures/aro_term_alpha_diversity_depth_corrected.svg", p_alpha_aro, width = 10.5, height = 6, web_fonts = "DejaVu Sans")

```


#### Pileou's eveness
```{r}

# Pielou's evenness on ARO-Term level: J = Shannon / log(S)
evenness_aro_df <- alpha_aro_df %>%
  left_join(richness_aro_df %>% select(sample, aro_richness), by = "sample") %>%
  mutate(
    pielou_J = if_else(aro_richness > 1, Shannon / log(aro_richness), NA_real_),
    category = as.factor(category)
  ) %>%
  filter(!is.na(pielou_J), is.finite(pielou_J))

# Depth-corrected linear model and Tukey posthoc
lm_evenness_aro <- lm(pielou_J ~ category + depth, data = evenness_aro_df)
anova_evenness_aro <- anova(lm_evenness_aro)
emm_evenness_aro <- emmeans::emmeans(lm_evenness_aro, ~ category)
posthoc_evenness_aro <- pairs(emm_evenness_aro, adjust = "tukey") %>%
  as.data.frame() %>%
  tidyr::separate(contrast, into = c("group1", "group2"), sep = " - ") %>%
  mutate(
    group1 = trimws(group1),
    group2 = trimws(group2),
    label = scales::pvalue(p.value, add_p = TRUE)
  )

posthoc_evenness_aro <- posthoc_evenness_aro %>%
  mutate(
    y.position = seq(
      from = max(evenness_aro_df$pielou_J, na.rm = TRUE) * 1.03,
      by = diff(range(evenness_aro_df$pielou_J, na.rm = TRUE)) * 0.10,
      length.out = n()
    )
  )

p_evenness_aro <- ggplot(evenness_aro_df, aes(x = category, y = pielou_J, fill = category)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.18, alpha = 0.55, size = 1.8) +
  stat_pvalue_manual(
    posthoc_evenness_aro,
    label = "label",
    y.position = "y.position",
    xmin = "group1",
    xmax = "group2",
    tip.length = 0.01,
    bracket.size = 0.35,
    family = "DejaVu Sans"
  ) +
  scale_fill_manual(values = sample_type_colors) +
  labs(
    title = "Pielou's Evenness (J) on ARO-Term level",
    subtitle = "Linear model (depth-corrected) with Tukey-adjusted pairwise p-values",
    x = NULL,
    y = "Evenness index (J)"
  ) +
  theme_publication() +
  theme(legend.position = "none")+
  scale_x_discrete(labels = color_codes$label_mapping)

print(p_evenness_aro)

cat("=== ARO-Term Pielou evenness (depth corrected) ===\n")
print(anova_evenness_aro)
print(posthoc_evenness_aro)

write_csv(evenness_aro_df, "../results/aro_term_pielou_evenness_depth_corrected_values.csv")
write_csv(
  tibble(
    measure = "Pielou_J",
    p_category = anova_evenness_aro["category", "Pr(>F)"],
    p_depth = anova_evenness_aro["depth", "Pr(>F)"]
  ),
  "../results/aro_term_pielou_evenness_depth_corrected_anova.csv"
)
write_csv(posthoc_evenness_aro, "../results/aro_term_pielou_evenness_depth_corrected_posthoc_tukey.csv")

ggsave(
  "../figures/aro_term_pielou_evenness_depth_corrected.svg",
  p_evenness_aro,
  width = 6,
  height = 6,
  web_fonts = "DejaVu Sans"
)

```


## CARD - Drug-Classes 
Aggregate `CPM` first to `ARO_Term`, then to `Drug_Class` per sample, and plot abundance by sample category.

```{r}
# Build a gene -> ARO map
gene_to_aro <- feats %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  transmute(
    Prevalence_Sequence_ID,
    ARO_Term = if_else(is.na(ARO_Term) | ARO_Term == "", Prevalence_Sequence_ID, ARO_Term)
  )

# CPM matrix to long format (gene x sample)
cpm_long <- cpm %>%
  as.data.frame() %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  pivot_longer(
    cols = -Prevalence_Sequence_ID,
    names_to = "sample",
    values_to = "cpm"
  )

# 1) Aggregate to ARO_Term per sample
cpm_aro_long <- cpm_long %>%
  left_join(gene_to_aro, by = "Prevalence_Sequence_ID") %>%
  group_by(sample, ARO_Term) %>%
  summarise(cpm_aro = sum(cpm, na.rm = TRUE), .groups = "drop")

# 2) Map ARO_Term -> Drug_Class (split multi-classes and weight equally)
aro_to_drug <- feats %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  transmute(
    ARO_Term = if_else(is.na(ARO_Term) | ARO_Term == "", Prevalence_Sequence_ID, ARO_Term),
    Drug_Class
  ) %>%
  distinct() %>%
  mutate(Drug_Class = str_split(Drug_Class, ";\\s*")) %>%
  unnest(Drug_Class) %>%
  mutate(Drug_Class = str_trim(Drug_Class)) %>%
  filter(!is.na(Drug_Class), Drug_Class != "") %>%
  distinct() %>%
  group_by(ARO_Term) %>%
  mutate(n_drug_classes = n(), weight = 1 / n_drug_classes) %>%
  ungroup()

# 3) Aggregate to Drug_Class per sample
cpm_drugclass_long <- cpm_aro_long %>%
  left_join(aro_to_drug, by = "ARO_Term") %>%
  filter(!is.na(Drug_Class)) %>%
  mutate(cpm_weighted = cpm_aro * weight) %>%
  group_by(sample, Drug_Class) %>%
  summarise(cpm_drugclass = sum(cpm_weighted, na.rm = TRUE), .groups = "drop") %>%
  left_join(
    meta %>%
      rownames_to_column("sample") %>%
      select(sample, category),
    by = "sample"
  ) %>%
  filter(!is.na(category))

# Save aggregated table for downstream analyses
write_csv(cpm_drugclass_long, "../results/cpm_drugclass_per_sample.csv")

cat("Rows in cpm_drugclass_long:", nrow(cpm_drugclass_long), "\n")
cat("Unique drug classes:", n_distinct(cpm_drugclass_long$Drug_Class), "\n")
```


### extract names of CARD drug classes

```{r}
aro_meta_long <- feats %>%
  select(ARO_Term, Drug_Class, Resistance_Mechanism) %>%
  distinct() %>%
  # split semicolon-separated fields into multiple rows
  mutate(
    Drug_Class = str_split(Drug_Class, ";\\s*"),
    Resistance_Mechanism = str_split(Resistance_Mechanism, ";\\s*")
  ) %>%
  unnest(Drug_Class) %>%
  unnest(Resistance_Mechanism) %>%
  mutate(
    Drug_Class = str_trim(Drug_Class),
    Resistance_Mechanism = str_trim(Resistance_Mechanism)
  ) %>%
  filter(!is.na(Drug_Class) & !is.na(Resistance_Mechanism) & 
         Drug_Class != "" & Resistance_Mechanism != "") %>%
  distinct()

unique(aro_meta_long$Drug_Class)
```
### Drug Class mean abundance
this should give a table, where we have the drug class and then the mean abundance per drug class from the cpm values over all samples.
Since I want to make a new annotation of the drug classes it's important to then be able to set a threshold in order to validate the new drug class names
```{r}
drug_class_mean_cpm <- cpm_drugclass_long %>%
  group_by(Drug_Class)%>%
  summarise(
    mean_drug_class_cpm = mean(cpm_drugclass)
  )%>%
  ungroup()%>%
  arrange(desc(mean_drug_class_cpm))

head(drug_class_mean_cpm)


ggplot(data = drug_class_mean_cpm, aes (x = mean_drug_class_cpm))+
  geom_histogram()

write_csv(drug_class_mean_cpm, "../results/drug_class_mean_cpm.csv")





```

### relative abundance barplot per category

```{r}
# Relative Abundanz pro Sample und Drug Class
# Order classes by overall abundance for consistent stack + legend order.
drug_class_levels <- cpm_drugclass_long %>%
  group_by(Drug_Class) %>%
  summarise(total_cpm = sum(cpm_drugclass, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_cpm)) %>%
  pull(Drug_Class)

drug_class_palette <- grDevices::hcl.colors(length(drug_class_levels), palette = "Dynamic")
names(drug_class_palette) <- drug_class_levels

# Keep "others" neutral if present.
other_idx <- which(tolower(names(drug_class_palette)) == "others")
if (length(other_idx) > 0) {
  drug_class_palette[other_idx] <- "#BDBDBD"
}

plot_rel_abundance_drugclass <- cpm_drugclass_long %>%
  mutate(Drug_Class = factor(Drug_Class, levels = drug_class_levels)) %>%
  ggplot(aes(x = category, y = cpm_drugclass, fill = Drug_Class)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = drug_class_palette, drop = FALSE) +
  labs(x = NULL, y = "Relative abundance", fill = "Drug class") +
  my_theme()

print(plot_rel_abundance_drugclass)

# ggsave
ggsave("../figures/drug_class_rel_abundances.svg", plot = plot_rel_abundance_drugclass, width = 12, height = 12, web_fonts = "DejaVu Sans")
```

## Final drug class annotation

### load manual annotation

```{r}
library(readxl)
drug_class_annot_manual <- read_excel("../drug_class_fam_mean_cpm_2.xlsx")

glimpse(drug_class_annot_manual)

length(unique(drug_class_annot_manual$Drug_Class_Resfinder))

# remove "biocides" from the annotation since it's not a drug class and we don't have a clear definition of which classes it should map to
drug_class_annot_manual <- drug_class_annot_manual %>%
  filter(Drug_Class_Resfinder != "Biocides")

cat("länge nach dem Biocides gefiltert wurden:", length(unique(drug_class_annot_manual$Drug_Class_Resfinder)), "drug classes")

```

### aggregate according to new annotation

```{r}
# Map old Drug_Class to the new manual annotation and aggregate CPM by new class.
drug_class_resfinder_map <- drug_class_annot_manual %>%
  transmute(
    Drug_Class = str_trim(as.character(Drug_Class)),
    Drug_Class_Resfinder = str_trim(as.character(Drug_Class_Resfinder))
  ) %>%
  filter(!is.na(Drug_Class), Drug_Class != "") %>%
  distinct(Drug_Class, .keep_all = TRUE)

cpm_resfinder_long <- cpm_drugclass_long %>%
  mutate(Drug_Class = str_trim(as.character(Drug_Class))) %>%
  left_join(drug_class_resfinder_map, by = "Drug_Class") %>%
  filter(!is.na(Drug_Class_Resfinder), Drug_Class_Resfinder != "") %>%
  group_by(sample, Drug_Class_Resfinder) %>%
  summarise(cpm_resfinder = sum(cpm_drugclass, na.rm = TRUE), .groups = "drop")

# join with metadata
cpm_resfinder_long <- cpm_resfinder_long %>%
  left_join(
    meta %>%
      rownames_to_column("sample"),
    by = "sample"
  )


write_csv(cpm_resfinder_long, "../results/cpm_drugclass_resfinder_per_sample.csv")

cat("Rows in cpm_resfinder_long:", nrow(cpm_resfinder_long), "\n")
cat("Unique Drug_Class_Resfinder labels:", n_distinct(cpm_resfinder_long$Drug_Class_Resfinder), "\n")
```

### ARO-term boxplots for all new drug classes + long annotation export

```{r}
# Build ARO-term level CPM per sample from full data (not DESeq subset)
gene_to_aro <- feats %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  transmute(
    Gene = as.character(Prevalence_Sequence_ID),
    ARO_Term = if_else(is.na(ARO_Term) | ARO_Term == "", Gene, as.character(ARO_Term)),
    Drug_Class = as.character(Drug_Class)
  )

cpm_aro_long_all <- as.data.frame(cpm) %>%
  rownames_to_column("Gene") %>%
  pivot_longer(-Gene, names_to = "sample", values_to = "cpm") %>%
  left_join(gene_to_aro %>% select(Gene, ARO_Term), by = "Gene") %>%
  filter(!is.na(ARO_Term), ARO_Term != "") %>%
  group_by(sample, ARO_Term) %>%
  summarise(cpm_aro = sum(cpm, na.rm = TRUE), .groups = "drop") %>%
  left_join(meta %>% rownames_to_column("sample") %>% select(sample, category), by = "sample")

# Map ARO terms to new manual drug classes
aro_to_resfinder_long <- gene_to_aro %>%
  mutate(Drug_Class = str_trim(Drug_Class)) %>%
  separate_rows(Drug_Class, sep = ";\\s*") %>%
  mutate(Drug_Class = str_trim(Drug_Class)) %>%
  filter(!is.na(Drug_Class), Drug_Class != "") %>%
  left_join(drug_class_resfinder_map, by = "Drug_Class") %>%
  filter(!is.na(Drug_Class_Resfinder), Drug_Class_Resfinder != "") %>%
  distinct(ARO_Term, Drug_Class_Resfinder)

# Count how many new classes each ARO term maps to and derive weight 1/n
aro_class_counts <- aro_to_resfinder_long %>%
  dplyr::count(ARO_Term, name = "n_resfinder_classes") %>%
  mutate(
    weight_1_over_n = 1 / n_resfinder_classes,
    multi_annotation_flag = n_resfinder_classes > 1
  )

# Long annotation table with overall and per-category medians.
aro_mean_cpm_tbl <- cpm_aro_long_all %>%
  group_by(ARO_Term) %>%
  summarise(mean_cpm_aro = mean(cpm_aro, na.rm = TRUE),
            median_cpm_aro = median(cpm_aro, na.rm = TRUE),
                                    .groups = "drop")

aro_median_cpm_by_category_long <- cpm_aro_long_all %>%
  filter(!is.na(category), category != "") %>%
  group_by(ARO_Term, category) %>%
  summarise(
    median_cpm_aro_category = median(cpm_aro, na.rm = TRUE),
    .groups = "drop"
  )

aro_median_cpm_by_category_wide <- aro_median_cpm_by_category_long %>%
  pivot_wider(
    names_from = category,
    values_from = median_cpm_aro_category,
    names_prefix = "median_cpm_"
  )

aro_resfinder_annotation_long <- aro_to_resfinder_long %>%
  left_join(aro_mean_cpm_tbl, by = "ARO_Term") %>%
  left_join(aro_median_cpm_by_category_wide, by = "ARO_Term") %>%
  left_join(aro_class_counts, by = "ARO_Term") %>%
  arrange(Drug_Class_Resfinder, desc(mean_cpm_aro), ARO_Term)

# Weighted mean CPM contribution for a class if an ARO term has multiple classes
aro_resfinder_annotation_long <- aro_resfinder_annotation_long %>%
  mutate(weighted_mean_cpm_contribution = mean_cpm_aro * weight_1_over_n)

write_csv(
  aro_resfinder_annotation_long,
  "../results/aro_term_resfinder_annotation_long.csv"
)

write_csv(
  aro_median_cpm_by_category_long,
  "../results/aro_term_median_cpm_by_category_long.csv"
)

if (requireNamespace("writexl", quietly = TRUE)) {
  writexl::write_xlsx(
    list(
      aro_term_resfinder_annotation_long = aro_resfinder_annotation_long,
      aro_term_median_cpm_by_category = aro_median_cpm_by_category_long,
      aro_term_class_counts = aro_class_counts
    ),
    "../results/aro_term_resfinder_annotation_long.xlsx"
  )
} else {
  message("Package 'writexl' not installed; CSV export was written instead of XLSX.")
}

# Create one PDF per Drug_Class_Resfinder with ARO-term boxplots (multi-page if needed)
out_dir <- "../figures/aro_term_boxplots_by_resfinder"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

terms_per_page <- 12
class_levels <- sort(unique(aro_to_resfinder_long$Drug_Class_Resfinder))
pdf_index <- vector("list", length(class_levels))

for (i in seq_along(class_levels)) {
  class_i <- class_levels[i]

  class_terms <- aro_to_resfinder_long %>%
    filter(Drug_Class_Resfinder == class_i) %>%
    pull(ARO_Term) %>%
    unique()

  class_plot_df <- cpm_aro_long_all %>%
    filter(ARO_Term %in% class_terms, !is.na(category))

  # Order terms by median abundance inside class
  term_order <- class_plot_df %>%
    group_by(ARO_Term) %>%
    summarise(median_cpm = median(cpm_aro, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(median_cpm)) %>%
    pull(ARO_Term)

  class_plot_df <- class_plot_df %>%
    mutate(ARO_Term = factor(ARO_Term, levels = term_order))

  n_pages <- ceiling(length(term_order) / terms_per_page)
  safe_name <- gsub("[^A-Za-z0-9]+", "_", tolower(class_i))
  pdf_file <- file.path(out_dir, paste0("aro_terms_", safe_name, ".pdf"))

  grDevices::cairo_pdf(pdf_file, width = 12, height = 8, onefile = TRUE)
  for (page_i in seq_len(n_pages)) {
    idx_start <- (page_i - 1) * terms_per_page + 1
    idx_end <- min(page_i * terms_per_page, length(term_order))
    terms_page <- term_order[idx_start:idx_end]

    p_page <- class_plot_df %>%
      filter(ARO_Term %in% terms_page) %>%
      ggplot(aes(x = category, y = log10(cpm_aro + 1e-6), fill = category)) +
      geom_boxplot(outlier.alpha = 0.25, alpha = 0.8, outlier.shape = NA) +
      geom_point(alpha = 0.8, size = 0.8) +
      facet_wrap(~ ARO_Term, scales = "free_y", ncol = 4) +
      scale_fill_manual(values = sample_type_colors) +
      my_theme() +
      labs(
        title = paste0(class_i, " | ARO-term boxplots (page ", page_i, " / ", n_pages, ")"),
        x = "Sample category",
        y = "log10(CPM + 1e-6)"
      ) +
      theme(
        legend.position = "none",
        strip.text = element_text(size = 8)
      )

    print(p_page)
  }
  grDevices::dev.off()

  pdf_index[[i]] <- tibble(
    Drug_Class_Resfinder = class_i,
    n_aro_terms = length(term_order),
    n_pages = n_pages,
    pdf_file = pdf_file
  )
}

pdf_index_tbl <- bind_rows(pdf_index)
write_csv(pdf_index_tbl, "../results/aro_term_boxplot_pdf_index.csv")

cat("Generated", nrow(pdf_index_tbl), "class-level PDF files in", out_dir, "\n")
cat("Annotation long table rows:", nrow(aro_resfinder_annotation_long), "\n")
```

### ARO-term category dotplots for each new drug class

```{r}
# Same idea as peptide-antibiotic dotplot, generated automatically for each new class.
dot_out_dir <- "../figures/aro_term_category_dotplots_by_resfinder"
if (!dir.exists(dot_out_dir)) dir.create(dot_out_dir, recursive = TRUE)

aro_class_cat_summary <- cpm_aro_long_all %>%
  inner_join(aro_to_resfinder_long, by = "ARO_Term") %>%
  filter(!is.na(category), !is.na(Drug_Class_Resfinder)) %>%
  group_by(Drug_Class_Resfinder, ARO_Term, category) %>%
  summarise(
    median_log10_cpm = median(log10(cpm_aro + 1e-6), na.rm = TRUE),
    prevalence_pct = mean(cpm_aro > 0, na.rm = TRUE) * 100,
    mean_cpm = mean(cpm_aro, na.rm = TRUE),
    median_cpm = median(cpm_aro, na.rm = TRUE),
    n_samples = n(),
    .groups = "drop"
  )

plot_index_list <- vector("list", length(class_levels))
plot_obj_list <- vector("list", length(class_levels))
export_font_family <- "sans"

for (i in seq_along(class_levels)) {
  class_i <- class_levels[i]

  class_tbl <- aro_class_cat_summary %>%
    filter(Drug_Class_Resfinder == class_i)

  if (nrow(class_tbl) == 0) {
    next
  }

  aro_dot_order <- class_tbl %>%
    group_by(ARO_Term) %>%
    summarise(overall_median_log10 = median(median_log10_cpm, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(overall_median_log10)) %>%
    pull(ARO_Term)

  class_tbl <- class_tbl %>%
    mutate(ARO_Term = factor(ARO_Term, levels = aro_dot_order))

  n_terms <- nlevels(class_tbl$ARO_Term)
  axis_y_size <- dplyr::case_when(
    n_terms <= 60 ~ 7,
    n_terms <= 100 ~ 6,
    n_terms <= 150 ~ 5,
    TRUE ~ 4
  )

  p_class_dot <- ggplot(
    class_tbl,
    aes(x = category, y = ARO_Term, size = prevalence_pct, color = median_log10_cpm)
  ) +
    geom_point(alpha = 0.9) +
    scale_size_continuous(range = c(1.2, 8), limits = c(0, 100), name = "Prevalence (%)") +
    scale_color_viridis_c(option = "C", name = "Median log10(CPM + 1e-6)") +
    my_theme() +
    labs(
      title = paste0("Category summary of ARO terms: ", class_i),
      subtitle = "Color: median abundance, Size: prevalence across samples",
      x = "Sample Category",
      y = "ARO Term"
    ) +
    theme(
      text = element_text(family = export_font_family),
      axis.text.y = element_text(size = axis_y_size),
      legend.position = "right"
    )

  safe_name <- gsub("[^A-Za-z0-9]+", "_", tolower(class_i))
  out_svg <- file.path(dot_out_dir, paste0("aro_category_summary_", safe_name, ".svg"))
  out_height_raw <- max(5, 0.26 * n_terms + 3)
  out_height <- min(45, out_height_raw)
  height_was_capped <- out_height_raw > out_height

  ggsave(out_svg, p_class_dot, device = "svg", width = 12, height = out_height, dpi = 300)
  plot_obj_list[[i]] <- p_class_dot

  plot_index_list[[i]] <- tibble(
    Drug_Class_Resfinder = class_i,
    n_aro_terms = n_distinct(class_tbl$ARO_Term),
    plot_height_in = out_height,
    height_was_capped = height_was_capped,
    svg_file = out_svg
  )
}

plot_index_tbl <- bind_rows(plot_index_list)
write_csv(plot_index_tbl, "../results/aro_term_category_dotplots_resfinder_index.csv")

# Single multi-page PDF with all class plots (one page per class)
combined_pdf_file <- "../figures/aro_term_category_dotplots_all_resfinder_classes.pdf"
grDevices::cairo_pdf(combined_pdf_file, width = 12, height = 16, onefile = TRUE)
for (p in plot_obj_list) {
  if (!is.null(p)) print(p)
}
grDevices::dev.off()

cat("Generated", nrow(plot_index_tbl), "per-class ARO-term dotplots in", dot_out_dir, "\n")
cat("Combined multi-page PDF:", combined_pdf_file, "\n")
```

### Weighted category summary on new drug classes

```{r}
# Join sample-level ARO CPM to class mapping and apply 1/n weighting for multi-annotated ARO terms.
drugclass_weighted_long <- cpm_aro_long_all %>%
  left_join(
    aro_resfinder_annotation_long %>%
      select(ARO_Term, Drug_Class_Resfinder, n_resfinder_classes, weight_1_over_n, multi_annotation_flag) %>%
      distinct(),
    by = "ARO_Term"
  ) %>%
  filter(!is.na(Drug_Class_Resfinder), Drug_Class_Resfinder != "") %>%
  mutate(cpm_weighted = cpm_aro * weight_1_over_n)

# Optional: exclude glycopeptides from this overview (case-insensitive match).
drugclass_weighted_long_no_gly <- drugclass_weighted_long %>%
  filter(!str_detect(str_to_lower(Drug_Class_Resfinder), "glycopeptide"))

# Per sample x class weighted CPM.
drugclass_weighted_sample <- drugclass_weighted_long_no_gly %>%
  group_by(sample, category, Drug_Class_Resfinder) %>%
  summarise(cpm_class_weighted = sum(cpm_weighted, na.rm = TRUE), .groups = "drop")

# Category summary for dotplot.
drugclass_category_summary <- drugclass_weighted_sample %>%
  group_by(Drug_Class_Resfinder, category) %>%
  summarise(
    median_log10_cpm = median(log10(cpm_class_weighted + 1e-6), na.rm = TRUE),
    prevalence_pct = mean(cpm_class_weighted > 0, na.rm = TRUE) * 100,
    mean_cpm = mean(cpm_class_weighted, na.rm = TRUE),
    median_cpm = median(cpm_class_weighted, na.rm = TRUE),
    n_samples = n(),
    .groups = "drop"
  )

class_dot_order <- drugclass_category_summary %>%
  group_by(Drug_Class_Resfinder) %>%
  summarise(overall_median_log10 = median(median_log10_cpm, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(overall_median_log10)) %>%
  pull(Drug_Class_Resfinder)

drugclass_category_summary <- drugclass_category_summary %>%
  mutate(Drug_Class_Resfinder = factor(Drug_Class_Resfinder, levels = class_dot_order))

p_drugclass_category_dot <- ggplot(
  drugclass_category_summary,
  aes(x = category, y = Drug_Class_Resfinder, size = prevalence_pct, color = median_log10_cpm)
) +
  geom_point(alpha = 0.9) +
  scale_size_continuous(range = c(1.2, 8), limits = c(0, 100), name = "Prevalence (%)") +
  scale_color_viridis_c(option = "C", name = "Median log10(CPM + 1e-6)") +
  my_theme() +
  labs(
    title = "Weighted category summary of new Drug_Class_Resfinder",
    subtitle = "Color: median abundance, Size: prevalence, CPM weighted by 1/n for multi-annotated ARO terms",
    x = "Sample Category",
    y = "Drug Class (new annotation)"
  ) +
  theme(
    axis.text.y = element_text(size = 8),
    legend.position = "right"
  )

print(p_drugclass_category_dot)



ggsave(
  "../figures/drugclass_resfinder_category_summary_dotplot_weighted.svg",
  p_drugclass_category_dot,
  width = 11,
  height = 14,
  dpi = 300,
  web_fonts ="DejaVu Sans"
)

write_csv(
  drugclass_category_summary,
  "../results/drugclass_resfinder_category_summary_weighted.csv"
)

# Accounting check: compare unweighted and weighted totals per sample.
accounting_check <- drugclass_weighted_long_no_gly %>%
  group_by(sample) %>%
  summarise(
    total_unweighted = sum(cpm_aro, na.rm = TRUE),
    total_weighted = sum(cpm_weighted, na.rm = TRUE),
    ratio_weighted_to_unweighted = if_else(total_unweighted > 0, total_weighted / total_unweighted, NA_real_),
    .groups = "drop"
  ) %>%
  arrange(desc(abs(1 - ratio_weighted_to_unweighted)))

write_csv(accounting_check, "../results/drugclass_resfinder_weighting_accounting_check.csv")

cat("Classes in weighted dotplot summary:", n_distinct(drugclass_category_summary$Drug_Class_Resfinder), "\n")
cat("Mean weighted/unweighted ratio across samples:", round(mean(accounting_check$ratio_weighted_to_unweighted, na.rm = TRUE), 4), "\n")
```

### Quick prevalence check (new drug classes)

```{r}
# Quick prevalence of each new class across sample categories.
# Prevalence is defined as % of samples in a category with CPM > 0.
quick_prev_tbl <- cpm_resfinder_long %>%
  filter(!is.na(category)) %>%
  group_by(Drug_Class_Resfinder, category) %>%
  summarise(
    prevalence_pct = mean(cpm_resfinder > 0, na.rm = TRUE) * 100,
    n_positive = sum(cpm_resfinder > 0, na.rm = TRUE),
    n_samples = n(),
    .groups = "drop"
  )

quick_prev_order <- quick_prev_tbl %>%
  group_by(Drug_Class_Resfinder) %>%
  summarise(mean_prevalence = mean(prevalence_pct, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_prevalence)) %>%
  pull(Drug_Class_Resfinder)

quick_prev_tbl <- quick_prev_tbl %>%
  mutate(Drug_Class_Resfinder = factor(Drug_Class_Resfinder, levels = quick_prev_order))

p_quick_prev <- ggplot(
  quick_prev_tbl,
  aes(x = category, y = Drug_Class_Resfinder, fill = prevalence_pct)
) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(option = "C", limits = c(0, 100), name = "Prevalence (%)") +
  my_theme() +
  labs(
    title = "Quick prevalence check of new drug classes",
    subtitle = "Percent of samples with CPM > 0",
    x = "Sample Category",
    y = "Drug Class (new annotation)"
  ) +
  theme(axis.text.y = element_text(size = 7))

print(p_quick_prev)

write_csv(quick_prev_tbl, "../results/quick_prevalence_new_drug_classes_by_category.csv")


ggsave(
  "../figures/quick_prevalence_new_drug_classes_by_category.svg",
  p_quick_prev,
  width = 10,
  height = 12,
  dpi = 300
)

cat("Quick prevalence rows:", nrow(quick_prev_tbl), "\n")
```

### Quick prevalence check from raw counts only

```{r}
# Raw-count prevalence: class is present in a sample if summed raw counts > 0.
# This does not use CPM values.
counts_long_all <- counts_df %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  pivot_longer(
    cols = -Prevalence_Sequence_ID,
    names_to = "sample",
    values_to = "raw_count"
  )

gene_to_old_drugclass <- feats %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  transmute(
    Prevalence_Sequence_ID = as.character(Prevalence_Sequence_ID),
    Drug_Class = as.character(Drug_Class)
  ) %>%
  mutate(Drug_Class = str_trim(Drug_Class)) %>%
  separate_rows(Drug_Class, sep = ";\\s*") %>%
  mutate(Drug_Class = str_trim(Drug_Class)) %>%
  filter(!is.na(Drug_Class), Drug_Class != "")

counts_resfinder_long <- counts_long_all %>%
  left_join(gene_to_old_drugclass, by = "Prevalence_Sequence_ID") %>%
  left_join(drug_class_resfinder_map, by = "Drug_Class") %>%
  filter(!is.na(Drug_Class_Resfinder), Drug_Class_Resfinder != "") %>%
  group_by(sample, Drug_Class_Resfinder) %>%
  summarise(raw_count_resfinder = sum(raw_count, na.rm = TRUE), .groups = "drop")

# Zero-fill missing sample x class combinations for correct prevalence denominators.
all_rf_classes <- sort(unique(counts_resfinder_long$Drug_Class_Resfinder))
all_samples_meta <- rownames(meta)

counts_resfinder_long <- expand_grid(
  sample = all_samples_meta,
  Drug_Class_Resfinder = all_rf_classes
) %>%
  left_join(counts_resfinder_long, by = c("sample", "Drug_Class_Resfinder")) %>%
  mutate(raw_count_resfinder = replace_na(raw_count_resfinder, 0)) %>%
  left_join(meta %>% rownames_to_column("sample") %>% select(sample, category), by = "sample")

quick_prev_raw_tbl <- counts_resfinder_long %>%
  filter(!is.na(category)) %>%
  group_by(Drug_Class_Resfinder, category) %>%
  summarise(
    prevalence_pct = mean(raw_count_resfinder > 0, na.rm = TRUE) * 100,
    n_positive = sum(raw_count_resfinder > 0, na.rm = TRUE),
    n_samples = n(),
    .groups = "drop"
  )

quick_prev_raw_order <- quick_prev_raw_tbl %>%
  group_by(Drug_Class_Resfinder) %>%
  summarise(mean_prevalence = mean(prevalence_pct, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_prevalence)) %>%
  pull(Drug_Class_Resfinder)

quick_prev_raw_tbl <- quick_prev_raw_tbl %>%
  mutate(Drug_Class_Resfinder = factor(Drug_Class_Resfinder, levels = quick_prev_raw_order))

p_quick_prev_raw <- ggplot(
  quick_prev_raw_tbl,
  aes(x = category, y = Drug_Class_Resfinder, fill = prevalence_pct)
) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(option = "C", limits = c(0, 100), name = "Prevalence (%)") +
  my_theme() +
  labs(
    title = "Quick prevalence check of new drug classes (raw counts)",
    subtitle = "Percent of samples with summed raw counts > 0",
    x = "Sample Category",
    y = "Drug Class (new annotation)"
  ) +
  theme(axis.text.y = element_text(size = 7))

print(p_quick_prev_raw)

write_csv(quick_prev_raw_tbl, "../results/quick_prevalence_new_drug_classes_by_category_rawcounts.csv")

ggsave(
  "../figures/quick_prevalence_new_drug_classes_by_category_rawcounts.svg",
  p_quick_prev_raw,
  width = 10,
  height = 12,
  dpi = 300
)

cat("Quick raw-count prevalence rows:", nrow(quick_prev_raw_tbl), "\n")
```

### heatmaps drug class x sample with new annotation

```{r}

library(viridis)
# 1) Build full Drug_Class_Resfinder x sample matrix
all_drug_classes <- cpm_resfinder_long %>%
  distinct(Drug_Class_Resfinder) %>%
  pull(Drug_Class_Resfinder)

all_samples <- rownames(meta)

heatmap_tbl <- tidyr::expand_grid(
  Drug_Class_Resfinder = all_drug_classes,
  sample = all_samples
) %>%
  left_join(
    cpm_resfinder_long %>%
      select(sample, Drug_Class_Resfinder, cpm_resfinder),
    by = c("Drug_Class_Resfinder", "sample")
  ) %>%
  mutate(cpm_resfinder = replace_na(cpm_resfinder, 0))

heatmap_mat <- heatmap_tbl %>%
  pivot_wider(names_from = sample, values_from = cpm_resfinder) %>%
  as.data.frame()

rownames(heatmap_mat) <- heatmap_mat$Drug_Class_Resfinder
heatmap_mat$Drug_Class_Resfinder <- NULL
heatmap_mat <- as.matrix(heatmap_mat)

heatmap_log10 <- log10(heatmap_mat + 1e-6)


control_level <- "Non_farmer_controls"

# Original heatmap: all pigs pooled together
drugclass_mw_annotations_all_pigs <- cpm_resfinder_long %>%
  filter(category %in% c("Pigs", farmer_level, control_level)) %>%
  group_by(Drug_Class_Resfinder) %>%
  summarise(
    p_pigs_vs_ctrl = tryCatch(
      wilcox.test(cpm_resfinder[category == "Pigs"],
                  cpm_resfinder[category == control_level])$p.value,
      error = function(e) NA_real_
    ),
    median_pigs = median(cpm_resfinder[category == "Pigs"], na.rm = TRUE),
    median_ctrl_for_pigs = median(cpm_resfinder[category == control_level], na.rm = TRUE),
    p_farmer_vs_ctrl = tryCatch(
      wilcox.test(cpm_resfinder[category == farmer_level],
                  cpm_resfinder[category == control_level])$p.value,
      error = function(e) NA_real_
    ),
    median_farmer = median(cpm_resfinder[category == farmer_level], na.rm = TRUE),
    median_ctrl_for_farmer = median(cpm_resfinder[category == control_level], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    p_adj_pigs_vs_ctrl = p.adjust(p_pigs_vs_ctrl, method = "BH"),
    p_adj_farmer_vs_ctrl = p.adjust(p_farmer_vs_ctrl, method = "BH"),
    pigs_vs_ctrl_mw = case_when(
      is.na(p_adj_pigs_vs_ctrl) ~ "not_tested",
      p_adj_pigs_vs_ctrl < 0.05 & median_pigs > median_ctrl_for_pigs ~ "sig_higher_in_all_pigs",
      p_adj_pigs_vs_ctrl < 0.05 & median_pigs < median_ctrl_for_pigs ~ "sig_higher_in_controls",
      TRUE ~ "ns"
    ),
    farmer_vs_ctrl_mw = case_when(
      is.na(p_adj_farmer_vs_ctrl) ~ "not_tested",
      p_adj_farmer_vs_ctrl < 0.05 & median_farmer > median_ctrl_for_farmer ~ "sig_higher_in_pig_farmers",
      p_adj_farmer_vs_ctrl < 0.05 & median_farmer < median_ctrl_for_farmer ~ "sig_higher_in_controls",
      TRUE ~ "ns"
    )
  )
# Build annotation columns/colors locally so this section does not depend
# on objects from earlier heatmap code chunks.
sample_annot_var_local <- if ("sample_type" %in% colnames(meta)) "sample_type" else "category"
stage_var_local <- if ("category2" %in% colnames(meta)) {
  "category2"
} else if ("growth_stage" %in% colnames(meta)) {
  "growth_stage"
} else if ("grwoth_stage" %in% colnames(meta)) {
  "grwoth_stage"
} else {
  NA_character_
}

build_ann_col <- function(sample_ids) {
  ann <- data.frame(
    sample_type = as.character(meta[sample_ids, sample_annot_var_local]),
    category = as.character(meta[sample_ids, "category"]),
    row.names = sample_ids
  )

  if (!is.na(stage_var_local)) {
    pig_idx <- ann$category == "Pigs"
    stage_vals <- as.character(meta[sample_ids, stage_var_local])
    use_stage_idx <- pig_idx & !is.na(stage_vals) & stage_vals != ""
    ann$category[use_stage_idx] <- stage_vals[use_stage_idx]
  }

  ann
}

build_color_map <- function(levels_vec, preferred_map = NULL, palette_name = "Dark 2") {
  levels_vec <- unique(levels_vec)
  levels_vec <- levels_vec[!is.na(levels_vec) & levels_vec != ""]
  if (length(levels_vec) == 0) return(setNames(character(0), character(0)))

  out <- setNames(rep(NA_character_, length(levels_vec)), levels_vec)

  if (!is.null(preferred_map)) {
    preferred_map <- preferred_map[!is.na(names(preferred_map)) & names(preferred_map) != ""]
    overlap <- intersect(levels_vec, names(preferred_map))
    out[overlap] <- preferred_map[overlap]
  }

  missing_lvls <- names(out)[is.na(out)]
  if (length(missing_lvls) > 0) {
    fallback <- grDevices::hcl.colors(length(missing_lvls), palette = palette_name)
    out[missing_lvls] <- fallback
  }

  out
}

normalize_stage_block <- function(x) {
  x <- as.character(x)
  dplyr::case_when(
    x %in% c("Suckling", "Suckling_piglets") ~ "Suckling_piglets",
    x %in% c("Weaning", "Fattening", "Older_pigs") ~ "Older_pigs",
    TRUE ~ x
  )
}

order_samples_by_blocks <- function(mat, ann_col) {
  block_order <- c("Non_farmer_controls", "Pig_farmers", "Older_pigs", "Suckling_piglets")
  ids <- colnames(mat)
  block_vals <- normalize_stage_block(ann_col[ids, "category"])

  out <- character(0)
  for (b in block_order) {
    block_ids <- ids[block_vals == b]
    if (length(block_ids) == 0) next
    if (length(block_ids) <= 2) {
      out <- c(out, block_ids)
    } else {
      hc <- hclust(dist(t(mat[, block_ids, drop = FALSE]), method = "euclidean"), method = "complete")
      out <- c(out, block_ids[hc$order])
    }
  }

  # append any labels not covered by block_order
  out <- c(out, setdiff(ids, out))
  out
}

make_col_order_callback <- function(desired_order) {
  desired_rank <- setNames(seq_along(desired_order), desired_order)
  function(hc, mat) {
    ord_w <- desired_rank[colnames(mat)]
    ord_w[is.na(ord_w)] <- max(ord_w, na.rm = TRUE) + seq_len(sum(is.na(ord_w)))
    dend <- stats::reorder(as.dendrogram(hc), wts = ord_w, agglo.FUN = mean)
    as.hclust(dend)
  }
}

ann_col_base <- build_ann_col(colnames(heatmap_log10))

sample_preferred <- NULL
if (exists("sample_type_colors") && !is.null(sample_type_colors)) {
  sample_preferred <- sample_type_colors
}

category_preferred <- NULL
if (exists("category2_colors") && !is.null(category2_colors)) {
  category_preferred <- c(category_preferred, category2_colors)
}
if (exists("pig_stage_colors") && !is.null(pig_stage_colors)) {
  category_preferred <- c(category_preferred, pig_stage_colors)
}

sample_ann_colors_local <- build_color_map(ann_col_base$sample_type, sample_preferred, "Set 2")
category_ann_colors_local <- build_color_map(ann_col_base$category, category_preferred, "Dark 2")



# Build annotations for all-pigs heatmap
ann_col_all_pigs <- build_ann_col(colnames(heatmap_log10))

row_annot_all_pigs <- tibble(Drug_Class_Resfinder = rownames(heatmap_log10)) %>%
  left_join(
    drugclass_mw_annotations_all_pigs %>%
      select(Drug_Class_Resfinder, pigs_vs_ctrl_mw, farmer_vs_ctrl_mw),
    by = "Drug_Class_Resfinder"
  ) %>%
  mutate(
    pigs_vs_ctrl_mw = replace_na(pigs_vs_ctrl_mw, "not_tested"),
    farmer_vs_ctrl_mw = replace_na(farmer_vs_ctrl_mw, "not_tested")
  ) %>%
  as.data.frame()

rownames(row_annot_all_pigs) <- row_annot_all_pigs$Drug_Class_Resfinder
row_annot_all_pigs$Drug_Class_Resfinder <- NULL

row_annot_all_pigs$pigs_vs_ctrl_mw <- factor(
  row_annot_all_pigs$pigs_vs_ctrl_mw,
  levels = c("sig_higher_in_all_pigs", "sig_higher_in_controls", "ns", "not_tested")
)

row_annot_all_pigs$farmer_vs_ctrl_mw <- factor(
  row_annot_all_pigs$farmer_vs_ctrl_mw,
  levels = c("sig_higher_in_pig_farmers", "sig_higher_in_controls", "ns", "not_tested")
)

pigs_sig_colors_all <- c(
  "sig_higher_in_all_pigs" = "#B2182B",
  "sig_higher_in_controls" = "#2166AC",
  "ns" = "#BDBDBD",
  "not_tested" = "#F0F0F0"
)

farmer_sig_colors_all <- c(
  "sig_higher_in_pig_farmers" = "#D6604D",
  "sig_higher_in_controls" = "#2166AC",
  "ns" = "#BDBDBD",
  "not_tested" = "#F0F0F0"
)

# Reorder columns by fixed group blocks with within-block clustering
ord_all <- order_samples_by_blocks(heatmap_log10, ann_col_all_pigs)
heatmap_log10_ordered <- heatmap_log10[, ord_all, drop = FALSE]
ann_col_all_pigs <- ann_col_all_pigs[colnames(heatmap_log10_ordered), , drop = FALSE]
cb_all <- make_col_order_callback(colnames(heatmap_log10_ordered))


library(viridis)
rng <- range(heatmap_log10_ordered, na.rm = TRUE)

breaks_low  <- seq(rng[1], 0, length.out = 10)   # coarse for -6..0
breaks_high <- seq(0, rng[2], length.out = 150)  # dense for 0..4
my_breaks   <- unique(c(breaks_low, breaks_high))

my_cols <- viridis(length(my_breaks) - 1, option = "C")

hm_width <- 15
hm_height <- max(8, nrow(heatmap_log10) * 0.25 + 3)

# 1) PDF export
pheatmap(
  heatmap_log10_ordered,
  color = my_cols,
  breaks = my_breaks,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  clustering_callback = cb_all,
  show_colnames = FALSE,
  annotation_col = ann_col_all_pigs,
  annotation_row = row_annot_all_pigs,
  annotation_colors = list(
    sample_type = sample_ann_colors_local,
    category = category_ann_colors_local,
    pigs_vs_ctrl_mw = pigs_sig_colors_all,
    farmer_vs_ctrl_mw = farmer_sig_colors_all
  ),
  main = "Drug Class x Samples (All pigs pooled vs controls)",
  fontsize_row = 10,
  filename = "../figures/heatmap_all_drugclasses_resfinder_all_pigs_stats.pdf",
  width = hm_width,
  height = hm_height
)

# 2) SVG export (DejaVu Sans)
if (requireNamespace("svglite", quietly = TRUE)) {
  svglite::svglite(
    "../figures/heatmap_all_drugclasses_resfinder_all_pigs_stats.svg",
    width = hm_width,
    height = hm_height,
    system_fonts = list(
      sans = "DejaVu Sans",
      serif = "DejaVu Sans",
      mono = "DejaVu Sans"
    )
  )
} else {
  grDevices::svg(
    "../figures/heatmap_all_drugclasses_resfinder_all_pigs_stats.svg",
    width = hm_width,
    height = hm_height,
    family = "DejaVu Sans"
  )
}

hm_svg <- pheatmap(
  heatmap_log10_ordered,
  color = my_cols,
  breaks = my_breaks,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  clustering_callback = cb_all,
  show_colnames = FALSE,
  annotation_col = ann_col_all_pigs,
  annotation_row = row_annot_all_pigs,
  annotation_colors = list(
    sample_type = sample_ann_colors_local,
    category = category_ann_colors_local,
    pigs_vs_ctrl_mw = pigs_sig_colors_all,
    farmer_vs_ctrl_mw = farmer_sig_colors_all
  ),
  main = "Drug Class x Samples (All pigs pooled vs controls)",
  fontsize_row = 8,
  silent = TRUE
)
grid::grid.newpage()
grid::grid.draw(hm_svg$gtable)
grDevices::dev.off()

write_csv(
  drugclass_mw_annotations_all_pigs,
  "../results/heatmap_all_drugclasses_resfinder_mw_all_pigs_stats.csv"
)

# Helper to compute MW stats for a specific pig group using category2
compute_mw_by_pig_group <- function(df_long, pig_group_name) {
  pig_group_var <- if ("category2" %in% colnames(df_long)) {
    "category2"
  } else if ("growth_stage" %in% colnames(df_long)) {
    "growth_stage"
  } else if ("grwoth_stage" %in% colnames(df_long)) {
    "grwoth_stage"
  } else {
    stop("No pig stage column found. Expected one of: category2, growth_stage, grwoth_stage")
  }

  df_long %>%
    filter(category %in% c("Pigs", farmer_level, control_level)) %>%
    group_by(Drug_Class_Resfinder) %>%
    summarise(
      p_pig_group_vs_ctrl = tryCatch(
        wilcox.test(cpm_resfinder[.data[[pig_group_var]] == pig_group_name],
                    cpm_resfinder[category == control_level])$p.value,
        error = function(e) NA_real_
      ),
      median_pig_group = median(cpm_resfinder[.data[[pig_group_var]] == pig_group_name], na.rm = TRUE),
      median_ctrl = median(cpm_resfinder[category == control_level], na.rm = TRUE),
      p_farmer_vs_ctrl = tryCatch(
        wilcox.test(cpm_resfinder[category == farmer_level],
                    cpm_resfinder[category == control_level])$p.value,
        error = function(e) NA_real_
      ),
      median_farmer = median(cpm_resfinder[category == farmer_level], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      p_adj_pig_vs_ctrl = p.adjust(p_pig_group_vs_ctrl, method = "BH"),
      p_adj_farmer_vs_ctrl = p.adjust(p_farmer_vs_ctrl, method = "BH"),
      pig_vs_ctrl_mw = case_when(
        is.na(p_adj_pig_vs_ctrl) ~ "not_tested",
        p_adj_pig_vs_ctrl < 0.05 & median_pig_group > median_ctrl ~ paste0("sig_higher_in_", pig_group_name),
        p_adj_pig_vs_ctrl < 0.05 & median_pig_group < median_ctrl ~ "sig_higher_in_controls",
        TRUE ~ "ns"
      ),
      farmer_vs_ctrl_mw = case_when(
        is.na(p_adj_farmer_vs_ctrl) ~ "not_tested",
        p_adj_farmer_vs_ctrl < 0.05 & median_farmer > median_ctrl ~ "sig_higher_in_pig_farmers",
        p_adj_farmer_vs_ctrl < 0.05 & median_farmer < median_ctrl ~ "sig_higher_in_controls",
        TRUE ~ "ns"
      )
    )
}

# Helper to build and plot heatmap for a pig group
plot_heatmap_by_pig_group <- function(df_long, pig_group_name, fig_suffix, title_text) {
  mw_stats <- compute_mw_by_pig_group(df_long, pig_group_name)
  
  # Build column and row annotations
  ann_col_rf <- build_ann_col(colnames(heatmap_log10))
  
  row_annot_rf <- tibble(Drug_Class_Resfinder = rownames(heatmap_log10)) %>%
    left_join(
      mw_stats %>%
        select(Drug_Class_Resfinder, pig_vs_ctrl_mw, farmer_vs_ctrl_mw),
      by = "Drug_Class_Resfinder"
    ) %>%
    mutate(
      pig_vs_ctrl_mw = replace_na(pig_vs_ctrl_mw, "not_tested"),
      farmer_vs_ctrl_mw = replace_na(farmer_vs_ctrl_mw, "not_tested")
    ) %>%
    as.data.frame()
  
  rownames(row_annot_rf) <- row_annot_rf$Drug_Class_Resfinder
  row_annot_rf$Drug_Class_Resfinder <- NULL
  
  pig_sig_label <- paste0("sig_higher_in_", pig_group_name)
  row_annot_rf$pig_vs_ctrl_mw <- factor(
    row_annot_rf$pig_vs_ctrl_mw,
    levels = c(pig_sig_label, "sig_higher_in_controls", "ns", "not_tested")
  )
  
  row_annot_rf$farmer_vs_ctrl_mw <- factor(
    row_annot_rf$farmer_vs_ctrl_mw,
    levels = c("sig_higher_in_pig_farmers", "sig_higher_in_controls", "ns", "not_tested")
  )
  
  pig_sig_colors_rf <- c(
    setNames("#B2182B", pig_sig_label),
    "sig_higher_in_controls" = "#2166AC",
    "ns" = "#BDBDBD",
    "not_tested" = "#F0F0F0"
  )
  
  farmer_sig_colors_rf <- c(
    "sig_higher_in_pig_farmers" = "#D6604D",
    "sig_higher_in_controls" = "#2166AC",
    "ns" = "#BDBDBD",
    "not_tested" = "#F0F0F0"
  )
  
  # Reorder columns by fixed group blocks with within-block clustering
  ord_rf <- order_samples_by_blocks(heatmap_log10, ann_col_rf)
  hm_ordered <- heatmap_log10[, ord_rf, drop = FALSE]
  ann_col_rf <- ann_col_rf[colnames(hm_ordered), , drop = FALSE]
  cb_rf <- make_col_order_callback(colnames(hm_ordered))
  
  ann_colors_list <- list(
    sample_type = sample_ann_colors_local,
    category = category_ann_colors_local,
    farmer_vs_ctrl_mw = farmer_sig_colors_rf
  )
  ann_colors_list[["pig_vs_ctrl_mw"]] <- pig_sig_colors_rf
  
  # Plot
  hm_width <- 15
  hm_height <- max(8, nrow(heatmap_log10) * 0.25 + 3)

  # 1) PDF export (same style as the all-pigs heatmap)
  pheatmap(
    hm_ordered,
    color = my_cols,
    breaks = my_breaks,
    cluster_rows = FALSE,
    cluster_cols = TRUE,
    clustering_callback = cb_rf,
    show_colnames = FALSE,
    annotation_col = ann_col_rf,
    annotation_row = row_annot_rf,
    annotation_colors = ann_colors_list,
    main = title_text,
    fontsize_row = 8,
    filename = paste0("../figures/heatmap_all_drugclasses_resfinder_", fig_suffix, ".pdf"),
    width = hm_width,
    height = hm_height
  )

  # 2) SVG export (same font handling as the all-pigs heatmap)
  svg_file <- paste0("../figures/heatmap_all_drugclasses_resfinder_", fig_suffix, ".svg")
  if (requireNamespace("svglite", quietly = TRUE)) {
    svglite::svglite(
      svg_file,
      width = hm_width,
      height = hm_height,
      system_fonts = list(
        sans = "DejaVu Sans",
        serif = "DejaVu Sans",
        mono = "DejaVu Sans"
      )
    )
  } else {
    grDevices::svg(
      svg_file,
      width = hm_width,
      height = hm_height,
      family = "DejaVu Sans"
    )
  }

  hm_svg <- pheatmap(
    hm_ordered,
    color = my_cols,
    breaks = my_breaks,
    cluster_rows = FALSE,
    cluster_cols = TRUE,
    clustering_callback = cb_rf,
    show_colnames = FALSE,
    annotation_col = ann_col_rf,
    annotation_row = row_annot_rf,
    annotation_colors = ann_colors_list,
    main = title_text,
    fontsize_row = 8,
    silent = TRUE
  )
  grid::grid.newpage()
  grid::grid.draw(hm_svg$gtable)
  grDevices::dev.off()
  
  write_csv(
    mw_stats,
    paste0("../results/heatmap_all_drugclasses_resfinder_mw_", fig_suffix, ".csv")
  )
  
  return(mw_stats)
}

# Generate heatmaps for Suckling_piglets and Older_pigs
mw_suckling <- plot_heatmap_by_pig_group(
  cpm_resfinder_long,
  pig_group_name = "Suckling_piglets",
  fig_suffix = "suckling_pigs_stats",
  title_text = "drug classes x all samples (Suckling piglets vs controls)"
)

mw_older <- plot_heatmap_by_pig_group(
  cpm_resfinder_long,
  pig_group_name = "Older_pigs",
  fig_suffix = "older_pigs_stats",
  title_text = "drug classes x all samples (Older pigs vs controls)"
)

```

### abundance barplots with Mann-Whitney-U statistics

```{r}
# Drug_Class_Resfinder abundance boxplots with pairwise Mann-Whitney tests.
plot_resfinder_boxplots_pdf <- function(df_long, group_var, pdf_file, stats_file) {
  stopifnot(group_var %in% colnames(df_long))

  resfinder_classes <- df_long %>%
    dplyr::count(Drug_Class_Resfinder, sort = TRUE) %>%
    pull(Drug_Class_Resfinder)

  if (length(resfinder_classes) == 0) {
    warning(paste0("No Drug_Class_Resfinder entries available for ", group_var, " boxplot PDF."))
    return(invisible(NULL))
  }

  all_wilcox <- list()
  grDevices::cairo_pdf(pdf_file, width = 6.5, height = 7.5, onefile = TRUE)

  for (dc in resfinder_classes) {
    df_plot <- df_long %>%
      filter(Drug_Class_Resfinder == dc) %>%
      mutate(
        group_label = as.character(.data[[group_var]]),
        log10_cpm = log10(cpm_resfinder + 1e-6)
      ) %>%
      filter(!is.na(group_label), group_label != "", is.finite(log10_cpm))

    if (group_var == "category2") {
      requested_order <- c("Non_farmer_controls", "Pig_farmers", "Suckling_piglets", "Older_pigs")
      other_levels <- setdiff(unique(df_plot$group_label), requested_order)
      df_plot <- df_plot %>%
        mutate(group_label = factor(group_label, levels = c(requested_order, sort(other_levels))))
    } else if (group_var == "category") {
      requested_order <- c("Non_farmer_controls", "Pig_farmers", "Pigs")
      other_levels <- setdiff(unique(df_plot$group_label), requested_order)
      df_plot <- df_plot %>%
        mutate(group_label = factor(group_label, levels = c(requested_order, sort(other_levels))))
    }

    if (nrow(df_plot) == 0 || dplyr::n_distinct(df_plot$group_label) < 2) {
      next
    }

    wilcox_tbl <- tryCatch(
      {
        df_plot %>%
          rstatix::pairwise_wilcox_test(log10_cpm ~ group_label, p.adjust.method = "BH") %>%
          rstatix::add_xy_position(x = "group_label", fun = "max", step.increase = 0.2) %>%
          mutate(
            Drug_Class_Resfinder = dc,
            grouping = group_var,
            p.adj.formatted = ifelse(p.adj < 0.001, 
                                     sprintf("%.2e", p.adj), 
                                     sprintf("%.4f", p.adj))
          )
      },
      error = function(e) {
        tibble()
      }
    )

    fill_colors <- if (group_var == "category2") {
      unlist(color_codes$category2)
    } else if (group_var == "category" && "sample_type_colors" %in% names(color_codes)) {
      unlist(color_codes$sample_type_colors)
    } else {
      sample_type_colors
    }

    missing_cols <- setdiff(as.character(unique(df_plot$group_label)), names(fill_colors))
    if (length(missing_cols) > 0) {
      extra_cols <- grDevices::hcl.colors(length(missing_cols), palette = "Set 2")
      names(extra_cols) <- missing_cols
      fill_colors <- c(fill_colors, extra_cols)
    }

    p <- ggplot(df_plot, aes(x = group_label, y = log10_cpm, fill = group_label)) +
      geom_boxplot(outlier.shape = NA, alpha = 0.75) +
      geom_jitter(width = 0.15, size = 1.3, alpha = 0.65) +
      labs(
        title = paste("Drug class (ResFinder):", dc),
        subtitle = paste("Pairwise Mann-Whitney (BH-adjusted) | grouped by", group_var),
        x = group_var,
        y = "log10(CPM + 1e-6)"
      ) +
      theme_bw(base_family = "DejaVu Sans", base_size = FIG_BASE_SIZE) +
      theme(
        plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(),
        legend.position = "none"
      )+
      scale_fill_manual(values = fill_colors, drop = FALSE)

    if (nrow(wilcox_tbl) > 0) {
      p <- p +
        stat_pvalue_manual(
          wilcox_tbl,
          label = "p.adj.formatted",
          xmin = "group1",
          xmax = "group2",
          y.position = "y.position",
          tip.length = 0.01,
          bracket.size = 0.3,
          size = 2.5
        )

      all_wilcox[[length(all_wilcox) + 1]] <- wilcox_tbl
    }

    print(p)
  }

  grDevices::dev.off()

  if (length(all_wilcox) > 0) {
    wilcox_all_tbl <- bind_rows(all_wilcox) %>%
      select(grouping, Drug_Class_Resfinder, group1, group2, n1, n2, p, p.adj, p.adj.signif)
    write_csv(wilcox_all_tbl, stats_file)
  }
}

# 1) PDF grouped by category: Pigs, Pig_farmers, Non_farmer_controls
plot_resfinder_boxplots_pdf(
  df_long = cpm_resfinder_long,
  group_var = "category",
  pdf_file = "../figures/boxplots_resfinder_abundance_by_category.pdf",
  stats_file = "../results/resfinder_mannwhitney_pairwise_by_category.csv"
)

# 2) PDF grouped by category2: separates Older_pigs and Suckling_piglets
plot_resfinder_boxplots_pdf(
  df_long = cpm_resfinder_long,
  group_var = "category2",
  pdf_file = "../figures/boxplots_resfinder_abundance_by_category2.pdf",
  stats_file = "../results/resfinder_mannwhitney_pairwise_by_category2.csv"
)

```


### Supplementary multipanel SVG pages (9 panels per page)

```{r}
# Build thesis-ready multipanel pages with fixed y-axis across panels.
plot_resfinder_multipanel_svg_pages <- function(df_long, group_var, out_prefix, panels_per_page = 9) {
  stopifnot(group_var %in% colnames(df_long))

  resfinder_classes <- df_long %>%
    dplyr::count(Drug_Class_Resfinder, sort = TRUE) %>%
    pull(Drug_Class_Resfinder)

  if (length(resfinder_classes) == 0) {
    warning(paste0("No Drug_Class_Resfinder entries available for ", group_var, " multipanel SVG export."))
    return(invisible(NULL))
  }

  # Match grouping order from the PDF function.
  if (group_var == "category2") {
    requested_order <- c("Non_farmer_controls", "Pig_farmers", "Suckling_piglets", "Older_pigs")
  } else if (group_var == "category") {
    requested_order <- c("Non_farmer_controls", "Pig_farmers", "Pigs")
  } else {
    requested_order <- sort(unique(as.character(df_long[[group_var]])))
  }

  # Same fill-color logic as above.
  fill_colors <- if (group_var == "category2") {
    unlist(color_codes$category2)
  } else if (group_var == "category" && "sample_type_colors" %in% names(color_codes)) {
    unlist(color_codes$sample_type_colors)
  } else {
    sample_type_colors
  }

  all_levels <- sort(unique(as.character(df_long[[group_var]])))
  missing_cols <- setdiff(all_levels, names(fill_colors))
  if (length(missing_cols) > 0) {
    extra_cols <- grDevices::hcl.colors(length(missing_cols), palette = "Set 2")
    names(extra_cols) <- missing_cols
    fill_colors <- c(fill_colors, extra_cols)
  }

  plot_list <- list()
  wilcox_all <- list()
  y_max_vals <- c()
  y_min_vals <- c()

  for (dc in resfinder_classes) {
    df_plot <- df_long %>%
      filter(Drug_Class_Resfinder == dc) %>%
      mutate(
        group_label = as.character(.data[[group_var]]),
        log10_cpm = log10(cpm_resfinder + 1e-6)
      ) %>%
      filter(!is.na(group_label), group_label != "", is.finite(log10_cpm))

    if (nrow(df_plot) == 0 || dplyr::n_distinct(df_plot$group_label) < 2) {
      next
    }

    other_levels <- setdiff(unique(df_plot$group_label), requested_order)
    level_order <- c(requested_order, sort(other_levels))
    df_plot <- df_plot %>%
      mutate(group_label = factor(group_label, levels = level_order))

    wilcox_tbl <- tryCatch(
      {
        df_plot %>%
          rstatix::pairwise_wilcox_test(log10_cpm ~ group_label, p.adjust.method = "BH") %>%
          rstatix::add_xy_position(x = "group_label", fun = "max", step.increase = 0.2) %>%
          mutate(
            Drug_Class_Resfinder = dc,
            grouping = group_var,
            p.adj.formatted = ifelse(
              p.adj < 0.001,
              sprintf("%.2e", p.adj),
              sprintf("%.4f", p.adj)
            )
          )
      },
      error = function(e) {
        tibble()
      }
    )

    if (nrow(wilcox_tbl) > 0) {
      y_max_vals <- c(y_max_vals, max(wilcox_tbl$y.position, na.rm = TRUE))
      wilcox_all[[length(wilcox_all) + 1]] <- wilcox_tbl
    }
    y_max_vals <- c(y_max_vals, max(df_plot$log10_cpm, na.rm = TRUE))
    y_min_vals <- c(y_min_vals, min(df_plot$log10_cpm, na.rm = TRUE))

    p <- ggplot(df_plot, aes(x = group_label, y = log10_cpm, fill = group_label)) +
      geom_boxplot(outlier.shape = NA, alpha = 0.75) +
      geom_jitter(width = 0.15, size = 1.0, alpha = 0.6) +
      labs(
        title = dc,
        x = group_var,
        y = "log10(CPM + 1e-6)"
      ) +
      theme_bw(base_family = "DejaVu Sans", base_size = FIG_BASE_SIZE) +
      scale_x_discrete(labels = color_codes$label_mapping)+
      theme(
        plot.title = element_text(hjust = 0.5, size = 10),
        axis.text.x = element_text(size = 10, angle = 0, hjust = 0),
        axis.title = element_text(size = 9),
        axis.title.x = element_blank(),
        legend.position = "none"
      ) +
      scale_fill_manual(values = fill_colors, drop = FALSE)

    if (nrow(wilcox_tbl) > 0) {
      p <- p +
        stat_pvalue_manual(
          wilcox_tbl,
          label = "p.adj.formatted",
          xmin = "group1",
          xmax = "group2",
          y.position = "y.position",
          tip.length = 0.01,
          bracket.size = 0.25,
          size = 2.2,
          family = "DejaVu Sans"
        )
    }

    plot_list[[dc]] <- p
  }

  if (length(plot_list) == 0) {
    warning(paste0("No plottable Drug_Class_Resfinder entries for ", group_var, "."))
    return(invisible(NULL))
  }

  # Fixed y-axis for all panels/pages.
  y_min <- floor(min(y_min_vals, na.rm = TRUE) * 10) / 10
  y_max <- ceiling(max(y_max_vals, na.rm = TRUE) * 10) / 10
  plot_list <- lapply(plot_list, function(p) p + coord_cartesian(ylim = c(y_min, y_max)))

  page_index <- split(seq_along(plot_list), ceiling(seq_along(plot_list) / panels_per_page))

  for (i in seq_along(page_index)) {
    idx <- page_index[[i]]
    page_plots <- plot_list[idx]

    page_plot <- ggpubr::ggarrange(
      plotlist = page_plots,
      ncol = 3,
      nrow = 3,
      align = "hv"
    )

    out_file <- sprintf("%s_page_%02d.svg", out_prefix, i)
    ggsave(
      filename = out_file,
      plot = page_plot,
      width = 12,
      height = 11,
      dpi = 300,
      bg = "white",
      web_font ="DejaVu Sans"
    )
  }

  if (length(wilcox_all) > 0) {
    wilcox_all_tbl <- bind_rows(wilcox_all) %>%
      select(grouping, Drug_Class_Resfinder, group1, group2, n1, n2, p, p.adj, p.adj.signif)
    stats_out <- sub("\\.svg$", "_stats.csv", out_prefix)
    write_csv(wilcox_all_tbl, stats_out)
  }

  message("Saved multipanel SVG pages for ", group_var,
          " with fixed y-axis [", y_min, ", ", y_max, "]")
}

# category pages (9 panels per page)
plot_resfinder_multipanel_svg_pages(
  df_long = cpm_resfinder_long,
  group_var = "category",
  out_prefix = "../figures/boxplots_resfinder_abundance_by_category_multipanel",
  panels_per_page = 9
)

# category2 pages (9 panels per page)
plot_resfinder_multipanel_svg_pages(
  df_long = cpm_resfinder_long,
  group_var = "category2",
  out_prefix = "../figures/boxplots_resfinder_abundance_by_category2_multipanel",
  panels_per_page = 9
)
```

### Scatterplot: Suckling vs Older significance (Mann-Whitney-U)

```{r}
# Compare stage-specific significance across drug classes.
mw_safe <- function(x, y) {
  if (sum(!is.na(x)) < 2 || sum(!is.na(y)) < 2) {
    return(NA_real_)
  }
  tryCatch(wilcox.test(x, y)$p.value, error = function(e) NA_real_)
}

log2_median_ratio <- function(x, y, pseudo = 1e-6) {
  mx <- median(x, na.rm = TRUE)
  my <- median(y, na.rm = TRUE)
  if (!is.finite(mx) || !is.finite(my)) {
    return(NA_real_)
  }
  log2((mx + pseudo) / (my + pseudo))
}

resfinder_stage_tbl <- cpm_resfinder_long %>%
  mutate(
    stage_group = case_when(
      category2 == "Suckling_piglets" ~ "Suckling_piglets",
      category2 == "Older_pigs" ~ "Older_pigs",
      category == farmer_level ~ "Pig_farmers",
      category == "Non_farmer_controls" ~ "Non_farmer_controls",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(stage_group))

stage_mw_stats <- resfinder_stage_tbl %>%
  group_by(Drug_Class_Resfinder) %>%
  summarise(
    p_suckling_vs_ctrl = mw_safe(
      cpm_resfinder[stage_group == "Suckling_piglets"],
      cpm_resfinder[stage_group == "Non_farmer_controls"]
    ),
    p_older_vs_ctrl = mw_safe(
      cpm_resfinder[stage_group == "Older_pigs"],
      cpm_resfinder[stage_group == "Non_farmer_controls"]
    ),
    p_suckling_vs_farmers = mw_safe(
      cpm_resfinder[stage_group == "Suckling_piglets"],
      cpm_resfinder[stage_group == "Pig_farmers"]
    ),
    p_older_vs_farmers = mw_safe(
      cpm_resfinder[stage_group == "Older_pigs"],
      cpm_resfinder[stage_group == "Pig_farmers"]
    ),
    l2mr_suckling_vs_ctrl = log2_median_ratio(
      cpm_resfinder[stage_group == "Suckling_piglets"],
      cpm_resfinder[stage_group == "Non_farmer_controls"]
    ),
    l2mr_older_vs_ctrl = log2_median_ratio(
      cpm_resfinder[stage_group == "Older_pigs"],
      cpm_resfinder[stage_group == "Non_farmer_controls"]
    ),
    l2mr_suckling_vs_farmers = log2_median_ratio(
      cpm_resfinder[stage_group == "Suckling_piglets"],
      cpm_resfinder[stage_group == "Pig_farmers"]
    ),
    l2mr_older_vs_farmers = log2_median_ratio(
      cpm_resfinder[stage_group == "Older_pigs"],
      cpm_resfinder[stage_group == "Pig_farmers"]
    ),
    .groups = "drop"
  ) %>%
  mutate(
    p_adj_suckling_vs_ctrl = p.adjust(p_suckling_vs_ctrl, method = "BH"),
    p_adj_older_vs_ctrl = p.adjust(p_older_vs_ctrl, method = "BH"),
    p_adj_suckling_vs_farmers = p.adjust(p_suckling_vs_farmers, method = "BH"),
    p_adj_older_vs_farmers = p.adjust(p_older_vs_farmers, method = "BH")
  )

scatter_stage_sig <- bind_rows(
  stage_mw_stats %>%
    transmute(
      Drug_Class_Resfinder,
      reference_group = "vs Non-farmer controls",
      p_adj_suckling = p_adj_suckling_vs_ctrl,
      p_adj_older = p_adj_older_vs_ctrl,
      l2mr_suckling = l2mr_suckling_vs_ctrl,
      l2mr_older = l2mr_older_vs_ctrl
    ),
  stage_mw_stats %>%
    transmute(
      Drug_Class_Resfinder,
      reference_group = "vs Pig farmers",
      p_adj_suckling = p_adj_suckling_vs_farmers,
      p_adj_older = p_adj_older_vs_farmers,
      l2mr_suckling = l2mr_suckling_vs_farmers,
      l2mr_older = l2mr_older_vs_farmers
    )
) %>%
  mutate(
    x_sig = -log10(p_adj_suckling + 1e-300),
    y_sig = -log10(p_adj_older + 1e-300),
    max_sig = pmax(x_sig, y_sig, na.rm = TRUE),
    sig_pattern = case_when(
      p_adj_suckling < 0.05 & p_adj_older < 0.05 ~ "both_sig",
      p_adj_suckling < 0.05 & (is.na(p_adj_older) | p_adj_older >= 0.05) ~ "suckling_only",
      p_adj_older < 0.05 & (is.na(p_adj_suckling) | p_adj_suckling >= 0.05) ~ "older_only",
      TRUE ~ "none_sig"
    ),
    effect_agreement = case_when(
      is.na(l2mr_suckling) | is.na(l2mr_older) ~ "missing_effect",
      sign(l2mr_suckling) == sign(l2mr_older) ~ "same_direction",
      TRUE ~ "opposite_direction"
    )
  )

label_df <- scatter_stage_sig %>%
  group_by(reference_group) %>%
  slice_max(order_by = max_sig, n = 10, with_ties = FALSE) %>%
  ungroup()

sig_cut <- -log10(0.05)

p_stage_sig <- ggplot(
  scatter_stage_sig,
  aes(x = x_sig, y = y_sig, color = sig_pattern, shape = effect_agreement)
) +
  geom_vline(xintercept = sig_cut, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  geom_hline(yintercept = sig_cut, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  geom_point(alpha = 0.85, size = 2.2) +
  ggrepel::geom_text_repel(
    data = label_df,
    aes(label = Drug_Class_Resfinder),
    size = 2.8,
    max.overlaps = 40,
    box.padding = 0.25,
    point.padding = 0.2,
    segment.alpha = 0.5,
    show.legend = FALSE
  ) +
  facet_wrap(~ reference_group, ncol = 2) +
  scale_color_manual(
    values = c(
      "both_sig" = "#B2182B",
      "suckling_only" = "#EF8A62",
      "older_only" = "#67A9CF",
      "none_sig" = "#BDBDBD"
    ),
    name = "BH-significance pattern"
  ) +
  scale_shape_manual(
    values = c(
      "same_direction" = 16,
      "opposite_direction" = 17,
      "missing_effect" = 4
    ),
    name = "Effect direction"
  ) +
  labs(
    title = "Suckling vs Older: significance concordance by drug class",
    subtitle = "x = -log10(BH p) Suckling comparison, y = -log10(BH p) Older comparison",
    x = "-log10(BH p): Suckling",
    y = "-log10(BH p): Older"
  ) +
  theme_bw(base_family = "DejaVu Sans", base_size = FIG_BASE_SIZE) +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(face = "bold")
  )

print(p_stage_sig)

write_csv(stage_mw_stats, "../results/resfinder_stage_pairwise_mw_stats.csv")
write_csv(scatter_stage_sig, "../results/resfinder_stage_sig_scatter_input.csv")

ggsave(
  "../figures/resfinder_stage_sig_scatterplot.svg",
  p_stage_sig,
  width = 12,
  height = 6.8,
  web_fonts = "DejaVu Sans"
)
```

### Interpretation

-   Data basis: sample-level abundance table `cpm_resfinder_long` (manual Drug_Class_Resfinder annotation), excluding missing category/class/value entries where needed.
-   Global category differences: Kruskal-Wallis test per drug class, followed by Benjamini-Hochberg multiple-testing correction.
-   Pairwise group interpretation: Wilcoxon rank-sum tests between predefined groups (for example Pigs vs Pig_farmers, stage-specific contrasts), with BH-adjusted p-values.
-   Effect size direction: log2 median ratio between groups to classify classes as higher in group A, higher in group B, or similar.
-   Similarity rule (Pigs vs Pig_farmers): class labeled as same when BH p \>= 0.05 and absolute log2 median ratio \<= 0.5.
-   Reporting outputs: ranked result tables exported to results and visual summaries (barplots/scatter with thresholds and labels) exported to figures.

#### Top 3 most abundant manually annotated drug classes per category

```{r}
# Use the final manual annotation object (Drug_Class_Resfinder).
# We test each class across all categories with Kruskal-Wallis,
# BH-correct p-values, and report top 3 classes per category by median CPM.

resfinder_kw <- cpm_resfinder_long %>%
  filter(!is.na(category), !is.na(Drug_Class_Resfinder), !is.na(cpm_resfinder)) %>%
  group_by(Drug_Class_Resfinder) %>%
  summarise(
    n_samples = n(),
    n_categories = n_distinct(category),
    kw_p = if (n_categories >= 2) {
      tryCatch(
        stats::kruskal.test(cpm_resfinder ~ category)$p.value,
        error = function(e) NA_real_
      )
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  mutate(kw_p_adj = p.adjust(kw_p, method = "BH"))

top3_resfinder_by_category <- cpm_resfinder_long %>%
  filter(!is.na(category), !is.na(Drug_Class_Resfinder), !is.na(cpm_resfinder)) %>%
  group_by(category, Drug_Class_Resfinder) %>%
  summarise(
    median_cpm = median(cpm_resfinder, na.rm = TRUE),
    mean_cpm = mean(cpm_resfinder, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(resfinder_kw %>% select(Drug_Class_Resfinder, kw_p, kw_p_adj), by = "Drug_Class_Resfinder") %>%
  group_by(category) %>%
  arrange(desc(median_cpm), .by_group = TRUE) %>%
  slice_head(n = 3) %>%
  ungroup() %>%
  arrange(category, desc(median_cpm))

top3_resfinder_by_category_print <- top3_resfinder_by_category %>%
  mutate(
    median_cpm = round(median_cpm, 2),
    mean_cpm = round(mean_cpm, 2),
    kw_p = signif(kw_p, 3),
    kw_p_adj = signif(kw_p_adj, 3)
  )

print(top3_resfinder_by_category_print, n = Inf)

write_csv(
  top3_resfinder_by_category,
  "../results/top3_resfinder_drugclasses_by_category.csv"
)

# Plot: median CPM bars with IQR error bars and optional sample-level jitter
top3_resfinder_samples <- top3_resfinder_by_category %>%
  select(category, Drug_Class_Resfinder) %>%
  distinct() %>%
  inner_join(
    cpm_resfinder_long %>%
      select(category, Drug_Class_Resfinder, sample, cpm_resfinder),
    by = c("category", "Drug_Class_Resfinder")
  )

top3_resfinder_iqr <- top3_resfinder_samples %>%
  group_by(category, Drug_Class_Resfinder) %>%
  summarise(
    median_cpm = median(cpm_resfinder, na.rm = TRUE),
    q1_cpm = quantile(cpm_resfinder, 0.25, na.rm = TRUE),
    q3_cpm = quantile(cpm_resfinder, 0.75, na.rm = TRUE),
    .groups = "drop"
  )


plot_top3_resfinder <- ggplot(
  top3_resfinder_iqr,
  aes(
    x = forcats::fct_reorder(Drug_Class_Resfinder, median_cpm, .desc = TRUE),
    y = median_cpm
  )
) +
  geom_col(fill = "grey55", alpha = 0.9) +
  geom_errorbar(aes(ymin = q1_cpm, ymax = q3_cpm), width = 0.22, linewidth = 0.45) +
  geom_jitter(
    data = top3_resfinder_samples,
    aes(
      x = forcats::fct_reorder(Drug_Class_Resfinder, cpm_resfinder, .fun = median, .desc = TRUE),
      y = cpm_resfinder
    ),
    width = 0.13,
    height = 0,
    size = 1.5,
    alpha = 0.45,
    color = "black",
    inherit.aes = FALSE
  ) +
  facet_wrap(
  ~ category,
  scales = "free_x",
  labeller = labeller(category = as_labeller(
    function(x) ifelse(x %in% names(color_codes$label_mapping),
                       color_codes$label_mapping[x],
                       x)
  ))
) +
  labs(
    title = "Top 3 drug classes by category",
    subtitle = "Bars = median CPM, error bars = IQR (Q1-Q3), points = samples",
    x = "Drug class",
    y = "CPM"
  ) +
  theme_bw(base_family = "DejaVu Sans", base_size = FIG_BASE_SIZE) +
  theme_publication()+
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
  

print(plot_top3_resfinder)


ggsave(
  "../figures/top3_resfinder_drugclasses_by_category.svg",
  plot_top3_resfinder,
  width = 8,
  height = 6,
  web_fonts = "DejaVu Sans"
)
```

#### Pigs have more of compared to human/farmer

```{r}
# Practical interpretation workflow:
# 1) one compact stats table per contrast
# 2) one ranking plot (top effects among BH-significant classes)

compare_resfinder_groups <- function(df, group_var, group_a, group_b, out_stub) {
  stopifnot(group_var %in% colnames(df))

  tbl <- df %>%
    filter(!is.na(.data[[group_var]]), .data[[group_var]] %in% c(group_a, group_b)) %>%
    group_by(Drug_Class_Resfinder) %>%
    summarise(
      n_a = sum(.data[[group_var]] == group_a),
      n_b = sum(.data[[group_var]] == group_b),
      median_a = median(cpm_resfinder[.data[[group_var]] == group_a], na.rm = TRUE),
      median_b = median(cpm_resfinder[.data[[group_var]] == group_b], na.rm = TRUE),
      p_value = tryCatch(
        wilcox.test(
          cpm_resfinder[.data[[group_var]] == group_a],
          cpm_resfinder[.data[[group_var]] == group_b]
        )$p.value,
        error = function(e) NA_real_
      ),
      .groups = "drop"
    ) %>%
    mutate(
      padj = p.adjust(p_value, method = "BH"),
      log2_median_ratio = log2((median_a + 1e-6) / (median_b + 1e-6)),
      direction = case_when(
        is.na(padj) ~ "not_tested",
        padj < 0.05 & log2_median_ratio > 0 ~ paste0("higher_in_", group_a),
        padj < 0.05 & log2_median_ratio < 0 ~ paste0("higher_in_", group_b),
        TRUE ~ "ns"
      )
    ) %>%
    arrange(padj, desc(abs(log2_median_ratio)))

  write_csv(tbl, paste0("../results/interpretation_", out_stub, ".csv"))

  plot_tbl <- tbl %>%
    filter(!is.na(padj), padj < 0.05) %>%
    slice_head(n = 20)

  p <- ggplot(plot_tbl, aes(x = reorder(Drug_Class_Resfinder, log2_median_ratio), y = log2_median_ratio, fill = direction)) +
    geom_col(alpha = 0.9) +
    coord_flip() +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    labs(
      title = paste0(group_a, " vs ", group_b),
      subtitle = "Top BH-significant classes by effect size",
      x = "Drug class",
      y = "log2 median ratio"
    ) +
    theme_bw(base_family = "DejaVu Sans", base_size = FIG_BASE_SIZE) +
    theme(legend.position = "bottom")+
    theme_publication_with_title()

  print(head(tbl, 20))
  print(p)

  ggsave(
    paste0("../figures/interpretation_", out_stub, ".svg"),
    p,
    width = 10,
    height = 7,
    web_fonts = "DejaVu Sans"
  )

  tbl
}

# Pigs vs Pig farmers
interp_pigs_vs_farmers <- compare_resfinder_groups(
  cpm_resfinder_long,
  group_var = "category",
  group_a = "Pigs",
  group_b = farmer_level,
  out_stub = "pigs_vs_pig_farmers"
)

# Pigs vs controls
interp_pigs_vs_controls <- compare_resfinder_groups(
  cpm_resfinder_long,
  group_var = "category",
  group_a = "Pigs",
  group_b = "Non_farmer_controls",
  out_stub = "pigs_vs_controls"
)
```

#### suckling pigs have more /less drugs then older pigs

```{r}
if ("category2" %in% colnames(cpm_resfinder_long)) {
  interp_suckling_vs_older <- compare_resfinder_groups(
    cpm_resfinder_long,
    group_var = "category2",
    group_a = "Suckling_piglets",
    group_b = "Older_pigs",
    out_stub = "suckling_vs_older"
  )
} else {
  message("Column 'category2' is missing; cannot run suckling vs older contrast.")
}
```

#### controls have sig more of

```{r}
# Collect classes that are significantly higher in controls.
controls_higher_tbl <- bind_rows(
  interp_pigs_vs_controls %>%
    mutate(contrast = "Pigs_vs_controls"),
  compare_resfinder_groups(
    cpm_resfinder_long,
    group_var = "category",
    group_a = farmer_level,
    group_b = "Non_farmer_controls",
    out_stub = "pig_farmers_vs_controls"
  ) %>%
    mutate(contrast = "Pig_farmers_vs_controls")
) %>%
  filter(direction == "higher_in_Non_farmer_controls") %>%
  arrange(padj)

print(controls_higher_tbl, n = Inf)
write_csv(controls_higher_tbl, "../results/interpretation_controls_higher_classes.csv")
```

#### suckling/oder pigs and farmer have sig more of than controls

Interpretation: This section identifies Drug_Class_Resfinder classes that are significantly more abundant in exposed groups (Suckling piglets, Older pigs, and Pig farmers) than in Non-farmer controls, based on Mann-Whitney tests with BH correction. Classes retained in `exposed_higher_vs_controls` represent a shared exposure-associated AMR signature, while contrast-specific hits indicate age-stage or occupation-specific enrichment patterns.

```{r}
# Exposure groups higher than controls in each contrast.
interp_farmers_vs_controls <- compare_resfinder_groups(
  cpm_resfinder_long,
  group_var = "category",
  group_a = farmer_level,
  group_b = "Non_farmer_controls",
  out_stub = "pig_farmers_vs_controls"
)

if (exists("interp_suckling_vs_older")) {
  interp_suckling_vs_controls <- compare_resfinder_groups(
    cpm_resfinder_long,
    group_var = "category2",
    group_a = "Suckling_piglets",
    group_b = "Non_farmer_controls",
    out_stub = "suckling_vs_controls"
  )

  interp_older_vs_controls <- compare_resfinder_groups(
    cpm_resfinder_long,
    group_var = "category2",
    group_a = "Older_pigs",
    group_b = "Non_farmer_controls",
    out_stub = "older_vs_controls"
  )

  exposed_higher_vs_controls <- bind_rows(
    interp_suckling_vs_controls %>% mutate(contrast = "Suckling_vs_controls"),
    interp_older_vs_controls %>% mutate(contrast = "Older_vs_controls"),
    interp_farmers_vs_controls %>% mutate(contrast = "Pig_farmers_vs_controls")
  ) %>%
    filter(direction %in% c("higher_in_Suckling_piglets", "higher_in_Older_pigs", paste0("higher_in_", farmer_level))) %>%
    arrange(contrast, padj)

  print(exposed_higher_vs_controls, n = Inf)
  write_csv(exposed_higher_vs_controls, "../results/interpretation_exposed_groups_higher_vs_controls.csv")
} else {
  message("Run the previous chunk first (or ensure category2 is present) to compute stage-specific contrasts.")
}
```

#### shared AMR signature scatterplot (pigs vs controls x farmers vs controls)

```{r}
# One point per drug class:
# x = log2 median ratio (Pigs vs Non_farmer_controls)
# y = log2 median ratio (Pig_farmers vs Non_farmer_controls)

if (!exists("interp_pigs_vs_controls")) {
  interp_pigs_vs_controls <- compare_resfinder_groups(
    cpm_resfinder_long,
    group_var = "category",
    group_a = "Pigs",
    group_b = "Non_farmer_controls",
    out_stub = "pigs_vs_controls"
  )
}

if (!exists("interp_farmers_vs_controls")) {
  interp_farmers_vs_controls <- compare_resfinder_groups(
    cpm_resfinder_long,
    group_var = "category",
    group_a = farmer_level,
    group_b = "Non_farmer_controls",
    out_stub = "pig_farmers_vs_controls"
  )
}

drugclass_fc_scatter <- interp_pigs_vs_controls %>%
  select(
    Drug_Class_Resfinder,
    l2mr_pigs_vs_ctrl = log2_median_ratio,
    padj_pigs_vs_ctrl = padj
  ) %>%
  inner_join(
    interp_farmers_vs_controls %>%
      select(
        Drug_Class_Resfinder,
        l2mr_farmers_vs_ctrl = log2_median_ratio,
        padj_farmers_vs_ctrl = padj
      ),
    by = "Drug_Class_Resfinder"
  ) %>%
  mutate(
    quadrant = case_when(
      l2mr_pigs_vs_ctrl > 0 & l2mr_farmers_vs_ctrl > 0 ~ "higher_in_both_exposed",
      l2mr_pigs_vs_ctrl > 0 & l2mr_farmers_vs_ctrl <= 0 ~ "higher_in_pigs_only",
      l2mr_pigs_vs_ctrl <= 0 & l2mr_farmers_vs_ctrl > 0 ~ "higher_in_farmers_only",
      TRUE ~ "higher_in_controls"
    ),
    mw_pigs = case_when(
      is.na(padj_pigs_vs_ctrl) ~ "not_tested",
      padj_pigs_vs_ctrl < 0.05 & l2mr_pigs_vs_ctrl > 0 ~ "sig_higher_in_all_pigs",
      padj_pigs_vs_ctrl < 0.05 & l2mr_pigs_vs_ctrl < 0 ~ "sig_higher_in_controls",
      TRUE ~ "ns"
    ),
    mw_farmers = case_when(
      is.na(padj_farmers_vs_ctrl) ~ "not_tested",
      padj_farmers_vs_ctrl < 0.05 & l2mr_farmers_vs_ctrl > 0 ~ "sig_higher_in_pig_farmers",
      padj_farmers_vs_ctrl < 0.05 & l2mr_farmers_vs_ctrl < 0 ~ "sig_higher_in_controls",
      TRUE ~ "ns"
    ),
    concordance = case_when(
      mw_pigs == "not_tested" | mw_farmers == "not_tested" ~ "not_tested",
      mw_pigs == "sig_higher_in_all_pigs" & mw_farmers == "sig_higher_in_pig_farmers" ~ "sig_both_exposed",
      mw_pigs == "sig_higher_in_all_pigs" & mw_farmers != "sig_higher_in_pig_farmers" ~ "sig_pigs_only",
      mw_pigs != "sig_higher_in_all_pigs" & mw_farmers == "sig_higher_in_pig_farmers" ~ "sig_farmers_only",
      mw_pigs == "sig_higher_in_controls" | mw_farmers == "sig_higher_in_controls" ~ "sig_controls",
      TRUE ~ "ns"
    ),
    max_sig = pmax(-log10(padj_pigs_vs_ctrl + 1e-300), -log10(padj_farmers_vs_ctrl + 1e-300), na.rm = TRUE)
  )

label_df <- drugclass_fc_scatter %>%
  arrange(desc(max_sig)) %>%
  slice_head(n = 18)

p_drugclass_fc_scatter <- ggplot(
  drugclass_fc_scatter,
  aes(x = l2mr_pigs_vs_ctrl, y = l2mr_farmers_vs_ctrl, color = quadrant, shape = concordance)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  geom_point(alpha = 0.9, size = 2.5) +
  ggrepel::geom_text_repel(
    data = label_df,
    aes(label = Drug_Class_Resfinder),
    family = "DeJaVu Sans",
    size = 5,
    max.overlaps = 40,
    box.padding = 0.25,
    point.padding = 0.2,
    segment.alpha = 0.5,
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = c(
      "higher_in_both_exposed" = "#1B9E77",
      "higher_in_pigs_only" = "#D95F02",
      "higher_in_farmers_only" = "#7570B3",
      "higher_in_controls" = "#666666"
    ),
    labels = c(
      "higher_in_both_exposed" = "Higher in both exposed groups",
      "higher_in_pigs_only" = "Higher in pigs only",
      "higher_in_farmers_only" = "Higher in farmers only",
      "higher_in_controls" = "Higher in controls"
    )
  ) +
  scale_shape_manual(
    values = c(
      "sig_both_exposed" = 16,
      "sig_pigs_only" = 17,
      "sig_farmers_only" = 15,
      "sig_controls" = 18,
      "ns" = 1,
      "not_tested" = 4
    ),
    labels = c(
      "sig_both_exposed" = "sig in both exposed groups",
      "sig_pigs_only" = "sig_higher_in_all_pigs only",
      "sig_farmers_only" = "sig_higher_in_pig_farmers only",
      "sig_controls" = "sig_higher_in_controls",
      "ns" = "ns",
      "not_tested" = "not_tested"
    )
  ) +
  labs(
    title = "Drug-class effect concordance across exposed groups",
    subtitle = "x: Pigs vs controls, y: Pig farmers vs controls (log2 median ratio)",
    x = "log2 median ratio (Pigs vs Controls)",
    y = "log2 median ratio (Farmers vs Controls)",
    color = "exposed groups",
    shape = "Mann-Whitney concordance"
  ) +
  theme_bw(base_family = "DejaVu Sans", base_size = FIG_BASE_SIZE) +
  theme(legend.position = "right")

p_drugclass_fc_scatter <- p_drugclass_fc_scatter +
  theme(
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.position = "bottom",
    legend.box = "horizontal"
  )+
  labs(title = NULL, subtitle = NULL)+
  guides(
    color = guide_legend(
      title.position = "top",
      nrow = 3,       # 3 color entries under "exposed groups"
      byrow = TRUE
    ),
    shape = guide_legend(
      title.position = "top",
      nrow = 4,       # 4 shape entries under "Mann-Whitney concordance"
      byrow = TRUE
    )
  )

print(p_drugclass_fc_scatter)

write_csv(
  drugclass_fc_scatter,
  "../results/interpretation_drugclass_log2ratio_scatter_pigs_farmers_vs_controls.csv"
)

ggsave(
  "../figures/interpretation_drugclass_log2ratio_scatter_pigs_farmers_vs_controls.svg",
  p_drugclass_fc_scatter,
  width = 7,
  height = 8,
  web_fonts = "DejaVu Sans"
)
```

#### drugs the same in pigs and pig farmers

```{r}
# Drug classes with similar abundance between pigs and pig farmers.
# "Same" is defined as:
# 1) no significant difference (BH-adjusted p >= 0.05)
# 2) small effect size (absolute log2 median ratio <= effect_tolerance)

if (!exists("interp_pigs_vs_farmers")) {
  interp_pigs_vs_farmers <- compare_resfinder_groups(
    cpm_resfinder_long,
    group_var = "category",
    group_a = "Pigs",
    group_b = farmer_level,
    out_stub = "pigs_vs_pig_farmers"
  )
}

effect_tolerance <- 0.5

same_pigs_farmers_tbl <- interp_pigs_vs_farmers %>%
  mutate(
    same_status = case_when(
      is.na(padj) ~ "not_tested",
      padj >= 0.05 & abs(log2_median_ratio) <= effect_tolerance ~ "same",
      TRUE ~ "different"
    )
  ) %>%
  arrange(same_status, desc(padj), abs(log2_median_ratio))

same_only_tbl <- same_pigs_farmers_tbl %>%
  filter(same_status == "same") %>%
  select(
    Drug_Class_Resfinder,
    n_a,
    n_b,
    median_a,
    median_b,
    log2_median_ratio,
    p_value,
    padj,
    same_status
  )

cat("Drug classes classified as same (Pigs vs Pig_farmers):", nrow(same_only_tbl), "\n")
print(same_only_tbl, n = Inf)

write_csv(
  same_only_tbl,
  "../results/interpretation_same_drugclasses_pigs_vs_pigfarmers.csv"
)

write_csv(
  same_pigs_farmers_tbl,
  "../results/interpretation_same_drugclasses_pigs_vs_pigfarmers_full.csv"
)

# Quick overview plot: classes near zero effect and non-significant are highlighted.
p_same_pigs_farmers <- ggplot(
  same_pigs_farmers_tbl,
  aes(x = log2_median_ratio, y = -log10(padj + 1e-300), color = same_status)
) +
  geom_vline(xintercept = c(-effect_tolerance, effect_tolerance), linetype = "dashed", color = "grey60", linewidth = 0.4) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey60", linewidth = 0.4) +
  geom_point(alpha = 0.85, size = 2.2) +
  ggrepel::geom_text_repel(
    aes(label = Drug_Class_Resfinder),
    size = 2.7,
    max.overlaps = Inf,
    box.padding = 0.2,
    point.padding = 0.15,
    segment.alpha = 0.5,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c("same" = "#1B9E77", "different" = "#D95F02", "not_tested" = "#999999")) +
  labs(
    title = "Pigs vs Pig farmers: similar vs different drug classes",
    subtitle = paste0("same = BH p >= 0.05 and |log2 median ratio| <= ", effect_tolerance),
    x = "log2 median ratio (Pigs / Pig_farmers)",
    y = "-log10(BH p-value)",
    color = "Classification"
  ) +
  theme_bw(base_family = "DejaVu Sans", base_size = FIG_BASE_SIZE) +
  theme(legend.position = "bottom")+
  theme_publication()

print(p_same_pigs_farmers)

ggsave(
  "../figures/interpretation_same_drugclasses_pigs_vs_pigfarmers.svg",
  p_same_pigs_farmers,
  width = 9,
  height = 6,
  web_fonts = "DejaVu Sans"
)
```

#### drugs different between humans (controls and pig farmers) vs. pigs

Interpretation: This final contrast highlights Drug_Class_Resfinder classes that separate the pig microbiome signal from the human-associated signal (controls + pig farmers). Classes with BH-significant p-values and non-trivial log2 median ratios indicate biologically relevant host-group differences, while non-significant classes suggest overlap between human and pig resistome composition.

```{r}

# Explicit contrast: Pigs vs Humans (Humans = Non_farmer_controls + Pig_farmers)
cpm_resfinder_human_vs_pig <- cpm_resfinder_long %>%
  mutate(
    host_group = case_when(
      category == "Pigs" ~ "Pigs",
      category %in% c("Non_farmer_controls", farmer_level) ~ "Humans",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(host_group))

interp_humans_vs_pigs <- compare_resfinder_groups(
  cpm_resfinder_human_vs_pig,
  group_var = "host_group",
  group_a = "Pigs",
  group_b = "Humans",
  out_stub = "pigs_vs_humans_combined"
)

# Split for concise interpretation tables
human_higher_tbl <- interp_humans_vs_pigs %>%
  filter(direction == "higher_in_Humans") %>%
  arrange(padj, desc(abs(log2_median_ratio)))

pig_higher_tbl <- interp_humans_vs_pigs %>%
  filter(direction == "higher_in_Pigs") %>%
  arrange(padj, desc(abs(log2_median_ratio)))

write_csv(
  human_higher_tbl,
  "../results/interpretation_humans_higher_vs_pigs_combined.csv"
)

write_csv(
  pig_higher_tbl,
  "../results/interpretation_pigs_higher_vs_humans_combined.csv"
)

cat("Human-higher classes (BH<0.05 direction call):", nrow(human_higher_tbl), "\n")
cat("Pig-higher classes (BH<0.05 direction call):", nrow(pig_higher_tbl), "\n")
print(head(interp_humans_vs_pigs, 20))

```

#### non-sig between pigs and farmers

```{r}
# Same base contrast as the pigs vs pig farmers section, but focused on non-differential classes.
if (!exists("interp_pigs_vs_farmers")) {
  interp_pigs_vs_farmers <- compare_resfinder_groups(
    cpm_resfinder_long,
    group_var = "category",
    group_a = "Pigs",
    group_b = farmer_level,
    out_stub = "pigs_vs_pig_farmers"
  )
}

effect_tolerance <- 0.5

non_diff_pigs_farmers_tbl <- interp_pigs_vs_farmers %>%
  mutate(
    non_diff_status = case_when(
      is.na(padj) ~ "not_tested",
      padj >= 0.05 ~ "non_differential",
      TRUE ~ "differential"
    ),
    near_zero_effect = abs(log2_median_ratio) <= effect_tolerance
  ) %>%
  arrange(non_diff_status, desc(padj), abs(log2_median_ratio))

non_diff_only_tbl <- non_diff_pigs_farmers_tbl %>%
  filter(non_diff_status == "non_differential") %>%
  select(
    Drug_Class_Resfinder,
    n_a,
    n_b,
    median_a,
    median_b,
    log2_median_ratio,
    p_value,
    padj,
    near_zero_effect,
    non_diff_status
  )

write_csv(
  non_diff_only_tbl,
  "../results/interpretation_non_differential_drugclasses_pigs_vs_pigfarmers.csv"
)

write_csv(
  non_diff_pigs_farmers_tbl,
  "../results/interpretation_non_differential_drugclasses_pigs_vs_pigfarmers_full.csv"
)

cat("Non-differential classes (BH p >= 0.05):", nrow(non_diff_only_tbl), "\n")

# Barplot style (same family as other interpretation plots), focused on non-differential classes.
non_diff_plot_tbl <- non_diff_only_tbl %>%
  mutate(effect_class = if_else(near_zero_effect, "near_zero_effect", "non_zero_effect")) %>%
  arrange(abs(log2_median_ratio), desc(padj)) %>%
  slice_head(n = 30)

p_non_diff_pigs_farmers <- ggplot(
  non_diff_plot_tbl,
  aes(
    x = reorder(Drug_Class_Resfinder, log2_median_ratio),
    y = log2_median_ratio,
    fill = effect_class
  )
) +
  geom_col(alpha = 0.9) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_hline(yintercept = c(-effect_tolerance, effect_tolerance), linetype = "dotted", color = "grey65", linewidth = 0.35) +
  scale_fill_manual(
    values = c(
      "near_zero_effect" = "#1B9E77",
      "non_zero_effect" = "#80CDC1"
    ),
    labels = c(
      "near_zero_effect" = paste0("|log2 ratio| <= ", effect_tolerance),
      "non_zero_effect" = paste0("|log2 ratio| > ", effect_tolerance)
    )
  ) +
  labs(
    title = "Pigs vs Pig farmers: non-differential drug classes",
    subtitle = "Top non-differential classes (BH p >= 0.05), ranked by smallest absolute effect",
    x = "Drug class",
    y = "log2 median ratio (Pigs / Pig_farmers)",
    fill = "Effect size class"
  ) +
  theme_bw(base_family = "DejaVu Sans", base_size = FIG_BASE_SIZE) +
  theme(legend.position = "bottom") +
  theme_publication()

print(p_non_diff_pigs_farmers)

ggsave(
  "../figures/interpretation_non_differential_drugclasses_pigs_vs_pigfarmers.svg",
  p_non_diff_pigs_farmers,
  width = 10,
  height = 7,
  web_fonts = "DejaVu Sans"
)
```

#### relative abundance bar plots new drug classes, cpm values

```{r}
# Build 100% stacked bar chart from the new Resfinder drug-class annotation.
# We first compute per-sample relative abundance, then average within category.
rel_resfinder_by_sample <- cpm_resfinder_long %>%
  filter(!is.na(category), category != "") %>%
  group_by(sample) %>%
  mutate(total_cpm_sample = sum(cpm_resfinder, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    rel_abundance = if_else(total_cpm_sample > 0, cpm_resfinder / total_cpm_sample, 0)
  )

rel_resfinder_by_category_raw <- rel_resfinder_by_sample %>%
  group_by(category, Drug_Class_Resfinder) %>%
  summarise(
    mean_rel_abundance = mean(rel_abundance, na.rm = TRUE),
    .groups = "drop"
  )

rel_resfinder_by_category <- rel_resfinder_by_category_raw %>%
  group_by(category) %>%
  mutate(
    mean_rel_abundance = mean_rel_abundance / sum(mean_rel_abundance, na.rm = TRUE)
  ) %>%
  ungroup()

rel_resfinder_plot <- rel_resfinder_by_category

drug_class_levels_rel <- rel_resfinder_by_category %>%
  group_by(Drug_Class_Resfinder) %>%
  summarise(total_mean_rel_abundance = sum(mean_rel_abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_mean_rel_abundance)) %>%
  pull(Drug_Class_Resfinder)

drug_class_palette_rel <- grDevices::hcl.colors(length(drug_class_levels_rel), palette = "Dynamic")
names(drug_class_palette_rel) <- drug_class_levels_rel

p_rel_abundance_new_drugclass <- ggplot(
  rel_resfinder_by_category %>% mutate(Drug_Class_Resfinder = factor(Drug_Class_Resfinder, levels = drug_class_levels_rel)),
  aes(x = category, y = mean_rel_abundance, fill = Drug_Class_Resfinder)
) +
  geom_col(width = 0.9, color = "white", linewidth = 0.2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Relative ARG abundance by new drug-class annotation",
    subtitle = "Category mean of per-sample relative abundances using manually annotated Drug_Class_Resfinder labels",
    x = "Study population",
    y = "Relative abundance",
    fill = "Drug class"
  ) +
  scale_fill_manual(values = drug_class_palette_rel, drop = FALSE) +
  theme_publication() +
  theme(legend.position = "right")

print(p_rel_abundance_new_drugclass)

write_csv(rel_resfinder_by_sample, "../results/relative_abundance_new_drugclass_by_sample.csv")
write_csv(rel_resfinder_plot, "../results/relative_abundance_new_drugclass_by_category.csv")

ggsave(
  "../figures/relative_abundance_new_drugclass_by_category.svg",
  p_rel_abundance_new_drugclass,
  width = 12,
  height = 7,
  web_fonts = "DejaVu Sans"
)
```

#### relative abundance bar plots with MLS combined

```{r}
# Combine MLS antibiotics (Macrolide, Lincosamide, Streptogramin) into single category
# and recalculate relative abundance with updated color scheme

rel_resfinder_mls_combined <- cpm_resfinder_long %>%
  filter(!is.na(category), category != "") %>%
  mutate(
    Drug_Class_Resfinder_Combined = case_when(
      str_to_lower(str_trim(Drug_Class_Resfinder)) %in% c(
        "macrolide", "macrolides",
        "lincosamide", "lincosamides",
        "streptogramin", "streptogramins"
      ) ~ "MLS",
      TRUE ~ Drug_Class_Resfinder
    )
  ) %>%
  group_by(sample, Drug_Class_Resfinder_Combined) %>%
  summarise(
    cpm_combined = sum(cpm_resfinder, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(meta %>% rownames_to_column("sample") %>% select(sample, category), by = "sample")

# Compute per-sample relative abundance
rel_resfinder_mls_by_sample <- rel_resfinder_mls_combined %>%
  group_by(sample) %>%
  mutate(total_cpm_sample = sum(cpm_combined, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    rel_abundance = if_else(total_cpm_sample > 0, cpm_combined / total_cpm_sample, 0)
  )

# Average per category
rel_resfinder_mls_by_category <- rel_resfinder_mls_by_sample %>%
  group_by(category, Drug_Class_Resfinder_Combined) %>%
  summarise(
    mean_rel_abundance = mean(rel_abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(category) %>%
  mutate(
    mean_rel_abundance = mean_rel_abundance / sum(mean_rel_abundance, na.rm = TRUE)
  ) %>%
  ungroup()

mls_levels <- rel_resfinder_mls_by_category %>%
  group_by(Drug_Class_Resfinder_Combined) %>%
  summarise(total_mean_rel_abundance = sum(mean_rel_abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_mean_rel_abundance)) %>%
  pull(Drug_Class_Resfinder_Combined)

mls_colors <- grDevices::hcl.colors(length(mls_levels), palette = "Dynamic")
names(mls_colors) <- mls_levels

# Keep MLS visually emphasized.
if ("MLS" %in% names(mls_colors)) {
  mls_colors["MLS"] <- "#D62728"
}

p_rel_abundance_mls_combined <- ggplot(
  rel_resfinder_mls_by_category %>%
    mutate(Drug_Class_Resfinder_Combined = factor(Drug_Class_Resfinder_Combined, levels = mls_levels)),
  aes(x = category, y = mean_rel_abundance, fill = Drug_Class_Resfinder_Combined)
) +
  geom_col(width = 0.85, color = "white", linewidth = 0.25) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(
    values = mls_colors,
    name = "Drug class",
    guide = guide_legend(ncol = 2)
  ) +
  labs(
    title = "Relative ARG abundance by drug class (MLS combined)",
    subtitle = "MLS antibiotics (Macrolide, Lincosamide, Streptogramin) aggregated as single category",
    x = "Study population",
    y = "Relative abundance"
  ) +
  theme_publication() +
  theme(legend.position = "right")

print(p_rel_abundance_mls_combined)

write_csv(rel_resfinder_mls_by_sample, "../results/relative_abundance_mls_combined_by_sample.csv")
write_csv(rel_resfinder_mls_by_category, "../results/relative_abundance_mls_combined_by_category.csv")

ggsave(
  "../figures/relative_abundance_mls_combined_by_category.svg",
  p_rel_abundance_mls_combined,
  width = 12,
  height = 7,
  web_fonts = "DejaVu Sans"
)

cat("✓ MLS-combined relative abundance plot generated and saved\n")
```

#### 3x3 summary barplot of selected drug classes

```{r}
# Publication-style 3x3 panel of mean CPM abundance by category for selected classes.
target_class_map <- c(
  "sptrepgoramin" = "Streptogramin",
  "lincosamides" = "Lincosamides",
  "macrolides" = "Macrolides",
  "aiminogylcosides" = "Aminoglycosides",
  "sulfonamides" = "Sulfonamide",
  "beta-lactam" = "Beta-lactam",
  "mupirocin" = "Mupirocin",
  "glycopeptides" = "Glycopeptides",
  "tetracylcines" = "Tetracyclines"
)

target_classes <- unname(target_class_map)

sample_tbl <- meta %>%
  rownames_to_column("sample") %>%
  select(sample, category)

# Complete sample x class grid so missing observations are treated as true zeros.
summary_3x3_long <- tidyr::expand_grid(
  sample = sample_tbl$sample,
  Drug_Class_Resfinder = target_classes
) %>%
  left_join(cpm_resfinder_long %>% select(sample, Drug_Class_Resfinder, cpm_resfinder),
            by = c("sample", "Drug_Class_Resfinder")) %>%
  left_join(sample_tbl, by = "sample") %>%
  mutate(cpm_resfinder = replace_na(cpm_resfinder, 0))

summary_3x3_stats <- summary_3x3_long %>%
  group_by(Drug_Class_Resfinder, category) %>%
  summarise(
    mean_cpm = mean(cpm_resfinder, na.rm = TRUE),
    se_cpm = sd(cpm_resfinder, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

panel_labels <- c(
  "Streptogramin" = "Streptogramin",
  "Lincosamides" = "Lincosamides",
  "Macrolides" = "Macrolides",
  "Aminoglycosides" = "Aminoglycosides",
  "Sulfonamide" = "Sulfonamides",
  "Beta-lactam" = "Beta-lactam",
  "Mupirocin" = "Mupirocin",
  "Glycopeptides" = "Glycopeptides",
  "Tetracyclines" = "Tetracyclines"
)

summary_3x3_stats <- summary_3x3_stats %>%
  mutate(
    Drug_Class_Resfinder = factor(Drug_Class_Resfinder, levels = target_classes, labels = panel_labels[target_classes]),
    category = factor(category, levels = c("Non_farmer_controls", "Pig_farmers", "Pigs"))
  )

summary_3x3_plot <- summary_3x3_long %>%
  mutate(
    Drug_Class_Resfinder = factor(Drug_Class_Resfinder, levels = target_classes, labels = panel_labels[target_classes]),
    category = factor(category, levels = c("Non_farmer_controls", "Pig_farmers", "Pigs"))
  )

p_drugclass_3x3 <- ggplot(summary_3x3_plot, aes(x = category, y = cpm_resfinder, fill = category)) +
  geom_boxplot(width = 0.7, alpha = 0.85, outlier.shape = NA, linewidth = 0.3) +
  geom_jitter(width = 0.18, size = 1.0, alpha = 0.35, color = "black") +
  ggpubr::stat_compare_means(
    method = "wilcox.test",
    comparisons = list(
      c("Non_farmer_controls", "Pig_farmers"),
      c("Non_farmer_controls", "Pigs"),
      c("Pig_farmers", "Pigs")
    ),
    p.adjust.method = "BH",
    label = "p.format",
    family = "DejaVu Sans",
    hide.ns = TRUE,
    step.increase = 0.08,
    size = 4
  ) +
  scale_x_discrete(labels = color_codes$label_mapping)+
  facet_wrap(~ Drug_Class_Resfinder, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = color_codes$sample_types) +
  scale_y_continuous(labels = scales::label_number(accuracy = 1)) +
  labs(
    title = "ARG abundance distribution across selected drug classes",
    subtitle = "Per-sample CPM by category (pairwise Wilcoxon rank-sum tests, BH-adjusted)",
    x = NULL,
    y = "Abundance (CPM)",
    fill = "Category"
  ) +
  theme_publication() +
  theme(
    #text = element_text(family =  "DejaVu Sans"),
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
    
  )

print(p_drugclass_3x3)

write_csv(summary_3x3_stats, "../results/drugclass_3x3_summary_stats.csv")
write_csv(summary_3x3_plot, "../results/drugclass_3x3_sample_values.csv")

library(svglite)
ggsave(
  "../figures/drugclass_3x3_summary_barplot.svg",
  p_drugclass_3x3,
  width = 13,
  height = 14,
  device = svglite::svglite,
  system_fonts = list(sans = "DejaVu Sans")
)
```

## Hypothesis Check: High Richness but Low Evenness in Pigs and Pig Farmers

```{r}
# Test whether pigs/farmers show higher richness but stronger dominance and lower evenness.

# Build ARO-term x sample matrix from currently active CPM matrix (global CPM<1 filter already applied).
gene_to_aro_hyp <- feats %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  transmute(
    Prevalence_Sequence_ID,
    ARO_Term = if_else(is.na(ARO_Term) | ARO_Term == "", Prevalence_Sequence_ID, as.character(ARO_Term))
  )

cpm_aro_hyp_long <- cpm %>%
  as.data.frame() %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  pivot_longer(cols = -Prevalence_Sequence_ID, names_to = "sample", values_to = "cpm") %>%
  left_join(gene_to_aro_hyp, by = "Prevalence_Sequence_ID") %>%
  group_by(sample, ARO_Term) %>%
  summarise(cpm_aro = sum(cpm, na.rm = TRUE), .groups = "drop")

cpm_aro_hyp_mat <- cpm_aro_hyp_long %>%
  pivot_wider(names_from = sample, values_from = cpm_aro, values_fill = 0) %>%
  as.data.frame()

rownames(cpm_aro_hyp_mat) <- cpm_aro_hyp_mat$ARO_Term
cpm_aro_hyp_mat$ARO_Term <- NULL
cpm_aro_hyp_mat <- as.matrix(cpm_aro_hyp_mat)

samples_keep_hyp <- intersect(colnames(cpm_aro_hyp_mat), rownames(meta))
samples_keep_hyp <- intersect(samples_keep_hyp, names(sample_depths))
cpm_aro_hyp_mat <- cpm_aro_hyp_mat[, samples_keep_hyp, drop = FALSE]

# Per-sample composition metrics on ARO-term level
sample_metrics_hyp <- purrr::map_dfr(colnames(cpm_aro_hyp_mat), function(s) {
  x <- cpm_aro_hyp_mat[, s]
  x <- x[is.finite(x)]
  s_total <- sum(x)
  p <- if (s_total > 0) x / s_total else rep(0, length(x))
  p_pos <- p[p > 0]
  richness <- sum(x > 0)
  shannon <- if (length(p_pos) > 0) -sum(p_pos * log(p_pos)) else NA_real_
  simpson <- if (length(p_pos) > 0) 1 - sum(p_pos^2) else NA_real_
  pielou <- if (!is.na(shannon) && richness > 1) shannon / log(richness) else NA_real_
  dominance <- if (length(p_pos) > 0) max(p_pos) else NA_real_
  top10_share <- if (length(p_pos) > 0) sum(sort(p_pos, decreasing = TRUE)[1:min(10, length(p_pos))]) else NA_real_
  rare_share <- if (s_total > 0) sum(x[x > 0 & x < 10]) / s_total else NA_real_

  tibble(
    sample = s,
    aro_richness = richness,
    shannon = shannon,
    simpson = simpson,
    pielou_J = pielou,
    dominance_max = dominance,
    top10_share = top10_share,
    rare_share_cpm_lt10 = rare_share
  )
}) %>%
  left_join(meta %>% rownames_to_column("sample"), by = "sample") %>%
  mutate(
    category = as.factor(category),
    depth = as.numeric(sample_depths[sample])
  )

# Depth-corrected models for key hypothesis metrics
metrics_to_test <- c("aro_richness", "shannon", "simpson", "pielou_J", "dominance_max", "top10_share", "rare_share_cpm_lt10")

hypothesis_lm_stats <- purrr::map_dfr(metrics_to_test, function(m) {
  fit <- lm(as.formula(paste(m, "~ category + depth")), data = sample_metrics_hyp)
  a <- anova(fit)
  tibble(
    metric = m,
    p_category = a["category", "Pr(>F)"],
    p_depth = a["depth", "Pr(>F)"]
  )
})

print(hypothesis_lm_stats)

# Rank-abundance (Whittaker) by category using category medians
rank_curve_tbl <- sample_metrics_hyp %>%
  select(sample, category) %>%
  distinct() %>%
  inner_join(cpm_aro_hyp_long, by = "sample") %>%
  group_by(category, ARO_Term) %>%
  summarise(median_cpm = median(cpm_aro, na.rm = TRUE), .groups = "drop") %>%
  group_by(category) %>%
  mutate(rel = median_cpm / sum(median_cpm, na.rm = TRUE)) %>%
  filter(rel > 0) %>%
  arrange(category, desc(rel)) %>%
  group_by(category) %>%
  mutate(rank = row_number()) %>%
  ungroup()

p_rank_abundance_hyp <- ggplot(rank_curve_tbl, aes(x = rank, y = rel, color = category)) +
  geom_line(linewidth = 1) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = sample_type_colors) +
  labs(
    title = "ARO-term rank-abundance curves by category",
    subtitle = "Steeper curves indicate stronger dominance and lower evenness",
    x = "Rank (log scale)",
    y = "Relative abundance (log scale)",
    color = "Category"
  ) +
  theme_publication()

print(p_rank_abundance_hyp)

# Compact panel for interpretation metrics
plot_metrics <- c("dominance_max", "top10_share", "rare_share_cpm_lt10")

plot_metrics_tbl <- sample_metrics_hyp %>%
  select(sample, category, all_of(plot_metrics)) %>%
  pivot_longer(cols = all_of(plot_metrics), names_to = "metric", values_to = "value")

hypothesis_kw_stats <- plot_metrics_tbl %>%
  group_by(metric) %>%
  rstatix::kruskal_test(value ~ category)

hypothesis_metric_limits <- plot_metrics_tbl %>%
  group_by(metric) %>%
  summarise(
    y_max = max(value, na.rm = TRUE),
    y_range = diff(range(value, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    y_step = if_else(
      is.finite(y_range) & y_range > 0,
      y_range * 0.08,
      pmax(abs(y_max) * 0.1, 0.1)
    )
  )

hypothesis_pairwise_stats <- plot_metrics_tbl %>%
  group_by(metric) %>%
  pairwise_wilcox_test(value ~ category, p.adjust.method = "BH") %>%
  add_significance() %>%
  left_join(hypothesis_metric_limits, by = "metric") %>%
  group_by(metric) %>%
  arrange(p.adj, .by_group = TRUE) %>%
  mutate(y.position = y_max + row_number() * y_step) %>%
  ungroup() %>%
  mutate(
    p.adj_label = if_else(
      is.na(p.adj),
      NA_character_,
      if_else(p.adj < 0.001, "<0.001", formatC(p.adj, format = "f", digits = 3))
    )
  )

print(hypothesis_kw_stats)
print(hypothesis_pairwise_stats)

p_hyp_metrics <- ggplot(plot_metrics_tbl, aes(x = category, y = value, fill = category)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.75) +
  geom_jitter(width = 0.18, alpha = 0.5, size = 1.2) +
  facet_wrap(~ metric, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = sample_type_colors) +
  stat_pvalue_manual(
    hypothesis_pairwise_stats,
    label = "p.adj_label",
    hide.ns = TRUE,
    tip.length = 0.01,
    size = 5,
    family = "DejaVu Sans"
  ) +
  labs(
    title = "Hypothesis diagnostics on ARO-term composition",
    subtitle = "Expected pattern: higher richness with lower evenness and higher dominance; per-panel Kruskal-Wallis and BH-adjusted pairwise Wilcoxon tests",
    x = NULL,
    y = "Value"
  ) +
  theme_publication() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 0, size = 12))+
  scale_x_discrete(labels = color_codes$label_mapping)

print(p_hyp_metrics)

write_csv(sample_metrics_hyp, "../results/hypothesis_aro_term_sample_metrics.csv")
write_csv(hypothesis_lm_stats, "../results/hypothesis_aro_term_depth_corrected_lm_pvalues.csv")
write_csv(hypothesis_kw_stats, "../results/hypothesis_aro_term_metrics_kruskal.csv")
write_csv(hypothesis_pairwise_stats, "../results/hypothesis_aro_term_metrics_pairwise_wilcox.csv")
write_csv(rank_curve_tbl, "../results/hypothesis_aro_term_rank_abundance_by_category.csv")

ggsave("../figures/hypothesis_aro_term_rank_abundance.svg", p_rank_abundance_hyp, width = 8, height = 6, web_fonts = "DejaVu Sans")
ggsave("../figures/hypothesis_aro_term_metrics_panel.svg", p_hyp_metrics, width = 12, height = 8, web_fonts = "DejaVu Sans")
```

### Rare-feature fraction by count (ARO-term level)

```{r}
# Rare-feature fraction by count:
#   #(0 < CPM < 10) / #(CPM > 0)

rare_fraction_count_df <- purrr::map_dfr(colnames(cpm_aro_hyp_mat), function(s) {
  x <- cpm_aro_hyp_mat[, s]
  x <- x[is.finite(x)]
  n_detected <- sum(x > 0)
  n_rare <- sum(x > 0 & x < 10)

  tibble(
    sample = s,
    n_detected_terms = n_detected,
    n_rare_terms_cpm_0_10 = n_rare,
    rare_feature_fraction_count = if_else(n_detected > 0, n_rare / n_detected, NA_real_)
  )
}) %>%
  left_join(meta %>% rownames_to_column("sample"), by = "sample") %>%
  mutate(category = as.factor(category))

pairwise_rare_fraction_count <- rare_fraction_count_df %>%
  filter(!is.na(rare_feature_fraction_count)) %>%
  pairwise_wilcox_test(rare_feature_fraction_count ~ category, p.adjust.method = "BH") %>%
  add_significance() %>%
  filter(p.adj.signif != "ns") %>%
  mutate(
    y.position = seq(
      from = max(rare_fraction_count_df$rare_feature_fraction_count, na.rm = TRUE) * 1.05,
      to = max(rare_fraction_count_df$rare_feature_fraction_count, na.rm = TRUE) * 1.20,
      length.out = dplyr::n()
    )
  )

p_rare_fraction_count <- ggplot(
  rare_fraction_count_df,
  aes(x = category, y = rare_feature_fraction_count, fill = category)
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.75) +
  geom_jitter(width = 0.18, alpha = 0.5, size = 1.2) +
  scale_fill_manual(values = sample_type_colors) +
  labs(
    title = "Rare-feature fraction by count",
    subtitle = "#(0 < CPM < 10) / #(CPM > 0) on ARO-term level",
    x = NULL,
    y = "Rare-feature fraction"
  ) +
  theme_publication() +
  theme(legend.position = "none") +
  scale_x_discrete(labels = color_codes$label_mapping)

if (nrow(pairwise_rare_fraction_count) > 0) {
  p_rare_fraction_count <- p_rare_fraction_count +
    stat_pvalue_manual(
      pairwise_rare_fraction_count,
      label = "p.adj",
      tip.length = 0.01,
      size = 4,
      family = "DejaVu Sans"
    )
}

print(p_rare_fraction_count)

write_csv(rare_fraction_count_df, "../results/hypothesis_aro_term_rare_feature_fraction_count_values.csv")
write_csv(pairwise_rare_fraction_count, "../results/hypothesis_aro_term_rare_feature_fraction_count_pairwise_wilcox.csv")

ggsave(
  "../figures/hypothesis_aro_term_rare_feature_fraction_count.svg",
  p_rare_fraction_count,
  width = 8,
  height = 6,
  web_fonts = "DejaVu Sans"
)
```

### Gene-level rare candidates in pigs and pig farmers

```{r}
# Identify genes that are frequently rare (0 < CPM < 10) in both exposed groups.
# Controls are not used as a filter criterion in this relaxed version.

farmer_level_local <- if ("Pig_farmers" %in% unique(meta$category)) {
  "Pig_farmers"
} else if ("Pig_farmer" %in% unique(meta$category)) {
  "Pig_farmer"
} else {
  stop("No pig farmer level found in meta$category.")
}

target_groups <- c(farmer_level_local, "Pigs")

gene_cpm_hyp_long <- cpm %>%
  as.data.frame() %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  pivot_longer(cols = -Prevalence_Sequence_ID, names_to = "sample", values_to = "cpm_gene") %>%
  left_join(
    meta %>% rownames_to_column("sample") %>% select(sample, category),
    by = "sample"
  ) %>%
  mutate(category = trimws(as.character(category))) %>%
  filter(category %in% target_groups)

gene_group_stats_long <- gene_cpm_hyp_long %>%
  group_by(Prevalence_Sequence_ID, category) %>%
  summarise(
    n_samples = n(),
    prev_nonzero = mean(cpm_gene > 0, na.rm = TRUE),
    prev_rare_0_10 = mean(cpm_gene > 0 & cpm_gene < 10, na.rm = TRUE),
    median_cpm = median(cpm_gene, na.rm = TRUE),
    mean_cpm = mean(cpm_gene, na.rm = TRUE),
    .groups = "drop"
  )

gene_group_stats <- gene_group_stats_long %>%
  mutate(category = as.character(category)) %>%
  pivot_wider(
    names_from = category,
    values_from = c(n_samples, prev_nonzero, prev_rare_0_10, median_cpm, mean_cpm),
    names_glue = "{.value}_{category}"
  )

# Tunable threshold (set lower if you want a broader candidate list)
min_rare_prev <- 0.10

pig_rare_col <- "prev_rare_0_10_Pigs"
farmer_rare_col <- paste0("prev_rare_0_10_", farmer_level_local)

required_cols <- c(pig_rare_col, farmer_rare_col)
missing_cols <- setdiff(required_cols, colnames(gene_group_stats))
if (length(missing_cols) > 0) {
  stop("Missing required columns in gene_group_stats: ", paste(missing_cols, collapse = ", "))
}

rare_gene_candidates_exposed_only <- gene_group_stats %>%
  filter(
    .data[[pig_rare_col]] >= min_rare_prev,
    .data[[farmer_rare_col]] >= min_rare_prev
  ) %>%
  mutate(
    mean_rare_prev_exposed = (.data[[pig_rare_col]] + .data[[farmer_rare_col]]) / 2
  )

rare_gene_ranked_exposed_only <- gene_group_stats %>%
  mutate(
    mean_rare_prev_exposed = (.data[[pig_rare_col]] + .data[[farmer_rare_col]]) / 2,
    min_rare_prev_exposed = pmin(.data[[pig_rare_col]], .data[[farmer_rare_col]])
  ) %>%
  left_join(
    feats %>%
      rownames_to_column("Prevalence_Sequence_ID") %>%
      select(Prevalence_Sequence_ID, ARO_Term = any_of("ARO_Term")),
    by = "Prevalence_Sequence_ID"
  ) %>%
  arrange(desc(min_rare_prev_exposed), desc(mean_rare_prev_exposed), desc(.data[[pig_rare_col]]), desc(.data[[farmer_rare_col]]))

rare_gene_candidates_exposed_only <- rare_gene_ranked_exposed_only %>%
  filter(
    .data[[pig_rare_col]] >= min_rare_prev,
    .data[[farmer_rare_col]] >= min_rare_prev
  )

cat("Rare-gene candidates (rare in pigs and pig farmers):", nrow(rare_gene_candidates_exposed_only), "\n")

print(
  rare_gene_ranked_exposed_only %>%
    select(Prevalence_Sequence_ID, ARO_Term, all_of(c(pig_rare_col, farmer_rare_col)), mean_rare_prev_exposed, min_rare_prev_exposed) %>%
    slice_head(n = 30)
)

write_csv(gene_group_stats, "../results/hypothesis_gene_level_group_prevalence_summary.csv")
write_csv(rare_gene_ranked_exposed_only, "../results/hypothesis_rare_gene_ranked_exposed_only_with_aro.csv")
write_csv(rare_gene_candidates_exposed_only, "../results/hypothesis_rare_gene_candidates_exposed_only_with_aro.csv")
```

### Visualize ARO terms from rare exposed genes

```{r}
# Build selected ARO-term set from rare genes and visualize across groups.
rare_exposed_aro_terms <- rare_gene_candidates_exposed_only %>%
  mutate(
    ARO_Term = if_else(is.na(ARO_Term) | ARO_Term == "", Prevalence_Sequence_ID, as.character(ARO_Term))
  ) %>%
  distinct(ARO_Term) %>%
  pull(ARO_Term)

cat("Unique ARO terms from rare exposed genes:", length(rare_exposed_aro_terms), "\n")

cpm_aro_rare_terms_long <- cpm_aro_hyp_long %>%
  filter(ARO_Term %in% rare_exposed_aro_terms) %>%
  left_join(meta %>% rownames_to_column("sample") %>% select(sample, category), by = "sample") %>%
  mutate(
    category = as.character(category),
    category = if_else(category == farmer_level_local, "Pig_farmers", category)
  ) %>%
  filter(category %in% c("Non_farmer_controls", "Pig_farmers", "Pigs"))

if (nrow(cpm_aro_rare_terms_long) == 0) {
  message("No rows available for selected rare exposed ARO terms.")
} else {
  cpm_aro_rare_terms_long <- cpm_aro_rare_terms_long %>%
    mutate(category = factor(category, levels = c("Non_farmer_controls", "Pig_farmers", "Pigs")))

  # 1) Multi-page boxplots for all selected ARO terms
  terms_sorted <- cpm_aro_rare_terms_long %>%
    group_by(ARO_Term) %>%
    summarise(overall_median = median(cpm_aro, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(overall_median)) %>%
    pull(ARO_Term)

  cpm_aro_rare_terms_long <- cpm_aro_rare_terms_long %>%
    mutate(ARO_Term = factor(ARO_Term, levels = terms_sorted))

  terms_per_page <- 12
  term_pages <- split(terms_sorted, ceiling(seq_along(terms_sorted) / terms_per_page))

  boxplot_pdf <- "../figures/rare_exposed_aro_terms_boxplots.pdf"
  if (capabilities("cairo")) {
    cairo_pdf(boxplot_pdf, width = 12, height = 10, family = "sans")
  } else {
    pdf(boxplot_pdf, width = 12, height = 10, family = "sans")
  }
  for (i in seq_along(term_pages)) {
    page_terms <- term_pages[[i]]
    p_page <- cpm_aro_rare_terms_long %>%
      filter(ARO_Term %in% page_terms) %>%
      ggplot(aes(x = category, y = cpm_aro, fill = category)) +
      geom_boxplot(outlier.shape = NA, alpha = 0.8) +
      geom_jitter(width = 0.18, alpha = 0.4, size = 0.9, color = "black") +
      facet_wrap(~ ARO_Term, ncol = 3, scales = "free_y") +
      scale_fill_manual(values = sample_type_colors) +
      labs(
        title = paste0("Rare exposed ARO terms by category (page ", i, " / ", length(term_pages), ")"),
        x = NULL,
        y = "CPM"
      ) +
      theme_publication() +
      theme(legend.position = "none") +
      scale_x_discrete(labels = color_codes$label_mapping)
    print(p_page)
  }
  dev.off()

  page_index_tbl <- tibble(
    ARO_Term = unlist(term_pages),
    page = rep(seq_along(term_pages), times = lengths(term_pages))
  )

  write_csv(page_index_tbl, "../results/rare_exposed_aro_terms_boxplot_page_index.csv")

  # 2) Presence overlap across groups (Venn-style)
  presence_tbl <- cpm_aro_rare_terms_long %>%
    group_by(ARO_Term, category) %>%
    summarise(present = any(cpm_aro > 0, na.rm = TRUE), .groups = "drop") %>%
    tidyr::complete(
      ARO_Term,
      category = c("Non_farmer_controls", "Pig_farmers", "Pigs"),
      fill = list(present = FALSE)
    ) %>%
    pivot_wider(names_from = category, values_from = present)

  overlap_membership_tbl <- presence_tbl %>%
    transmute(
      ARO_Term,
      overlap_pattern = case_when(
        Non_farmer_controls & Pig_farmers & Pigs ~ "All three",
        Non_farmer_controls & Pig_farmers & !Pigs ~ "Controls + Farmers",
        Non_farmer_controls & !Pig_farmers & Pigs ~ "Controls + Pigs",
        !Non_farmer_controls & Pig_farmers & Pigs ~ "Farmers + Pigs",
        Non_farmer_controls & !Pig_farmers & !Pigs ~ "Controls only",
        !Non_farmer_controls & Pig_farmers & !Pigs ~ "Farmers only",
        !Non_farmer_controls & !Pig_farmers & Pigs ~ "Pigs only",
        TRUE ~ "Absent in all"
      )
    )

  overlap_tbl <- overlap_membership_tbl %>%
    dplyr::count(overlap_pattern, name = "n_aro_terms") %>%
    arrange(desc(n_aro_terms))

  p_overlap <- ggplot(overlap_tbl, aes(x = reorder(overlap_pattern, n_aro_terms), y = n_aro_terms)) +
    geom_col(fill = "#4C78A8", alpha = 0.9) +
    coord_flip() +
    labs(
      title = "Overlap of rare exposed ARO terms across groups",
      subtitle = "Venn-style intersection counts",
      x = NULL,
      y = "Number of ARO terms"
    ) +
    theme_publication()

  ggsave("../figures/rare_exposed_aro_terms_overlap_barplot.svg", p_overlap, width = 8, height = 5, web_fonts = "DejaVu Sans")

  if (requireNamespace("ggvenn", quietly = TRUE)) {
    venn_list <- list(
      Non_farmer_controls = presence_tbl %>% filter(Non_farmer_controls) %>% pull(ARO_Term),
      Pig_farmers = presence_tbl %>% filter(Pig_farmers) %>% pull(ARO_Term),
      Pigs = presence_tbl %>% filter(Pigs) %>% pull(ARO_Term)
    )

    p_venn <- ggvenn::ggvenn(
      venn_list,
      fill_alpha = 0.4,
      stroke_size = 0.6,
      text_size = 4,
      set_name_size = 4.5
    ) +
      labs(title = "Venn diagram of rare exposed ARO terms")

    ggsave("../figures/rare_exposed_aro_terms_venn.svg", p_venn, width = 7, height = 6, web_fonts = "DejaVu Sans")
  }

  # 3) Mean CPM summary for selected ARO terms
  rare_aro_summary <- cpm_aro_rare_terms_long %>%
    group_by(ARO_Term, category) %>%
    summarise(
      mean_cpm = mean(cpm_aro, na.rm = TRUE),
      median_cpm = median(cpm_aro, na.rm = TRUE),
      prevalence_nonzero = mean(cpm_aro > 0, na.rm = TRUE),
      .groups = "drop"
    )

  rare_aro_dot_tbl <- rare_aro_summary %>%
    group_by(ARO_Term) %>%
    mutate(max_mean = max(mean_cpm, na.rm = TRUE)) %>%
    ungroup() %>%
    mutate(ARO_Term = forcats::fct_reorder(ARO_Term, max_mean, .desc = TRUE))

  p_rare_aro_dot <- ggplot(
    rare_aro_dot_tbl,
    aes(x = category, y = ARO_Term, size = prevalence_nonzero, color = mean_cpm)
  ) +
    geom_point(alpha = 0.9) +
    scale_color_viridis_c(option = "C", trans = "log10") +
    scale_size_continuous(range = c(1.2, 6)) +
    labs(
      title = "Mean CPM and prevalence of rare exposed ARO terms",
      subtitle = "Color = mean CPM (log scale), size = prevalence",
      x = NULL,
      y = "ARO term",
      color = "Mean CPM",
      size = "Prevalence"
    ) +
    theme_publication() +
    scale_x_discrete(labels = color_codes$label_mapping)

  ggsave("../figures/rare_exposed_aro_terms_mean_cpm_prevalence_dotplot.svg", p_rare_aro_dot, width = 10, height = 16, web_fonts = "DejaVu Sans")

  write_csv(overlap_tbl, "../results/rare_exposed_aro_terms_overlap_counts.csv")
  write_csv(presence_tbl, "../results/rare_exposed_aro_terms_presence_by_group.csv")
  write_csv(overlap_membership_tbl, "../results/rare_exposed_aro_terms_overlap_membership.csv")
  write_csv(rare_aro_summary, "../results/rare_exposed_aro_terms_mean_cpm_summary.csv")
  
  # Build long annotation table with overall and per-category statistics for Excel export
  rare_aro_category_wide <- rare_aro_summary %>%
    pivot_wider(
      names_from = category,
      values_from = c(mean_cpm, median_cpm, prevalence_nonzero),
      names_glue = "{.value}_{category}"
    )
  
  rare_aro_annotation_long <- cpm_aro_rare_terms_long %>%
    group_by(ARO_Term) %>%
    summarise(
      mean_cpm_overall = mean(cpm_aro, na.rm = TRUE),
      median_cpm_overall = median(cpm_aro, na.rm = TRUE),
      prevalence_overall = mean(cpm_aro > 0, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(rare_aro_category_wide, by = "ARO_Term") %>%
    left_join(overlap_membership_tbl, by = "ARO_Term") %>%
    arrange(desc(mean_cpm_overall), ARO_Term)
  
  # Export comprehensive long annotation table to Excel
  if (requireNamespace("writexl", quietly = TRUE)) {
    writexl::write_xlsx(
      list(
        rare_aro_annotation = rare_aro_annotation_long,
        rare_aro_by_category = rare_aro_summary,
        overlap_membership = overlap_membership_tbl,
        presence_table = presence_tbl,
        overlap_counts = overlap_tbl
      ),
      "../results/rare_exposed_aro_terms_comprehensive_annotation.xlsx"
    )
    cat("✓ Wrote Excel file: rare_exposed_aro_terms_comprehensive_annotation.xlsx\n")
  } else {
    message("Package 'writexl' not installed; CSV export was written instead of XLSX.")
  }
  
  cat("Saved rare exposed ARO-term visualizations and summaries.\n")
}
```

### Comprehensive ARO-term annotation for all 1502 terms

```{r}
# Build comprehensive annotation table for ALL ARO-terms (not filtered to rare exposed)
# This mirrors the structure from ARO_Term.qmd for supplementary data

# 1) Ensure drug class annotation is loaded
if (!exists("drug_class_annot_manual") || nrow(drug_class_annot_manual) == 0) {
  cat("Loading drug class annotation...\n")
  drug_class_annot_manual <- read_excel("../drug_class_fam_mean_cpm.xlsx")
}

if (!exists("drug_class_resfinder_map") || nrow(drug_class_resfinder_map) == 0) {
  drug_class_resfinder_map <- drug_class_annot_manual %>%
    select(Drug_Class, Drug_Class_Resfinder) %>%
    distinct()
}

# 2) Build all-terms CPM matrix long format
cpm_aro_all_long <- cpm_aro_hyp_mat %>%
  as.data.frame() %>%
  rownames_to_column("ARO_Term") %>%
  pivot_longer(-ARO_Term, names_to = "sample", values_to = "cpm_aro") %>%
  left_join(meta %>% rownames_to_column("sample") %>% select(sample, category), by = "sample") %>%
  filter(!is.na(ARO_Term), ARO_Term != "")

# 3) Compute overall and category-wise statistics
aro_mean_cpm_all <- cpm_aro_all_long %>%
  group_by(ARO_Term) %>%
  summarise(
    mean_cpm_aro = mean(cpm_aro, na.rm = TRUE),
    median_cpm_aro = median(cpm_aro, na.rm = TRUE),
    .groups = "drop"
  )

aro_median_cpm_all_by_category <- cpm_aro_all_long %>%
  filter(!is.na(category), category != "") %>%
  group_by(ARO_Term, category) %>%
  summarise(
    median_cpm_aro_category = median(cpm_aro, na.rm = TRUE),
    .groups = "drop"
  )

aro_median_cpm_all_by_category_wide <- aro_median_cpm_all_by_category %>%
  pivot_wider(
    names_from = category,
    values_from = median_cpm_aro_category,
    names_prefix = "median_cpm_"
  )

# 4) Map ARO terms to Drug_Class_Resfinder
aro_to_resfinder_all <- feats %>%
  rownames_to_column("Prevalence_Sequence_ID") %>%
  select(Prevalence_Sequence_ID, ARO_Term = any_of("ARO_Term"), Drug_Class) %>%
  filter(!is.na(ARO_Term), ARO_Term != "", !is.na(Drug_Class), Drug_Class != "") %>%
  mutate(Drug_Class = str_trim(Drug_Class)) %>%
  separate_rows(Drug_Class, sep = ";\\s*") %>%
  mutate(Drug_Class = str_trim(Drug_Class)) %>%
  filter(!is.na(Drug_Class), Drug_Class != "") %>%
  left_join(drug_class_resfinder_map, by = "Drug_Class") %>%
  filter(!is.na(Drug_Class_Resfinder), Drug_Class_Resfinder != "") %>%
  distinct(ARO_Term, Drug_Class_Resfinder)

# 5) Count mapping diversity
aro_class_counts_all <- aro_to_resfinder_all %>%
  dplyr::count(ARO_Term, name = "n_resfinder_classes") %>%
  mutate(
    weight_1_over_n = 1 / n_resfinder_classes,
    multi_annotation_flag = n_resfinder_classes > 1
  )

# 6) Build comprehensive annotation table (all terms)
aro_resfinder_annotation_all <- aro_to_resfinder_all %>%
  left_join(aro_mean_cpm_all, by = "ARO_Term") %>%
  left_join(aro_median_cpm_all_by_category_wide, by = "ARO_Term") %>%
  left_join(aro_class_counts_all, by = "ARO_Term") %>%
  mutate(weighted_mean_cpm_contribution = mean_cpm_aro * weight_1_over_n) %>%
  arrange(Drug_Class_Resfinder, desc(mean_cpm_aro), ARO_Term)

cat("✓ Built comprehensive annotation for", n_distinct(aro_resfinder_annotation_all$ARO_Term), "ARO-terms\n")
cat("  Across", n_distinct(aro_resfinder_annotation_all$Drug_Class_Resfinder), "drug class categories\n")

# 7) Export to CSV and Excel
write_csv(
  aro_resfinder_annotation_all,
  "../results/aro_term_resfinder_annotation_all.csv"
)

write_csv(
  aro_median_cpm_all_by_category,
  "../results/aro_term_median_cpm_all_by_category.csv"
)

if (requireNamespace("writexl", quietly = TRUE)) {
  writexl::write_xlsx(
    list(
      aro_term_resfinder_annotation = aro_resfinder_annotation_all,
      aro_term_median_cpm_by_category = aro_median_cpm_all_by_category,
      aro_term_class_counts = aro_class_counts_all
    ),
    "../results/aro_term_resfinder_annotation.xlsx"
  )
  cat("✓ Wrote Excel file: aro_term_resfinder_annotation_all_hypothesis.xlsx\n")
} else {
  message("Package 'writexl' not installed; CSV export was written instead of XLSX.")
}
```

### ARO-term driver analysis (contribution models + bubble plot)

```{r}
# Driver table at ARO-term level based on per-sample relative abundance contributions.
aro_term_contrib_long <- cpm_aro_hyp_long %>%
  left_join(sample_metrics_hyp %>% select(sample, category, depth), by = "sample") %>%
  group_by(sample) %>%
  mutate(
    total_cpm_sample = sum(cpm_aro, na.rm = TRUE),
    p = if_else(total_cpm_sample > 0, cpm_aro / total_cpm_sample, 0),
    shannon_contrib = if_else(p > 0, -p * log(p), 0),
    simpson_contrib = p^2,
    is_rare_1_10 = cpm_aro >= 1 & cpm_aro < 10,
    p_rank = min_rank(desc(p)),
    is_top10 = p_rank <= 10 & p > 0
  ) %>%
  ungroup() %>%
  select(sample, category, depth, ARO_Term, cpm_aro, p, shannon_contrib, simpson_contrib, is_rare_1_10, is_top10)

fit_term_contrib <- function(df_term, response_col) {
  if (n_distinct(df_term$category) < 2) {
    return(tibble(
      p_category = NA_real_,
      p_depth = NA_real_,
      mean_contrib = mean(df_term[[response_col]], na.rm = TRUE),
      prevalence_nonzero = mean(df_term[[response_col]] > 0, na.rm = TRUE),
      effect_pigs_vs_controls = NA_real_,
      effect_farmers_vs_controls = NA_real_,
      effect_pigs_vs_farmers = NA_real_,
      max_abs_effect = NA_real_,
      dominant_group = NA_character_
    ))
  }

  fit <- tryCatch(
    lm(as.formula(paste(response_col, "~ category + depth")), data = df_term),
    error = function(e) NULL
  )

  if (is.null(fit)) {
    return(tibble(
      p_category = NA_real_,
      p_depth = NA_real_,
      mean_contrib = mean(df_term[[response_col]], na.rm = TRUE),
      prevalence_nonzero = mean(df_term[[response_col]] > 0, na.rm = TRUE),
      effect_pigs_vs_controls = NA_real_,
      effect_farmers_vs_controls = NA_real_,
      effect_pigs_vs_farmers = NA_real_,
      max_abs_effect = NA_real_,
      dominant_group = NA_character_
    ))
  }

  a <- anova(fit)

  med_tbl <- df_term %>%
    group_by(category) %>%
    summarise(med = median(.data[[response_col]], na.rm = TRUE), .groups = "drop")

  med_controls <- med_tbl$med[med_tbl$category == "Non_farmer_controls"]
  med_farmers <- med_tbl$med[med_tbl$category == "Pig_farmers"]
  med_pigs <- med_tbl$med[med_tbl$category == "Pigs"]

  med_controls <- ifelse(length(med_controls) == 0, NA_real_, med_controls)
  med_farmers <- ifelse(length(med_farmers) == 0, NA_real_, med_farmers)
  med_pigs <- ifelse(length(med_pigs) == 0, NA_real_, med_pigs)

  effect_pc <- med_pigs - med_controls
  effect_fc <- med_farmers - med_controls
  effect_pf <- med_pigs - med_farmers

  dominant_group <- med_tbl %>%
    filter(is.finite(med)) %>%
    arrange(desc(med)) %>%
    slice_head(n = 1) %>%
    pull(category)

  if (length(dominant_group) == 0) dominant_group <- NA_character_

  tibble(
    p_category = a["category", "Pr(>F)"],
    p_depth = a["depth", "Pr(>F)"],
    mean_contrib = mean(df_term[[response_col]], na.rm = TRUE),
    prevalence_nonzero = mean(df_term[[response_col]] > 0, na.rm = TRUE),
    effect_pigs_vs_controls = effect_pc,
    effect_farmers_vs_controls = effect_fc,
    effect_pigs_vs_farmers = effect_pf,
    max_abs_effect = max(abs(c(effect_pc, effect_fc, effect_pf)), na.rm = TRUE),
    dominant_group = as.character(dominant_group)
  )
}

driver_shannon <- aro_term_contrib_long %>%
  group_by(ARO_Term) %>%
  group_modify(~ fit_term_contrib(.x, "shannon_contrib")) %>%
  ungroup() %>%
  mutate(
    p_adj = p.adjust(p_category, method = "BH"),
    response = "shannon_contrib"
  ) %>%
  arrange(p_adj, desc(max_abs_effect))

driver_simpson <- aro_term_contrib_long %>%
  group_by(ARO_Term) %>%
  group_modify(~ fit_term_contrib(.x, "simpson_contrib")) %>%
  ungroup() %>%
  mutate(
    p_adj = p.adjust(p_category, method = "BH"),
    response = "simpson_contrib"
  ) %>%
  arrange(p_adj, desc(max_abs_effect))

top_driver_shannon <- driver_shannon %>%
  filter(!is.na(p_adj)) %>%
  slice_head(n = 50)

top_driver_simpson <- driver_simpson %>%
  filter(!is.na(p_adj)) %>%
  slice_head(n = 50)

# Bubble plot requested: x = effect size, y = -log10(BH p), size = mean contribution.
bubble_df <- driver_shannon %>%
  filter(is.finite(max_abs_effect), is.finite(p_adj), is.finite(mean_contrib)) %>%
  mutate(score = -log10(p_adj + 1e-300)) %>%
  arrange(desc(score), desc(max_abs_effect)) %>%
  slice_head(n = 120)

label_df <- bubble_df %>%
  mutate(label_rank = score * max_abs_effect) %>%
  arrange(desc(label_rank)) %>%
  slice_head(n = 20)

p_driver_bubble <- ggplot(
  bubble_df,
  aes(x = max_abs_effect, y = score, size = mean_contrib, color = dominant_group)
) +
  geom_point(alpha = 0.8) +
  ggrepel::geom_text_repel(
    data = label_df,
    aes(label = ARO_Term),
    size = 2.8,
    max.overlaps = 40,
    box.padding = 0.25,
    point.padding = 0.2,
    segment.alpha = 0.5,
    show.legend = FALSE
  ) +
  scale_size_continuous(name = "Mean contribution") +
  labs(
    title = "ARO-term drivers of Shannon contribution differences",
    subtitle = "x = max absolute median effect across group contrasts, y = -log10(BH-adjusted p)",
    x = "Effect size (max absolute median difference)",
    y = "-log10(BH-adjusted p-value)",
    color = "Highest median group"
  )

print(p_driver_bubble)

write_csv(aro_term_contrib_long, "../results/aro_term_contribution_long_table.csv")
write_csv(driver_shannon, "../results/aro_term_driver_shannon.csv")
write_csv(driver_simpson, "../results/aro_term_driver_simpson.csv")
write_csv(top_driver_shannon, "../results/aro_term_driver_shannon_top50.csv")
write_csv(top_driver_simpson, "../results/aro_term_driver_simpson_top50.csv")

ggsave(
  "../figures/aro_term_driver_bubble_shannon.svg",
  p_driver_bubble,
  width = 10,
  height = 7,
  web_fonts = "DejaVu Sans"
)
```