---
title: "Alypha Diversity Analysis"
---

# $𝛼$ diversity {#sec-alpha-diversity}

$𝛼$ diversity is calculated on the raw reads

## Dependencies

`utils.R` **-script**

```{r}
source("utils.R")
```

**Packages**

```{r}
#| echo: true
#| message: false
#| warning: false
#| results: hide


#devtools::install_github("gmteunisse/fantaxtic")
#devtools::install_github("gmteunisse/ggnested")
library(tidyverse)
library(BiocManager)
library(phyloseq)
library(ggnested)
library(vegan)

# for font
library(ggplot2)
library(ggpubr)
library(patchwork)
library(showtext)
library(systemfonts)
library(rstatix)

font_add("DejaVu Sans", regular = systemfonts::font_info("DejaVu Sans")$path[1])
showtext_auto()

font_families()


# load theme
theme_publication <- function(base_size = 14, base_family = "DejaVu Sans") {
ggplot2::theme_bw(
base_size = base_size,
base_family = base_family
) +
ggplot2::theme(
plot.title = ggplot2::element_text(hjust = 0.5),
axis.title = ggplot2::element_text(size = ggplot2::rel(1.05)),
axis.text = ggplot2::element_text(size = ggplot2::rel(1.2)),
legend.title = ggplot2::element_text(size = ggplot2::rel(0.95)),
legend.text = ggplot2::element_text(size = ggplot2::rel(0.9)),
strip.text = ggplot2::element_text(size = ggplot2::rel(0.95))
)
}

```

**Color Palette**

```{r - load color palette }
# Load color codes from YAML file
color_codes <- yaml::read_yaml("../../Metadata/color_codes.yaml")

# Extract specific color palettes
sample_type_colors <- unlist(color_codes$sample_types)
pig_stage_colors <- unlist(color_codes$pig_stages)
category2_colors <- unlist(color_codes$category2)
```

## Data: Loading and Filtering

-   MetaPhlAn-Output estimated count data table (MF_SGB_abundance.txt)
-   MetaPhlAn-Output estimated read count based on all taxonomic levels (MF_abundance.txt)
-   metaData

```{r - load data}
# metaphlan raw read data (MF_SGB_abundance.txt)
metaphlanData <- read.table("../data/raw/MF_SGB_abundance.txt", header = TRUE, sep = "")
metaphlanData$clade_name <- as.character(metaphlanData$clade_name)

# load metadata
metaData <- read.table("../../metadata/metadata_silas.csv", header = TRUE, sep = ",") %>%
  column_to_rownames(var = "id")

# assigns 0 to all NA's in my metaphlan data table
metaphlanData[is.na(metaphlanData)] <- 0

# load the MF_abundance.txt file for getting access to `UNCLASSIFIED`reads and filter those reads
MF_abundance <- read.table("../data/raw/MF_abundance.txt", header = T, sep = "") %>%
  filter(grepl("UNCLASSIFIED", clade_name))

# adjust the colnames of MF_abundance to match with metaphlanData
colnames(MF_abundance) <- gsub("^X", "C", colnames(MF_abundance))


# marge MF_abudnance with metaphlanData
metaphlanData_all <- rbind(metaphlanData, MF_abundance)

```

-   find species which are assigned to UNKOWN

    ```{r - find `UKNOWN` assignemnt}
    # Assuming your MetaPhlAn table is in a data frame called mpa
    # Count the number of taxonomic levels per clade
    mpa <- metaphlanData
    # Assuming your MetaPhlAn table is in a data frame called mpa
    # Count the number of taxonomic levels per clade
    mpa$tax_level_count <- sapply(strsplit(mpa$clade_name, "\\|", fixed = FALSE), length)

    # View rows with fewer than 8 levels (less than full taxonomy path)
    mpa_incomplete <- mpa[mpa$tax_level_count < 8, ]

    # Show unique clade names that are incomplete
    unique(mpa_incomplete$clade_name)

    ```

-   filtering of all duplicated samples (ppB)

```{r - filter metaphlan Data }
# metaphlanData
metaphlanData <- metaphlanData[, !grepl("ppB", colnames(metaphlanData))]


```

## Sample Numbers per Group

```{r - sample numbers metadata barplots}
# sample count per category
meta_counts_category <- metaData %>%
  rownames_to_column(var = "sample") %>%
  dplyr::count(category, name = "n") %>%
  mutate(
    category = factor(category,
                      levels = c("Non_farmer_controls", "Pig_farmers", "Pigs"))
  )

# define growth-stage legend for all categories
meta_counts_stage <- metaData %>%
  rownames_to_column(var = "sample") %>%
  mutate(
    growth_stage_plot = case_when(
      category == "Non_farmer_controls" ~ "Control",
      is.na(growth_stage) ~ "Unknown",
      TRUE ~ as.character(growth_stage)
    )
  ) %>%
  dplyr::count(category, growth_stage_plot, name = "n") %>%
  mutate(
    category = factor(category,
                      levels = c("Non_farmer_controls", "Pig_farmers", "Pigs"))
  )

growth_stage_colors_plot <- c(
  "Control" = sample_type_colors[["Non_farmer_controls"]],
  pig_stage_colors
)

if ("Unknown" %in% meta_counts_stage$growth_stage_plot) {
  growth_stage_colors_plot <- c(growth_stage_colors_plot, "Unknown" = "grey70")
}

# Plot A: total samples per category
p_counts_category <- ggplot(meta_counts_category,
                            aes(x = category, y = n, fill = category)) +
  geom_col(width = 0.6, color = "black") +
  geom_text(aes(label = n), vjust = 1.4, size = 5, family = "DejaVu Sans") +
  scale_fill_manual(values = sample_type_colors) +
  scale_x_discrete(labels = color_codes$label_mapping) +
  labs(
    # title = "Number of Samples per Category",
    x = "",
    y = "Sample count"
  ) +
  theme_publication() +
  theme(legend.position = "none")

# Plot B: category split by growth stage (controls shown as Control)
p_counts_stage <- ggplot(meta_counts_stage,
                         aes(x = category, y = n, fill = growth_stage_plot)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, color = "black") +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.8),
    vjust = 1.4,
    size = 4.5,
    family = "DejaVu Sans"
  ) +
  scale_fill_manual(values = growth_stage_colors_plot) +
  scale_x_discrete(labels = color_codes$label_mapping) +
  labs(
    #title = "Sample Counts by Category and Growth Stage",
    x = "",
    y = "Sample count",
    fill = "Grwoth Stage"
  ) +
  theme_publication()

print(p_counts_category)
print(p_counts_stage)

# save plots
ggsave("../figures/sample_counts_per_category.svg", plot = p_counts_category, width = 6, height = 7)
ggsave("../figures/sample_counts_category_growth_stage.svg", plot = p_counts_stage, width = 7, height = 6.5)
```

## Phyloseq-Object

```{r - create phyloseq-object}
meta_physeq <- metaphlan_to_phyloseq(metaphlanData, metaData, version = 4, verbose = TRUE, tax_lvl = "SGB")

ps <- meta_physeq
ps <- subset_taxa(ps, Kingdom !="k__Eukaryota") # remove eukaryotes

#View(ps@tax_table@.Data)

ps@tax_table@.Data <- substring(ps@tax_table@.Data, 4)
taxa_names(ps) <- substring(taxa_names(ps), 4)
colnames(ps@tax_table@.Data)<- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "SGB")
```

## Richness

```{r - richness plots}
# calculate sample counts
sample_counts <- data.frame(ps@sam_data) %>%
  group_by(category) %>%
  summarise(n = n(), .groups = 'drop')


# plot richness with amount of samples
plot_richness(ps, x = "category", color = "category",
              scales = "free_y")+
  scale_color_manual(values = sample_type_colors)+
  geom_violin()+
  geom_jitter(alpha = 0.5)+
  geom_text(data = sample_counts,
            aes(x = category, y = 0, label = paste0("n=",n)),
                vjust = 2, size = 3, color ="black")+
  theme_publication()+
  ggtitle("Species Richness per Category")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

### save the plot

```{r - save richness plot }

# save richness plot 
ggsave("../figures/richness_per_category.svg", width = 15, height = 6)
```

## 𝛼-diversity

### dataframe with all relevant data

we create a dataframe which includes 𝛼 diversity metrix (Shannon) and different other parameters we need:

-   total read count of all classified reads

-   total read count of all unclassified reads

-   sum of classified + unclassified

-   merged with metadata to make the plot

```{r - alpha diversity dataframe}
library(dplyr)
library(tibble)

# total `CLASSIFIED` read count per sample
classified <- data.frame(t(ps@otu_table)) %>%
  mutate(classified_reads = rowSums(.)) %>%
  select(classified_reads)

# extract Eukaryotic samples in my metaphlan Data
eukaryotic <- metaphlanData_all %>%
  filter(grepl("k__Eukaryota", clade_name)) %>%
  select(-c("clade_name", "clade_taxid")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "sample",
    values_to = "eukaryotic_reads"
  ) %>% 
  group_by(sample) %>%
  summarise(eukaryotic_reads = sum(eukaryotic_reads), .groups = "drop") %>%
  column_to_rownames(var = "sample")


# total `UNCLASSIFIED`
unclassified <- metaphlanData_all%>%
  filter(clade_name == "UNCLASSIFIED") %>%
  select(-c("clade_name", "clade_taxid")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "sample",
    values_to = "unclassified_reads"
  ) %>% 
  left_join(eukaryotic %>% rownames_to_column(var = "sample"), by = "sample") %>% # add eukaryotic reads
  mutate(
    unclassified_reads = unclassified_reads + eukaryotic_reads) %>% # add eukaryotic reads to unclassified reads
  column_to_rownames(var = "sample")


# calculate alpha diversity
df_ps <- as.data.frame(otu_table(ps))
df_ps <- t(df_ps)

shannon <- diversity(df_ps, index = "shannon") # diversity function from vegan package to calculate shannon diversity 
simpson <- diversity(df_ps, index = "simpson")

# make data.frame 
alpha_diversity <- data.frame(
  sample = rownames(df_ps),
  shannon = shannon,
  simpson = simpson) %>%
  left_join(classified %>% rownames_to_column(var = "sample"), by = "sample") %>%
  left_join(metaData %>% rownames_to_column(var = "sample"), by = "sample") %>%
  left_join(unclassified %>% rownames_to_column(var = "sample"), by = "sample") %>%
  mutate(
    total_reads = classified_reads + unclassified_reads
  ) %>%
  mutate(
    classified_total_ratio = classified_reads / total_reads
  )
```

### Classified vs. Unclassified reads

this plot should quickly show how much unclassified reads we have per category (Pig, Pigfarmer, Controls).

The unclassified reads we find in MF_abudnance.txt as they are not present in SGB_abundance.txt.

#### barplot - classified vs unclassified reads

```{r - barplot unclassified vs. classified reads}

barplot_classified_unclassified <- alpha_diversity %>%
  group_by(category)%>%
  summarise(
    mean_classified = mean(classified_reads),
    mean_unclassified = mean(unclassified_reads), 
    .groups = "drop") %>%
  pivot_longer(cols = starts_with("mean_"),
               names_to = "type", 
               values_to = "mean_reads"
  )

library(viridis)
# bar plot
barplot <- ggplot(barplot_classified_unclassified, 
                  aes(fill = type, y = mean_reads, x = category))+
  geom_bar(stat = "identity", position = "stack", width = 0.5, color ="black")+
  theme_publication()+
  scale_x_discrete(labels = color_codes$label_mapping)+
  ylab("Mean Reads per Category")+
  scale_fill_brewer(palette = "Dark2")+
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.text.x = element_text(size = 11)
  )
  
print(barplot)


```

**save this plot**

```{r - save barplot classified vs unclassified}
ggsave("../figures/barplot_classified_unclassified_reads.svg", plot = barplot,
       width = 6, height = 6)
```

#### boxplot classified to total ratio boxplot

```{r - boxplot classified to total ratio}
# boxplot with suckling / older pigs separated 

# set the order correctly in the plot
alpha_diversity$category2 <- factor(alpha_diversity$category2,
                                       levels = c("Suckling_piglets",
                                                  "Older_pigs",
                                                  "Pig_farmers",
                                                  "Non_farmer_controls"))


# set the order of catergory correctly
alpha_diversity$category <- factor(alpha_diversity$category,
                                   levels = c("Pigs",
                                              "Pig_farmers",
                                              "Non_farmer_controls"))

# custom formatter: keep tiny p-values readable
ratio_p_label <- function(p_value) {
  ifelse(p_value < 0.001, "p < 0.001", paste0("p = ", sprintf("%.3f", p_value)))
}


ratio_pvals <- alpha_diversity %>%
  pairwise_wilcox_test(classified_total_ratio ~ category, p.adjust.method = "BH") %>%
  add_xy_position(x = "category") %>%
  mutate(p_label = ratio_p_label(p.adj))



boxplot_classified_ratio <- ggplot(alpha_diversity,
                                   aes(x = category, y = classified_total_ratio, fill = category))+
  geom_boxplot(outlier.shape = NA, alpha = 0.5)+
  geom_jitter(alpha = 0.3, width = 0.2)+
  scale_fill_manual(values = sample_type_colors, drop = F)+
  stat_pvalue_manual(
    ratio_pvals,
    label = "p_label",
    hide.ns = TRUE,
    tip.length = 0.01,
    size = 4,
    family = "DejaVu Sans"
  )+
  #ggtitle("Classified to Total Reads Ratio per Category")+
  xlab("")+
  ylab("Classified / Total estimated reads")+
  scale_x_discrete(labels = color_codes$label_mapping)+
  theme(legend.position = "none")+
  theme_publication()+
  theme(legend.position = "none")


print(boxplot_classified_ratio)
                      
```

**save this plot**

```{r - save boxplot}
ggsave("../figures/boxplot_classified_total_reads_ratio.svg", plot = boxplot_classified_ratio,
       width = 6, height = 7)
```

### 𝛼- Diversity violin plots (not for paper)

$alpha$ diversity violin plots

-   A: pigs / pig workers / ctrls

-   B: pigs per stage

-   C: per farm (only pig data)

```{r - alpha diversity plot without read depth accounting}
# plot A: pig / pig workers / ctrls
plot_alpha_A <- ggplot(
  data = alpha_diversity, 
  aes(x = category, y = shannon, fill = category))+
    geom_violin(trim = FALSE, alpha = 0.5, position = "dodge")+
    geom_boxplot(width = 0.1, outlier.shape = NA, fill = "grey", notch= TRUE)+
  scale_fill_manual(values = sample_type_colors)+
  geom_jitter(alpha = 0.3, width = 0.2)+
  theme_publication()+
  ggtitle("Alpha Diversity of Pigs, Pig Worker and Controls")

plot(plot_alpha_A)

# plot B: pigs per stage

plot_alpha_B <- alpha_diversity %>%
  filter(species == "Pig") %>%
  mutate(growth_stage = factor(growth_stage,
                               levels = c("Suckling", "Weaning", "Fattening"))) %>%
  ggplot(aes(x = growth_stage, y = shannon, fill = growth_stage)) +
    geom_violin(trim = FALSE, alpha = 0.6, position = "dodge") +
    geom_boxplot(width = 0.1, outlier.shape = NA, fill = "grey", notch = TRUE) +
    scale_fill_manual(values = pig_stage_colors) +
    theme_publication() +
    geom_jitter(alpha = 0.3, width = 0.2) +
    labs(x = "", y = "Shannon diversity index (H)",
         title = "Alpha Diversity of Suckling, Weaning and Fattening pigs")

plot(plot_alpha_B)

# plot C: farms (only pig data)
plot_alpha_C <- alpha_diversity %>%
  filter(species == "pig") %>%
  ggplot(
    aes(x = farm_number, y = shannon, shape = stage))+
  geom_point()

plot(plot_alpha_C)
```

**save the plots**

```{r - save alpha diversity plots wiihtout read depth accounting}
ggsave("../figures/alpha_diversity_A_species.svg", plot = plot_alpha_A, width = 8, height = 6)
ggsave("../figures/alpha_diversity_B_stages_pigs.svg", plot = plot_alpha_B, width = 8, height = 6)
```

### 𝛼 diversity with accounting for read depth

#### linear model plot - by species

With a linear model, we are statistically evaluating whether different groups (defined by your "category" variable) have significant differences in Shannon diversity, while at the same time adjusting for the confounding effect of sequencing depth (total_reads).

the model fits Shannon diversity as a function of our sample categories and the total number of reads for each sample (sequencing depth). By including total_reads as a covariate, we control for the fact that samples with more reads might artificially appear to have higher diversity due to deeper sequencing. Therefore, true groups effects can be isolated from technical variability.

With the `emmenans` package we compute pairwise differences between all the group levels for the category variable and we adjust the p-value with Tukey test to correct for multiple comparisons.

```{r - alpha LM model species }
lm_model<- lm(shannon ~ category + total_reads, data = alpha_diversity)


plot(lm_model)

# pairwise group comparison
library(emmeans)
pairwise_results <- emmeans(lm_model, pairwise ~ category, adjust = "tukey")


# extract p values
pvals <- summary(pairwise_results$contrasts)


# p-value dataframe
pval_df <- pvals %>%
  mutate(
    group1 = sub(" - .*", "", contrast),
    group2 = sub(".* - ", "", contrast)  # add space before group2 if needed
  ) %>%
  select(group1, group2, p.value) %>%
  mutate(
    y.position = max(alpha_diversity$shannon, na.rm = TRUE) * 
                   (1 + 0.1 * row_number())
  )



# violin plot
library(ggplot2)
library(ggpubr)
library(scales) # for p-value formatting



clean_p <- function(x) {
  ifelse(
    x < 0.001,
    "<0.001",   
    sprintf("%.3f", x)           # e.g. 0.155
  )
}

pval_df$p_label <- clean_p(pval_df$p.value)

# define the order of groups
alpha_diversity$category <- factor(alpha_diversity$category,
                                   levels = c("Non_farmer_controls",
                                              "Pig_farmers",
                                              "Pigs"))

# plot violin plots including LM model
alpha_diversity_lm_plot <- ggboxplot(
    alpha_diversity,
    x = "category",
    y = "shannon",
    fill = "category"
  )+
  geom_jitter(alpha = 0.3, width = 0.2)+
  facet_wrap(~"Alpha Diversity per Group", nrow = 1)+
  scale_fill_manual(values = sample_type_colors)+
  theme_bw(base_family = "DejaVu Sans")+
  stat_pvalue_manual(
      pval_df,
      label = "p_label",
      xmin = "group1", xmax = "group2",
      y.position = "y.position",
      tip.length = 0.02,
      family = "DejaVu Sans",
      size = 6
    ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))+
  ylab("Shannon Diversity Index")+
  xlab("")+
  scale_x_discrete(labels = color_codes$label_mapping)+
  theme_publication()+
  theme(
    legend.position = "none",
    # axis.text.x = element_text(size = 11),
    # axis.text.y = element_text(size = 11),
    #axis.title.y = element_text(size = 12),
    strip.background = element_rect(fill = "grey85", color = "black"),
    strip.text = element_text(
      face ="bold", 
      size = 14,
      family = "DejaVu Sans"),
    plot.margin = margin(5.5,5.5,5.5,5.5)
    )
  

print(alpha_diversity_lm_plot)
```

**save this figure**

```{r - save alpha LM model species}
ggsave("../figures/alpha_diversity_species_LM.svg", plot =alpha_diversity_lm_plot, width = 8, height = 8 )
```

#### linear model - by pig grwoth stage

```{r linear model by stage}
library(scales) # for p-value formatting

# filter alpha diversity for pigs only
alpha_diversity_pigs <- alpha_diversity %>%
  filter(species == "Pig")

# make linear model
lm_model_stage <- lm(shannon ~ growth_stage + total_reads, data = alpha_diversity_pigs)

# pairwise group comparison
library(emmeans)
pairwise_results_stage <- emmeans(lm_model_stage, pairwise ~ growth_stage, adjust = "tukey")

# extract p values
pvals_stage <- summary(pairwise_results_stage$contrasts)


# p-value dataframe
pval_stage_df <- pvals_stage %>%
  mutate(
    group1 = sub(" - .*", "", contrast),
    group2 = sub(".* - ", "", contrast)  # add space before group2 if needed
  ) %>%
  select(group1, group2, p.value) %>%
  mutate(
    y.position = max(alpha_diversity$shannon, na.rm = TRUE) * 
                   (1 + 0.1 * row_number()),
    p.label = clean_p(p.value)
  )


# adjust the factor for correct order
alpha_diversity_pigs$growth_stage <- factor(alpha_diversity_pigs$growth_stage,
                                            levels= c("Suckling",
                                                      "Weaning",
                                                      "Fattening"))

# violin plot
library(ggplot2)
library(ggpubr)

alpha_diversity_stage_lm_plot <- ggboxplot(
    alpha_diversity_pigs,
    x = "growth_stage",
    y = "shannon",
    add = "boxplot", 
    fill = "growth_stage"
  )+
  geom_jitter(alpha = 0.3, width = 0.2)+
  facet_wrap(~"Alpha Diversity per Pig Growth Stage", nrow = 1)+
  scale_fill_manual(values = pig_stage_colors)+
  theme_bw(base_family = "DejaVu Sans")+
  stat_pvalue_manual(
      pval_stage_df,
      label = "p.label",
      xmin = "group1", xmax = "group2",
      y.position = "y.position",
      tip.length = 0.02,
      family = "DejaVu Sans",
      size = 6
    ) +
  xlab("")+
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))+
  xlab("Pig Growth Stage")+
  theme_publication()+
  theme(legend.position = "none",
        strip.background = element_rect(fill = "grey85", color = "black"),
        strip.text = element_text(size = 13,
                                  family = "DejaVu Sans"),
        # axis.text.x = element_text(size = 11),
        # axis.text.y = element_text(size = 11),
        # axis.title.y = element_text(size = 12),
        axis.title.x = element_blank(),
        plot.margin = margin(5.5,5.5,5.5,0))+
  ylim(1,8)

print(alpha_diversity_stage_lm_plot)

```

**save this plot**

```{r - save this plot}
ggsave("../figures/alpha_diversity_stage_LM.svg", plot = alpha_diversity_stage_lm_plot, height = 8, width = 8 )
```

### final alpha diversity plot
```{r}

# right blot (stage)
alpha_diversity_stage_lm_plot_r <- alpha_diversity_stage_lm_plot +
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  )

alpha_diversity_stage_lm_plot_r

library(patchwork)

combined <- alpha_diversity_lm_plot + alpha_diversity_stage_lm_plot_r +
  plot_layout(ncol = 2, guides = "collect") & theme(
  axis.text.x = element_text(size = 17),
  axis.text.y = element_text(size = 18))

combined

ggsave("../figures/alpha_diversity_combined_LM.svg", plot = combined,
       height = 6, 
       width = 8.5,
       device = svglite::svglite)

```

### Simpson diversity with accounting for read depth

```{r - simpson LM model species}
# linear model for simpson by category
lm_model_simpson <- lm(simpson ~ category + total_reads, data = alpha_diversity)

pairwise_results_simpson <- emmeans(lm_model_simpson, pairwise ~ category, adjust = "tukey")
pvals_simpson <- summary(pairwise_results_simpson$contrasts)

pval_simpson_df <- pvals_simpson %>%
  mutate(
    group1 = sub(" - .*", "", contrast),
    group2 = sub(".* - ", "", contrast)
  ) %>%
  select(group1, group2, p.value) %>%
  mutate(
    y.position = max(alpha_diversity$simpson, na.rm = TRUE) * (1 + 0.08 * row_number()),
    p.label = clean_p(p.value)
  )

alpha_diversity$category <- factor(alpha_diversity$category,
                                   levels = c("Non_farmer_controls", "Pig_farmers", "Pigs"))

simpson_lm_plot <- ggboxplot(
  alpha_diversity,
  x = "category",
  y = "simpson",
  fill = "category"
) +
  geom_jitter(alpha = 0.3, width = 0.2) +
  facet_wrap(~"Simpson Diversity per Group", nrow = 1) +
  scale_fill_manual(values = sample_type_colors) +
  theme_bw(base_family = "DejaVu Sans") +
  stat_pvalue_manual(
    pval_simpson_df,
    label = "p.label",
    xmin = "group1", xmax = "group2",
    y.position = "y.position",
    tip.length = 0.02,
    family = "DejaVu Sans",
    size = 6
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  ylab("Simpson Diversity Index") +
  xlab("") +
  scale_x_discrete(labels = color_codes$label_mapping) +
  theme_publication() +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "grey85", color = "black"),
    strip.text = element_text(face = "bold", size = 14, family = "DejaVu Sans"),
    plot.margin = margin(5.5, 5.5, 5.5, 5.5)
  )

print(simpson_lm_plot)
ggsave("../figures/simpson_diversity_species_LM.svg", plot = simpson_lm_plot, width = 8, height = 8)
```

```{r - simpson LM model stage}
# linear model for simpson by pig growth stage
alpha_diversity_pigs_simpson <- alpha_diversity %>%
  filter(species == "Pig") %>%
  mutate(growth_stage = factor(growth_stage, levels = c("Suckling", "Weaning", "Fattening")))

lm_model_stage_simpson <- lm(simpson ~ growth_stage + total_reads, data = alpha_diversity_pigs_simpson)
pairwise_results_stage_simpson <- emmeans(lm_model_stage_simpson, pairwise ~ growth_stage, adjust = "tukey")
pvals_stage_simpson <- summary(pairwise_results_stage_simpson$contrasts)

pval_stage_simpson_df <- pvals_stage_simpson %>%
  mutate(
    group1 = sub(" - .*", "", contrast),
    group2 = sub(".* - ", "", contrast)
  ) %>%
  select(group1, group2, p.value) %>%
  mutate(
    y.position = max(alpha_diversity_pigs_simpson$simpson, na.rm = TRUE) * (1 + 0.08 * row_number()),
    p.label = clean_p(p.value)
  )

simpson_stage_lm_plot <- ggboxplot(
  alpha_diversity_pigs_simpson,
  x = "growth_stage",
  y = "simpson",
  add = "boxplot",
  fill = "growth_stage"
) +
  geom_jitter(alpha = 0.3, width = 0.2) +
  facet_wrap(~"Simpson Diversity per Pig Growth Stage", nrow = 1) +
  scale_fill_manual(values = pig_stage_colors) +
  theme_bw(base_family = "DejaVu Sans") +
  stat_pvalue_manual(
    pval_stage_simpson_df,
    label = "p.label",
    xmin = "group1", xmax = "group2",
    y.position = "y.position",
    tip.length = 0.02,
    family = "DejaVu Sans",
    size = 6
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  xlab("") +
  ylab("Simpson Diversity Index") +
  theme_publication() +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "grey85", color = "black"),
    strip.text = element_text(size = 13, family = "DejaVu Sans"),
    axis.title.x = element_blank(),
    plot.margin = margin(5.5, 5.5, 5.5, 0)
  )

print(simpson_stage_lm_plot)
ggsave("../figures/simpson_diversity_stage_LM.svg", plot = simpson_stage_lm_plot, width = 8, height = 8)
```

```{r - final simpson diversity plot}
simpson_stage_lm_plot_r <- simpson_stage_lm_plot +
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  )

combined_simpson <- simpson_lm_plot + simpson_stage_lm_plot_r +
  plot_layout(ncol = 2, guides = "collect") &
  theme(
    axis.text.x = element_text(size = 17),
    axis.text.y = element_text(size = 18)
  )

combined_simpson

ggsave("../figures/simpson_diversity_combined_LM.svg", plot = combined_simpson,
       height = 6,
       width = 8.5,
       device = svglite::svglite)
```


## Rarefaction Plot

```{r - rarefraction plots }
library(phyloseq)

set.seed(1234)

ps.rarefied <- rarefy_even_depth(ps, rngseed = 1, sample.size = min(sample_sums(ps)),replace = F)
```

```{r}
library(vegan)
library(permute)

set.seed(1)
otu <- as(otu_table(ps), "matrix")
otu.rf <- rrarefy(otu, min(sample_sums(ps)))

# rarefaction curves
rarecurve(otu, step = 10000, label = F)

ps.rarefied <- ps
otu_table(ps.rarefied) <- otu_table(otu.rf, taxa_are_rows = taxa_are_rows(ps))
```


## Pielou's eveness
```{r}
# Pielou's evenness J = H / ln(S)
# H: Shannon diversity, S: number of observed taxa (richness)
otu_even <- as(otu_table(ps), "matrix")
if (taxa_are_rows(ps)) {
  otu_even <- t(otu_even)
}

pielou_df <- data.frame(
  sample = rownames(otu_even),
  richness = rowSums(otu_even > 0),
  stringsAsFactors = FALSE
) %>%
  mutate(
    pielou_evenness = ifelse(richness > 1, shannon / log(richness), NA_real_)
  )

# add metadata and keep order used in other category plots
pielou_df <- pielou_df %>%
  left_join(alpha_diversity %>% select(sample, category), by = "sample") %>%
  mutate(
    category = factor(category,
                      levels = c("Non_farmer_controls", "Pig_farmers", "Pigs"))
  )

pielou_comparisons <- list(
  c("Non_farmer_controls", "Pig_farmers"),
  c("Non_farmer_controls", "Pigs"),
  c("Pig_farmers", "Pigs")
)

pielou_plot <- ggplot(pielou_df,
                      aes(x = category, y = pielou_evenness, fill = category)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.35, size = 2, color = "black") +
  scale_fill_manual(values = sample_type_colors) +
  scale_x_discrete(labels = color_codes$label_mapping) +
  xlab("") +
  ylab("Pielou's Evenness (J)") +
  labs(title = "Pielou's Evenness per Category") +
  stat_compare_means(
    comparisons = pielou_comparisons,
    method = "wilcox.test",
    aes(label = after_stat(p.format)),
    step.increase = 0.08,
    size = 5,
    family = "DejaVu Sans"
  ) +
  theme_publication() +
  theme(legend.position = "none")

print(pielou_plot)

ggsave("../figures/pielou_evenness_per_category.svg",
       plot = pielou_plot,
       width = 7,
       height = 6)

```

## Summary

This analysis includes the alpha and beta diversity calculation for based on the phyloseq object created from MetaPhlAn output:

-   Richness calculation
-   calculation of alpha diversity dataframe including:
    -   shannon diversity
    -   classified reads
    -   unclassified reads
    -   total reads
    -   classified to total reads ratio
-   boxplot of classified vs unclassified reads per category
-   alpha diversity (without accounting for the read depth)
-   alpha diversity plots accouting for the rarefraction curve
-   rarefraction curve


