# Richness {#sec-richness}

## Dependencies

`utils.R` **-script**

```{r}
source("utils.R")
```

**Packages**

```{r}
#| echo: true
#| message: false
#| warning: false
#| results: hide


#devtools::install_github("gmteunisse/fantaxtic")
#devtools::install_github("gmteunisse/ggnested")
library(tidyverse)
library(BiocManager)
library(phyloseq)
library(ggnested)
library(vegan)
library(matrixStats)



# load theme
theme_publication <- function(base_size = 18, base_family = "DejaVu Sans") {
ggplot2::theme_bw(
base_size = base_size,
base_family = base_family
) +
ggplot2::theme(
plot.title = ggplot2::element_text(hjust = 0.5),
axis.title = ggplot2::element_text(size = ggplot2::rel(1.05)),
axis.text = ggplot2::element_text(size = ggplot2::rel(1.2)),
legend.title = ggplot2::element_text(size = ggplot2::rel(0.95)),
legend.text = ggplot2::element_text(size = ggplot2::rel(0.9)),
strip.text = ggplot2::element_text(size = ggplot2::rel(0.95))
)
}
```

## Data: Loading and Filtering

-   MetaPhlAn-Output estimated count data table

```{r}
# metaphlan raw read data (MF_SGB_abundance.txt)
metaphlanData <- read.table("../data/raw/MF_SGB_abundance.txt", header = TRUE, sep = "")
metaphlanData$clade_name <- as.character(metaphlanData$clade_name)
metaData <- read.table("../../metadata/metadata_silas.csv", header = TRUE, sep = ",") %>%
  column_to_rownames(var = "id")

# name the rows according to the sample names
rownames(metaData) <- metaData$Sample_name

# assigns 0 to all NA's in my metaphlan data table
metaphlanData[is.na(metaphlanData)] <- 0

```

-   filtering of all duplicated samples (ppB, HpB, H1B)

```{r}
# metaphlanData
metaphlanData <- metaphlanData[, !grepl("ppB", colnames(metaphlanData))]

```

-   load colors

```{r}
# Load color codes from YAML file
color_codes <- yaml::read_yaml("../../Metadata/color_codes.yaml")

# Extract specific color palettes
sample_type_colors <- unlist(color_codes$sample_types)
pig_stage_colors <- unlist(color_codes$pig_stages)
category2_colors <- unlist(color_codes$category2)
```

## Phyloseq-Object

```{r}
meta_physeq <- metaphlan_to_phyloseq(metaphlanData, metaData, version = 4, verbose = TRUE, tax_lvl = "SGB")

ps <- meta_physeq
ps <- subset_taxa(ps, Kingdom !="UNKNOWN") # remove the unknowns

#View(ps@tax_table@.Data)

ps@tax_table@.Data <- substring(ps@tax_table@.Data, 4)
taxa_names(ps) <- substring(taxa_names(ps), 4)
colnames(ps@tax_table@.Data)<- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "SGB")
```

### Phyloseq-Object: Filtering

```{r}
ps_filtered <- filter_species_by_group_prevalence(
  ps = ps,
  min_count = 10,
  min_prevalence = 0.1,
  group_var = "category"
)


ps <-  ps_filtered
```

## Richness Plots Theme for 1x1 plots

```{r}
theme_personal <- function(){
  list(
    theme_bw(base_family = "DejaVu Sans") +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5),
      scale_x_discrete(labels = color_codes$label_mapping)
    ),
    scale_color_manual(values = sample_type_colors),
    scale_x_discrete(labels = color_codes$label_mapping),
    xlab("")
  )
}

library(ggplot2)
theme_set(theme_bw(base_family = "DejaVu Sans"))
```

## Richness plots by category

```{r}
richness_plots <- plot_richness(ps, x = "category", color = "category",
              measures = c("Observed", "Shannon"))+
  theme_bw(base_family = "DejaVu Sans")+
  geom_boxplot()+
  scale_color_manual(values = sample_type_colors)+
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x=element_blank(),
    legend.position = "right"
  )+
  xlab("")+
  ggtitle("Richness and Alpha Diversity by Group")+
  scale_x_discrete(labels = color_codes$label_mapping)
  

print(richness_plots)

ggsave("../figures/richness_plots.svg", height = 6, width = 8)



```

### Add Statistics

```{r}
alpha_richness_df <- estimate_richness(ps, measures =  c("Observed", "Shannon"))

# add categorty
alpha_richness_df$category <- sample_data(ps)$category

# plot boxplots with statistics
library(ggpubr)

richness_plots + ggpubr::stat_compare_means(method = "wilcox.test")


# define comparisons
my_comparisons <- list(
  c("Non_farmer_controls", "Pig_farmers"),
  c("Non_farmer_controls", "Pigs"),
  c("Pig_farmers", "Pigs")
)


richness_shannon_plot_groups <- richness_plots + ggpubr::stat_compare_means(method = "wilcox.test",
                                            comparisons = my_comparisons,
                                            aes(label = after_stat(p.format)),
                                            label.y.npc = 0.95)


print(richness_shannon_plot_groups)

# save this plot
ggsave("../figures/richness_shannon_groups.svg", plot = richness_shannon_plot_groups)
```

### One-panel plots

#### Richness

```{r}
# Function to format p-values
clean_p <- function(x) {
  ifelse(
    x < 0.001,
    "<0.001",   
    sprintf("%.3f", x)           # e.g. 0.155
  )
}

# richness category
p_richness_group <- plot_richness(ps, x = "category", color = "category",
              measures = c("Observed"))+
  geom_boxplot(
    aes(fill = category),
    color = "black"
  )+
  geom_jitter(
    alpha = 0.3, width = 0.2, color = "black"
  )+
  facet_grid(. ~variable, labeller = as_labeller(c(Observed = "Richness per Group")))+
  theme_personal()+
  scale_fill_manual(values = sample_type_colors)


p_richness_group_df <- estimate_richness(ps, measures =  c("Observed"))

# add categorty
p_richness_group_df$category <- sample_data(ps)$category

# plot boxplots with statistics
library(ggpubr)

# define comparisons
my_comparisons_groups <- list(
  c("Non_farmer_controls", "Pig_farmers"),
  c("Non_farmer_controls", "Pigs"),
  c("Pig_farmers", "Pigs")
)

# Calculate p-values manually for each comparison
pval_results <- list()
for (i in 1:length(my_comparisons_groups)) {
  group1 <- my_comparisons_groups[[i]][1]
  group2 <- my_comparisons_groups[[i]][2]
  
  data1 <- p_richness_group_df$Observed[p_richness_group_df$category == group1]
  data2 <- p_richness_group_df$Observed[p_richness_group_df$category == group2]
  
  test <- wilcox.test(data1, data2)
  pval_results[[i]] <- data.frame(
    group1 = group1,
    group2 = group2,
    p.value = test$p.value
  )
}

pval_df_richness_group <- do.call(rbind, pval_results) %>%
  mutate(
    y.position = max(p_richness_group_df$Observed, na.rm = TRUE) * (1.05 + 0.10 * row_number()),
    p.label = clean_p(p.value)
  )

p_richness_group <- p_richness_group + ggpubr::stat_pvalue_manual(
    pval_df_richness_group,
    label = "p.label",
    xmin = "group1", xmax = "group2",
    y.position = "y.position",
    tip.length = 0.02,
    size = 6,
    family = "DejaVu Sans")+
  theme(plot.margin = margin(5.5,5.5,5.5,5.5))+
  ylab("Richness (species present per sample)")


# adjust the size
p_richness_group <- p_richness_group +
  theme_publication()+
  theme(
    legend.position = "none",
    strip.text = element_text(size = 15, face = "bold", family = "DejaVu Sans"),
    strip.background = element_rect(fill = "grey85", color = "black"),
    axis.text.x = element_text(size = 13)
  )


print(p_richness_group)

ggsave("../figures/richness_1x1_group.svg", plot = p_richness_group)
```

#### Shannon

```{r}
p_shannon_group <- plot_richness(ps, x = "category", color = "category",
              measures = c("Shannon"))+
  geom_boxplot()+
  facet_grid(. ~variable, labeller = as_labeller(c(Shannon = "Shannon Diversity Index per Group")))+
  theme_personal()


p_shannon_group_df <- estimate_richness(ps, measures =  c("Shannon"))

# add categorty
p_shannon_group_df$category <- sample_data(ps)$category

# plot boxplots with statistics
library(ggpubr)

# define comparisons
my_comparisons_groups <- list(
  c("Non_farmer_controls", "Pig_farmers"),
  c("Non_farmer_controls", "Pigs"),
  c("Pig_farmers", "Pigs")
)

# Calculate p-values manually for each comparison
pval_results_shannon <- list()
for (i in 1:length(my_comparisons_groups)) {
  group1 <- my_comparisons_groups[[i]][1]
  group2 <- my_comparisons_groups[[i]][2]
  
  data1 <- p_shannon_group_df$Shannon[p_shannon_group_df$category == group1]
  data2 <- p_shannon_group_df$Shannon[p_shannon_group_df$category == group2]
  
  test <- wilcox.test(data1, data2)
  pval_results_shannon[[i]] <- data.frame(
    group1 = group1,
    group2 = group2,
    p.value = test$p.value
  )
}

pval_df_shannon_group <- do.call(rbind, pval_results_shannon) %>%
  mutate(
    y.position = max(p_shannon_group_df$Shannon, na.rm = TRUE) * (1 + 0.1 * row_number()),
    p_label = clean_p(p.value)
  )

p_shannon_group <- p_shannon_group + ggpubr::stat_pvalue_manual(
    pval_df_shannon_group,
    label = "p_label",
    xmin = "group1", xmax = "group2",
    y.position = "y.position",
    tip.length = 0.02,
    size = 6,
    family = "DejaVu Sans")+
  theme(plot.margin = margin(5.5,5.5,5.5,5.5))+
  ylab("Shannon Diversity Index")+
  theme(
    strip.text = element_text(size = 14, face = "bold", family = "DejaVu Sans"),
    strip.background = element_rect(fill = "grey85", color = "black"),
    axis.text.x = element_text(size = 13)
  )+
  theme_publication()+
  theme(
    legend.position = "none"
  )


print(p_shannon_group)

ggsave("../figures/shannon_1x1_group.svg", plot = p_shannon_group)
```

## Richness plots by stage

```{r}
# plot boxplots with statistics
library(ggpubr)

# suset pigs from phyloseq object 
ps_pigs <- subset_samples(ps, species == "Pig")

# extract richness measures
alpha_richness_stage_df <- estimate_richness(ps_pigs, measures =  c("Observed", "Shannon"))

# add grwoth_stage
alpha_richness_stage_df$grwoth_stage <- sample_data(ps_pigs)$growth_stage

# define order (level)
alpha_richness_stage_df$grwoth_stage <- factor(alpha_richness_stage_df$grwoth_stage,
                                               levels = c("Suckling", "Weaning", "Fattening"))

# define order (level) in the phyloseq object
sample_data(ps_pigs)$growth_stage <- factor(
  sample_data(ps_pigs)$growth_stage, levels = c("Suckling", "Weaning", "Fattening"))



# define comparisons
my_comparisons_stage <- list(
  c("Fattening", "Suckling"),
  c("Fattening", "Weaning"),
  c("Suckling", "Weaning")
)


richness_plots_stage <- plot_richness(ps_pigs, x = "growth_stage", color = "growth_stage",
              measures = c("Observed"))+
  theme_bw(base_family = "DejaVu Sans")+
  geom_boxplot()+
  scale_color_manual(values = pig_stage_colors)+
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x=element_blank(),
    legend.position = "right"
  )+
  xlab("")+
  ggtitle("Richness by Growth Stage")+
  scale_x_discrete(labels = color_codes$label_mapping)

# Calculate p-values manually for richness stage plot
pval_results_stage_richness_multi <- list()
ps_pigs_richness <- estimate_richness(ps_pigs, measures = c("Observed"))
ps_pigs_richness$growth_stage <- sample_data(ps_pigs)$growth_stage

for (i in 1:length(my_comparisons_stage)) {
  group1 <- my_comparisons_stage[[i]][1]
  group2 <- my_comparisons_stage[[i]][2]
  
  data1 <- ps_pigs_richness$Observed[ps_pigs_richness$growth_stage == group1]
  data2 <- ps_pigs_richness$Observed[ps_pigs_richness$growth_stage == group2]
  
  test <- wilcox.test(data1, data2)
  pval_results_stage_richness_multi[[i]] <- data.frame(
    group1 = group1,
    group2 = group2,
    p.value = test$p.value
  )
}

pval_df_stage_richness_multi <- do.call(rbind, pval_results_stage_richness_multi) %>%
  mutate(
    y.position = max(ps_pigs_richness$Observed, na.rm = TRUE) * (1 + 0.1 * row_number()),
    p_label = clean_p(p.value)
  )

richness_plots_stage <- richness_plots_stage + ggpubr::stat_pvalue_manual(
    pval_df_stage_richness_multi,
    label = "p_label",
    xmin = "group1", xmax = "group2",
    y.position = "y.position",
    tip.length = 0.02,
    size = 5,
    family = "DejaVu Sans")

print(richness_plots_stage)

ggsave("../figures/richness_stage.svg", plot = richness_plots_stage)
```

### One-panel plots

#### Richness

```{r}
p_richness_stage <- plot_richness(ps_pigs, x = "growth_stage", color = "growth_stage",
              measures = c("Observed"))+
  geom_boxplot(aes(
    fill = growth_stage),
    color = "black",
    outlier.shape = NA)+
  geom_jitter(
    alpha = 0.3, width = 0.2
  )+
  facet_grid(. ~variable, labeller = as_labeller(c(Observed = "Richness per Pig Growth Stage")))+
  theme_personal()+
  scale_fill_manual(values = pig_stage_colors)
  


p_richness_stage_df <- estimate_richness(ps_pigs, measures =  c("Observed"))

# add growth stage
p_richness_stage_df$growth_stage <- sample_data(ps_pigs)$growth_stage

# plot boxplots with statistics
library(ggpubr)

# Calculate p-values manually for each comparison
pval_results_stage_richness <- list()
for (i in 1:length(my_comparisons_stage)) {
  group1 <- my_comparisons_stage[[i]][1]
  group2 <- my_comparisons_stage[[i]][2]
  
  data1 <- p_richness_stage_df$Observed[p_richness_stage_df$growth_stage == group1]
  data2 <- p_richness_stage_df$Observed[p_richness_stage_df$growth_stage == group2]
  
  test <- wilcox.test(data1, data2)
  pval_results_stage_richness[[i]] <- data.frame(
    group1 = group1,
    group2 = group2,
    p.value = test$p.value
  )
}

pval_df_richness_stage <- do.call(rbind, pval_results_stage_richness) %>%
  mutate(
    y.position = max(p_richness_stage_df$Observed, na.rm = TRUE) * (1.05 + 0.10 * row_number()),
    p.label = clean_p(p.value)
  )

p_richness_stage <- p_richness_stage + ggpubr::stat_pvalue_manual(
    pval_df_richness_stage,
    label = "p.label",
    xmin = "group1", xmax = "group2",
    y.position = "y.position",
    tip.length = 0.02,
    size = 6,
    family = "DejaVu Sans")+
  theme(
      axis.title.y = element_blank(),
      axis.text.y  = element_blank(),
      axis.ticks.y = element_blank(),
      plot.margin  = margin(5.5, 5.5, 5.5, 0)   # no left margin
    )

# make the header bigger
p_richness_stage <- p_richness_stage +
  theme_publication()+
  theme(
    strip.text = element_text(size = 15, face = "bold", family = "DejaVu Sans"),
    strip.background = element_rect(fill = "grey85", color = "black"),
    axis.text.x = element_text(size = 13),
    legend.position = "none"
  )

print(p_richness_stage)

# save the plot
ggsave("../figures/richness_1x1_stage.svg", plot = p_richness_stage)
```

#### Shannon

```{r}
p_shannon_stage <- plot_richness(ps_pigs, x = "growth_stage", color = "growth_stage",
              measures = c("Shannon"))+
  geom_boxplot()+
  facet_grid(. ~variable, labeller = as_labeller(c(Shannon = "Shannon Diversity Index per Pig Growth Stage")))+
  theme_personal()+
  scale_color_manual(values = pig_stage_colors)
  


p_shannon_stage_df <- estimate_richness(ps_pigs, measures =  c("Shannon"))

# add growth stage
p_shannon_stage_df$growth_stage <- sample_data(ps_pigs)$growth_stage

# plot boxplots with statistics
library(ggpubr)

# Calculate p-values manually for each comparison
pval_results_stage_shannon <- list()
for (i in 1:length(my_comparisons_stage)) {
  group1 <- my_comparisons_stage[[i]][1]
  group2 <- my_comparisons_stage[[i]][2]
  
  data1 <- p_shannon_stage_df$Shannon[p_shannon_stage_df$growth_stage == group1]
  data2 <- p_shannon_stage_df$Shannon[p_shannon_stage_df$growth_stage == group2]
  
  test <- wilcox.test(data1, data2)
  pval_results_stage_shannon[[i]] <- data.frame(
    group1 = group1,
    group2 = group2,
    p.value = test$p.value
  )
}

pval_df_shannon_stage <- do.call(rbind, pval_results_stage_shannon) %>%
  mutate(
    y.position = max(p_shannon_stage_df$Shannon, na.rm = TRUE) * (1 + 0.1 * row_number()),
    p_label = clean_p(p.value)
  )

p_shannon_stage <- p_shannon_stage + ggpubr::stat_pvalue_manual(
    pval_df_shannon_stage,
    label = "p_label",
    xmin = "group1", xmax = "group2",
    y.position = "y.position",
    tip.length = 0.02,
    size = 6,
    family = "DejaVu Sans")+
  theme(
      axis.title.y = element_blank(),
      axis.text.y  = element_blank(),
      axis.ticks.y = element_blank(),
      plot.margin  = margin(5.5, 5.5, 5.5, 0)   # no left margin
    )+
  ylab("Shannon Diversity Index")+
  theme(
    strip.text = element_text(size = 14, face = "bold", family = "DejaVu Sans"),
    strip.background = element_rect(fill = "grey85", color = "black"),
    axis.text.x = element_text(size = 13)
  )



print(p_shannon_stage)

# save this plot
ggsave("../figures/shannon_1x1_stage.svg", plot = p_shannon_stage)
```

## 4x4 Richness & Shannon plot

```{r}
library(patchwork)

(richness_shannon_plot_groups | richness_plots_stage)+
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

### Richness - 2x2

```{r}

# headroom for p-values
y_max <- max(
  p_richness_group$data$value,
  p_richness_stage$data$value,
  na.rm = TRUE
)

# keep enough headroom for all p-value brackets/labels in both panels
y_top_stats <- max(
  pval_df_richness_group$y.position,
  pval_df_richness_stage$y.position,
  na.rm = TRUE
)
ylims <- c(0, max(y_max * 1.15, y_top_stats * 1.05))


# hide the y-axis on the right plot
p_richness_stage_r <- p_richness_stage +
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  )

# adjust size of y-axis of left plot
p_richness_group <- p_richness_group +
  theme(
    axis.title.y = element_text(size = 18),  # y-label
    axis.text.y  = element_text(size = 17)   # y-axis numbers
  )

# combine the plots
library(patchwork)

layout_richness <-
  (p_richness_group | p_richness_stage_r) &
  coord_cartesian(ylim = ylims)

layout_richness

ggsave("../figures/richness_final_2x2.svg", plot = layout_richness, width = 8, height = 6)
```

### Shannon - 2x2

```{r}
# layout shannon

# common limits and grids
common_limits_shannon <- range(
  ggplot_build(p_shannon_group)$data[[1]]$y,
  ggplot_build(p_shannon_stage)$data[[1]]$y
)

extra <- 1.1 * diff(common_limits_shannon)
common_limits_shannon[1] <- common_limits_shannon[1] + extra

p_shannon_group  <- p_shannon_group  + scale_y_continuous(limits = common_limits_shannon)
p_shannon_stage  <- p_shannon_stage  + scale_y_continuous(limits = common_limits_shannon)


layout_shannon <- (p_shannon_group | p_shannon_stage) +
  plot_layout(
    guides = "collect",
    widths = c(1, 1),       
    design = "AB"           
  )

layout_shannon

ggsave("../figures/shannon_final_2x2.svg", plot = layout_shannon, width = 7, height = 6)
```

merge both together

```{r}
layout_final <- layout_richness / layout_shannon
layout_final

# save layout_final
ggsave("../figures/richness_shannon_final_4x4_panel.svg", plot = layout_final, width = 7, height = 10, dpi = 600)
```
