# Richness Filtering based on prevalence {#sec-richness-filtering}

## Dependencies

`utils.R` **-script**

```{r}
source("utils.R")
```

**Packages**

```{r}
#| echo: true
#| message: false
#| warning: false
#| results: hide


#devtools::install_github("gmteunisse/fantaxtic")
#devtools::install_github("gmteunisse/ggnested")
library(tidyverse)
library(BiocManager)
library(phyloseq)
library(ggnested)
library(vegan)
library(matrixStats) #for 'colAnys'-function in'filter_species_by_group_prevalence'-function
```

## Data: Loading and Filtering

-   metaphlan estimated count data table

```{r}
# metaphlan raw read data (MF_SGB_abundance.txt)
metaphlanData <- read.table("../data/raw/MF_SGB_abundance.txt", header = TRUE, sep = "")
metaphlanData$clade_name <- as.character(metaphlanData$clade_name)
metaData <- read.table("../metadata/metadata_samples.csv", header = TRUE, sep = ",")

# name the rows according to the sample names
rownames(metaData) <- metaData$Sample_name

# assigns 0 to all NA's in my metaphlan data table
metaphlanData[is.na(metaphlanData)] <- 0

```

-   filtering of all duplicated samples (ppB, HpB, H1B)

```{r}
# metaphlanData
metaphlanData <- metaphlanData[, !grepl("ppB|H1B|HpB", colnames(metaphlanData))]

```

## Phyloseq-Object

```{r}
meta_physeq <- metaphlan_to_phyloseq(metaphlanData, metaData, version = 4, verbose = TRUE, tax_lvl = "SGB")

ps <- meta_physeq
ps <- subset_taxa(ps, Kingdom !="UNKNOWN") # remove the unknowns

#View(ps@tax_table@.Data)

ps@tax_table@.Data <- substring(ps@tax_table@.Data, 4)
taxa_names(ps) <- substring(taxa_names(ps), 4)
colnames(ps@tax_table@.Data)<- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "SGB")
```

### Filter Phyloseq-Object based on different Cut-OFFs

prevalence cut offs of: 0, 5 and 10% Prevalence per Group

```{r}
# definition of minimal prevalence thresholds
min_prevalence_values <- c(0, 0.05, 0.1)

# list of filtered ps-objects
list_filtered_ps <- list()

# for-loop to filter based on min_prevalence_values

for (min_prev in min_prevalence_values){
  
  # Apply the filtering function
  ps_filtered <- filter_species_by_group_prevalence(
    ps = ps,
    min_count = 10,
    min_prevalence = min_prev,
    group_var = "animal"
  )
  
  # Store the result in a named list
  list_filtered_ps[[as.character(min_prev)]] <- ps_filtered
}


```

## Richness Plots based on different Cut-OFFs

```{r}
#| label: richness_plots_by_prevalence

# create an empty list for richness plots
richness_plots <- list()

# loop through each prevalence entery in list_filtered_ps
i <- 1

for (name in names(list_filtered_ps)){
  # assign the namne of list_filtered_ps to the variable ps_obj
  ps_obj <- list_filtered_ps[[name]]
  
  # plots
  richness_plots[[i]] <- plot_richness(ps_obj, x = "animal", color = "animal")+
  theme_minimal()+
  geom_boxplot()+
  labs(title = paste0("richnessplots: ", min_prev,"% prevalence per group"))+
  ggtitle(paste("Prevalence filter: ", name, "% per group"))
  
  i = i + 1
}

# Display all 3 plots next to each other
gridExtra::grid.arrange(grobs = richness_plots, nrow = 3)

```
