# NMDS - realtive abundance data {#sec-NMDS-rel-abundance-data}

## Dependencies

`utils.R` **-script**

```{r}
setwd("/Users/simeonstreit/Library/CloudStorage/OneDrive-UniversitaetBern/20_Pig_Project/Simeon")
source("datautils.R")

# Load color codes from YAML file
color_codes <- yaml::read_yaml("../../Metadata/color_codes.yaml")

# Extract specific color palettes
sample_type_colors <- unlist(color_codes$sample_types)
pig_stage_colors <- unlist(color_codes$pig_stages)
category2_colors <- unlist(color_codes$category2)


# set theme
# theme
theme_publication <- function(base_size = 14, base_family = "DejaVu Sans") {
ggplot2::theme_bw(
base_size = base_size,
base_family = base_family
) +
ggplot2::theme(
plot.title = ggplot2::element_text(hjust = 0.5),
axis.title = ggplot2::element_text(size = ggplot2::rel(1.05)),
axis.text = ggplot2::element_text(size = ggplot2::rel(0.95)),
legend.title = ggplot2::element_text(size = ggplot2::rel(0.95)),
legend.text = ggplot2::element_text(size = ggplot2::rel(0.9)),
strip.text = ggplot2::element_text(size = ggplot2::rel(0.95))
)
}
```

**Packages**

```{r}
#| echo: true
#| message: false
#| warning: false
#| results: hide


# devtools::install_github("gmteunisse/fantaxtic")
# devtools::install_github("gmteunisse/ggnested")
library(tidyverse)
library(BiocManager)
library(phyloseq)
library(ggnested)
library(vegan)
library(matrixStats) # for 'colAnys'-function in'filter_species_by_group_prevalence'-function.
library(microViz)
library(fantaxtic) # for making barplots
library(gplots) # for enhanced heatmaps


```

## Data: Loading and Filtering (relative-Abundance data)

```{r}
# load metaphlan percentage file (relative abundance file)
metaphlanData_percent <- read.table("../data/raw/MF_SGB_percent.txt", header = TRUE, sep = "", row.names = 1)
# convert rownames into first column
metaphlanData_percent <- rownames_to_column(metaphlanData_percent, var = "clade_name")
metaphlanData_percent$clade_name <- as.character(metaphlanData_percent$clade_name)

# load metadata file
metaData <- read.table("../../metadata/metadata_silas.csv", header = TRUE, sep = ",") %>%
    column_to_rownames(var = "id")

```

check metaphlan percentage file

```{r}
metaphlanData_percent_check <- metaphlanData_percent %>%
    select(-clade_name)

summary(colSums(metaphlanData_percent_check))
```

filter metaphlan data (all ppB) samples

```{r}
# metaphlanData percentage
metaphlanData_percent <- metaphlanData_percent[, !grepl("ppB", colnames(metaphlanData_percent))]

```

## Phyloseq-Object

```{r}
meta_physeq_percent <- metaphlan_to_phyloseq(metaphlanData_percent, metaData, version = 4, verbose = TRUE, tax_lvl = "SGB")
ps_percent <- meta_physeq_percent
ps_percent <- subset_taxa(ps_percent, Kingdom != "k__Eukaryota") # remove eukaryotes


# check the ps-percent - object
ps_kingdom <- tax_glom(ps_percent, taxrank = "Kingdom")
otu_mat <- as(otu_table(ps_kingdom), "matrix")
# print(otu_mat)


# look if we have species without t__
tax_names <- taxa_names(ps_percent)
not_t_taxa <- tax_names[!grepl("t__", tax_names)]
print(not_t_taxa)

```

### Filter Phyloseq-Object based on minimal relative abundance

retain taxa whose average relative abundance across all samples exceeds $10^5$.

```{r}
ps_percent_filt <- filter_taxa(
    ps_percent,
    function(x) mean(x) > 1e-5,
    prune = TRUE
)
```

## Ordination analysis using microViz

### PCA (CLR, SGB)
```{r}
# PCA with CLR transformation at SGB level
pca_ps <- ps_percent_filt %>%
    tax_transform("clr", rank = "Species") %>%
    ord_calc(method = "PCA")



```

```{r}
pca_plot <- ord_plot(pca_ps, color = "category2", size = 2, plot_taxa = 1:9) +
  stat_ellipse(aes(color = category2)) +
  #    ggtitle("PCA (CLR, SGB)") +
  labs(color = "Category") +
  scale_color_manual(
    values = unlist(color_codes$category2),
    labels = color_codes$label_mapping
  )
plot(pca_plot)
ggsave(
  "../figures/pca_clr_dim1_and_2.svg",
  plot = pca_plot,
  height = 6,
  width = 8
)
```

```{r}

pca_plot <- ord_plot(pca_ps, color = "category2", size = 2, axes = c(1, 3)) +
    stat_ellipse(aes(color = category2)) +
    ggtitle("PCA (CLR, SGB)") +
    labs(color = "Category") +
    scale_color_manual(values = unlist(color_codes$category2), labels = color_codes$label_mapping)
plot(pca_plot)
ggsave("../figures/pca_clr_dim1_and_3.svg", plot = pca_plot, height = 6, width = 8)

```



### PCoA (Bray-Curtis, SGB)
```{r}
# PCoA with Bray-Curtis at SGB level
pcoa_ps <- ps_percent_filt %>%
    tax_transform("identity", rank = "SGB") %>%
    dist_calc("bray") %>%
    ord_calc(method = "PCoA")

```


w30pp show microbiome similar to suckling pigs


```{r}
# Global PERMANOVA for category2 (same Bray-Curtis distance used in PCoA)
bc_dist_global <- phyloseq::distance(ps_percent_filt, method = "bray")
metadata_global <- data.frame(sample_data(ps_percent_filt))
metadata_global <- metadata_global[labels(bc_dist_global), , drop = FALSE]

permanova_global <- vegan::adonis2(
  bc_dist_global ~ category2,
  data = metadata_global,
  permutations = 999
)

format_p <- function(x) {
  ifelse(x < 0.001, "<0.001", sprintf("= %.3f", x))
}

permanova_label <- paste0(
  "PERMANOVA p ", format_p(permanova_global$`Pr(>F)`[1]),
  "\nR2 = ", round(permanova_global$R2[1], 3)
)

pcoa_plot <- ord_plot(pcoa_ps, color = "category2", size = 2, axes = c(1, 2)) +
    stat_ellipse(aes(group = category2, color = category2)) +
    #ggtitle("PCoA (Bray-Curtis, SGB)") +
    labs(color = "Category") +
    scale_color_manual(values = unlist(color_codes$category2), labels = color_codes$label_mapping)+
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = permanova_label,
    hjust = 1.1,
    vjust = 1.5,
    size = 4,
    family = "DejaVu Sans"
  ) +
  theme_bw(base_family = "DejaVu Sans")+
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 13),
    strip.text = element_text(size = 15)
  )


plt_data <- pcoa_plot$data


plot(pcoa_plot)
ggsave("../figures/pcoa_bray_sgb.svg", plot = pcoa_plot, height = 6, width = 8)
```


### PCoA (Bray-Curtis, humans vs pigs)

```{r}
# Build a binary grouping: Humans vs Pigs
ps_human_pig <- ps_percent_filt %>%
  ps_mutate(
    human_pig_group = case_when(
      species == "Human" ~ "Humans",
      species == "Pig" ~ "Pigs",
      TRUE ~ NA_character_
    )
  ) %>%
  ps_filter(!is.na(human_pig_group))

human_pig_pcoa_ps <- ps_human_pig %>%
  tax_transform("identity", rank = "SGB") %>%
  dist_calc("bray") %>%
  ord_calc(method = "PCoA")

# Global PERMANOVA for the binary human/pig grouping
human_pig_bc_dist <- phyloseq::distance(ps_human_pig, method = "bray")
human_pig_meta <- data.frame(sample_data(ps_human_pig))
human_pig_meta <- human_pig_meta[labels(human_pig_bc_dist), , drop = FALSE]

permanova_human_pig <- vegan::adonis2(
  human_pig_bc_dist ~ human_pig_group,
  data = human_pig_meta,
  permutations = 999
)

format_p_human_pig <- function(x) {
  ifelse(x < 0.001, "<0.001", sprintf("= %.3f", x))
}

permanova_human_pig_label <- paste0(
  "PERMANOVA p ", format_p_human_pig(permanova_human_pig$`Pr(>F)`[1]),
  "\nR2 = ", round(permanova_human_pig$R2[1], 3)
)

pcoa_human_pig_plot <- ord_plot(
  human_pig_pcoa_ps,
  color = "human_pig_group",
  size = 2,
  axes = c(1, 2)
) +
  stat_ellipse(aes(group = human_pig_group, color = human_pig_group)) +
  labs(color = "Group") +
  scale_color_manual(
    values = c(
      "Humans" = "#D4843D",
      "Pigs" = "#4C9F70"
    )
  ) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = permanova_human_pig_label,
    hjust = 1.1,
    vjust = 1.5,
    size = 4,
    family = "DejaVu Sans"
  ) +
  theme_bw(base_family = "DejaVu Sans") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 13),
    strip.text = element_text(size = 15)
  )

print(permanova_human_pig)
plot(pcoa_human_pig_plot)
ggsave("../figures/pcoa_bray_humans_vs_pigs.svg", plot = pcoa_human_pig_plot, height = 6, width = 8)
```


### PCoA (Bray-Curtis, pigs only, growth stages)

```{r}
# Pig-only subset for growth-stage-specific ordination
ps_pig <- ps_percent_filt %>%
  ps_filter(species == "Pig")

pig_pcoa_ps <- ps_pig %>%
  tax_transform("identity", rank = "SGB") %>%
  dist_calc("bray") %>%
  ord_calc(method = "PCoA")

# Global PERMANOVA for growth stage (permutations constrained within farm)
pig_bc_dist <- phyloseq::distance(ps_pig, method = "bray")
pig_meta <- data.frame(sample_data(ps_pig))
pig_meta <- pig_meta[labels(pig_bc_dist), , drop = FALSE]

permanova_growth_global <- vegan::adonis2(
  pig_bc_dist ~ growth_stage,
  data = pig_meta,
  permutations = 999,
  strata = pig_meta$farm
)

format_p_growth <- function(x) {
  ifelse(x < 0.001, "<0.001", sprintf("= %.3f", x))
}

permanova_growth_label <- paste0(
  "PERMANOVA p ", format_p_growth(permanova_growth_global$`Pr(>F)`[1]),
  "\nR2 = ", round(permanova_growth_global$R2[1], 3)
)

pcoa_growth_plot <- ord_plot(pig_pcoa_ps, color = "growth_stage", size = 2, axes = c(1, 2)) +
  stat_ellipse(aes(group = growth_stage, color = growth_stage)) +
  labs(color = "Growth stage") +
  scale_color_manual(values = unlist(color_codes$pig_stages)) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = permanova_growth_label,
    hjust = 1.1,
    vjust = 1.5,
    size = 4,
    family = "DejaVu Sans"
  ) +
  theme_bw(base_family = "DejaVu Sans") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 13),
    strip.text = element_text(size = 15)
  )

print(permanova_growth_global)
plot(pcoa_growth_plot)
ggsave("../figures/pcoa_bray_sgb_growth_stage_only.svg", plot = pcoa_growth_plot, height = 6, width = 8, web_fonts = "DejaVu Sans")
```



## PERMANOVA 

```{r}
# H0: Human vs Pig is the strongest factor
# This should have the highest R² value

permanova_h0_species <- ps_percent_filt %>%
    tax_transform("identity", rank = "SGB") %>%
    dist_calc("bray") %>%
    dist_permanova(
        variables = "species",
        n_perms = 999,
        seed = 42
    )

print("H0: Human vs Pig (strongest factor):")
print(permanova_h0_species)

```

human pig R2=0.20


```{r}
# H1 + H3 combined: Test growth_stage effect across ALL pigs
# This tests: are there differences among growth stages?
# Expected: suckling vs non-suckling >> weaning vs fattening

permanova_h1_h3_combined <- ps_percent_filt %>%
  ps_filter(species == "Pig") %>%
  tax_transform("identity", rank = "SGB") %>%
  dist_calc("bray") %>%
  dist_permanova(
    variables = c("growth_stage","farm"),
    n_perms = 999,
    seed = 42
  )

print("H1+H3: Growth stage effect in all pigs:")
print(permanova_h1_h3_combined)

# Pairwise PERMANOVA to see which groups differ
# This shows: suckling vs weaning, suckling vs fattening, weaning vs fattening
pig_bc_dist <- phyloseq::distance(
  ps_percent_filt %>% ps_filter(species == "Pig"), 
  method = "bray"
)

pig_metadata_full <- data.frame(sample_data(ps_percent_filt)) %>%
  filter(species == "Pig")

# Pairwise comparisons using pairwiseAdonis
library(pairwiseAdonis)
pairwise_growth <- pairwise.adonis2(
  pig_bc_dist ~ growth_stage + farm,
  data = pig_metadata_full,
  nperm = 999
)

print("Pairwise comparisons between growth stages:")
print(pairwise_growth)

```

All pig groups are different.

suckling pigs are more different than others. 

Suckling_vs_Weaning R2=0.38
Fattening_vs_Weaning R2= 0.14

Farm is not significant




```{r}



# H4: Farm is NOT important (accounting for pooling of samples)
# Note: If each sample is a pool of pigs from different growth stages on the same farm,
# we should test farm effect WITHIN each growth stage separately

# H4a: Farm effect across all pig samples
permanova_h4_overall <- ps_percent_filt %>%
  ps_filter(species == "Pig") %>%
  tax_transform("identity", rank = "SGB") %>%
  dist_calc("bray") %>%
  dist_permanova(
    variables = "farm",
    n_perms = 999,
    seed = 42
  )

print("H4a: Farm effect in pigs (overall):")
print(permanova_h4_overall)

# H4b: Test farm effect WITHIN growth stages (using strata)
# Permutations are restricted within growth stages
pig_bc_dist <- phyloseq::distance(
  ps_percent_filt %>% ps_filter(species == "Pig"), 
  method = "bray"
)

pig_metadata_full <- data.frame(sample_data(ps_percent_filt)) %>%
  filter(species == "Pig")

permanova_h4_stratified <- adonis2(
  pig_bc_dist ~ farm,
  data = pig_metadata_full,
  permutations = 999,
  strata = pig_metadata_full$growth_stage  # Block by growth stage
)

print("H4b: Farm effect blocking by growth_stage:")
print(permanova_h4_stratified)

# H4c: Sequential test - which is stronger: growth_stage or farm?
permanova_h4_sequential <- adonis2(
  pig_bc_dist ~ growth_stage + farm,
  data = pig_metadata_full,
  permutations = 999,
  by = "terms"  # Sequential: growth_stage first, then farm
)

print("H4c: Sequential PERMANOVA (growth_stage + farm):")
print(permanova_h4_sequential)

# H4d: Marginal test - independent effect of each after accounting for the other
permanova_h4_marginal <- adonis2(
  pig_bc_dist ~ growth_stage + farm,
  data = pig_metadata_full,
  permutations = 999,
  by = "margin"  # Marginal effects
)

print("H4d: Marginal PERMANOVA (growth_stage + farm):")
print(permanova_h4_marginal)

```



```{r}

# H5: Farmers vs Controls vs Pigs
# Interpretation:
# - This tests if the 4 categories differ in composition
# - High R² = strong differences among all groups
# - Post-hoc pairwise tests tell you WHICH groups differ
# - Expected: Pigs cluster separately from humans, farmers may be intermediate

permanova_h5 <- ps_percent_filt %>%
  #ps_filter(category2 %in% c("Pig_farmers", "Non_farmer_controls", "Older_pigs", "Weaning_pigs")) %>%
  tax_transform("identity", rank = "SGB") %>%
  dist_calc("bray") %>%
  dist_permanova(
    variables = "category2",
    n_perms = 999,
    seed = 42
  )

print("H5: Farmers vs Controls vs Pigs:")
print(permanova_h5)

# Pairwise comparisons for H5
#h5_subset <- ps_percent_filt %>%
  #ps_filter(category2 %in% c("Pig_farmers", "Non_farmer_controls", "Older_pigs", "Weaning_pigs"))

h5_bc_dist <- phyloseq::distance(ps_percent_filt, method = "bray")
h5_metadata <- data.frame(sample_data(ps_percent_filt))

library(pairwiseAdonis)
pairwise_h5 <- vegan::pairwise.adonis2(
  h5_bc_dist ~ category2,
  data = h5_metadata,
  nperm = 999
)

print("Pairwise comparisons for H5:")
print(pairwise_h5)


```





```{r}

metadata <- data.frame(sample_data(ps_percent_filt))
bc_dist <- phyloseq::distance(ps_percent_filt, method = "bray")

# Get sample names
farmer_samples <- rownames(metadata)[metadata$category2 == "Pig_farmers"]
control_samples <- rownames(metadata)[metadata$category2 == "Non_farmer_controls"]
older_pig_samples <- rownames(metadata)[metadata$category2 == "Older_pigs"]

bc_matrix <- as.matrix(bc_dist)

# Calculate mean distance from each farmer to all older pigs
farmer_mean_dists <- sapply(farmer_samples, function(farmer) {
  mean(bc_matrix[farmer, older_pig_samples])
})

# Calculate mean distance from each control to all older pigs
control_mean_dists <- sapply(control_samples, function(control) {
  mean(bc_matrix[control, older_pig_samples])
})

# Now you have ONE value per human sample (independent observations)
wilcox_corrected <- wilcox.test(farmer_mean_dists, control_mean_dists, 
                                alternative = "less")

cat("Mean distance per sample approach:\n")
cat("Farmers (n=", length(farmer_mean_dists), "):\n")
cat("  Mean:", mean(farmer_mean_dists), "\n")
cat("  Median:", median(farmer_mean_dists), "\n")

cat("\nControls (n=", length(control_mean_dists), "):\n")
cat("  Mean:", mean(control_mean_dists), "\n")
cat("  Median:", median(control_mean_dists), "\n")

cat("\nWilcoxon test (per-sample means):\n")
cat("  W:", wilcox_corrected$statistic, "\n")
cat("  P-value:", wilcox_corrected$p.value, "\n")

# Effect size
cohen_d <- (mean(farmer_mean_dists) - mean(control_mean_dists)) / 
           sqrt((sd(farmer_mean_dists)^2 + sd(control_mean_dists)^2) / 2)
cat("  Cohen's d:", cohen_d, "\n")

# Visualize
plot_data <- data.frame(
  distance = c(farmer_mean_dists, control_mean_dists),
  group = c(rep("Farmers", length(farmer_mean_dists)),
            rep("Controls", length(control_mean_dists)))
)

library(ggplot2)

# Format p-values for plot labels
format_p <- function(x) {
  ifelse(x < 0.001, "<0.001", sprintf("= %.3f", x))
}

stat_label <- paste0(
  "Wilcoxon p ", format_p(wilcox_corrected$p.value),
  "\nCohen's d = ", round(cohen_d, 2)
)

p <- ggplot(plot_data, aes(x = group, y = distance, fill = group)) +
  geom_boxplot(width = 0.2, alpha = 0.8) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 2) +
  theme_bw(base_size = 12, base_family = "DejaVu Sans") +
  labs(
    # title = "Mean Bray-Curtis Distance to Older Pigs",
    # subtitle = paste0("Each point = one human sample's mean distance to all Older_pig samples\n",
                     #"Wilcoxon p = ", format(wilcox_corrected$p.value, digits = 3),
                     #", Cohen's d = ", round(cohen_d, 3)),
    x = "",
    y = "Mean Bray-Curtis Distance to Older Pigs"
  ) +
  annotate(
    "text",
    x = 1.5,
    y = max(plot_data$distance, na.rm = TRUE) * 1.08,
    label = stat_label,
    family = "DejaVu Sans",
    size = 4
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.18))) +
  scale_fill_manual(values = c("Farmers" = "#D4843D", "Controls" = "#4A7C7E"))+
  theme_publication()+
  theme(legend.position = "none")

print(p)
ggsave("../figures/per_sample_mean_distance.svg", p, width = 5, height = 6)

```

```{r}

library(lme4)
library(lmerTest)

bc_dist <- phyloseq::distance(ps_percent_filt, method = "bray")

# Prepare data: each row is a distance between a human and a pig
human_samples <- c(farmer_samples, control_samples)
distance_data <- data.frame()

for (human in human_samples) {
  for (pig in older_pig_samples) {
    distance_data <- rbind(distance_data, data.frame(
      human_id = human,
      pig_id = pig,
      distance = bc_matrix[human, pig],
      group = metadata[human, "category2"]
    ))
  }
}

# Mixed model: group as fixed effect, human and pig as random effects
model <- lmer(distance ~ group + (1|human_id) + (1|pig_id), 
              data = distance_data)

summary(model)
anova(model)

```


# Within group and between group

```{r}

# Load required libraries
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(tidyr)

# Assuming your phyloseq object is called 'ps'
# Extract metadata
metadata <- data.frame(sample_data(ps_percent_filt))

# Calculate Bray-Curtis distance matrix
bc_dist <- phyloseq::distance(ps_percent_filt, method = "bray")
bc_matrix <- as.matrix(bc_dist)

# Create a function to extract distances by category pairs
extract_distances <- function(dist_matrix, metadata, category_col = "category2") {
  
  distances_df <- data.frame()
  
  samples <- rownames(dist_matrix)
  categories <- metadata[[category_col]][match(samples, rownames(metadata))]
  
  # Loop through all pairs of samples
  for (i in 1:(length(samples)-1)) {
    for (j in (i+1):length(samples)) {
      distances_df <- rbind(distances_df, data.frame(
        sample1 = samples[i],
        sample2 = samples[j],
        category1 = categories[i],
        category2 = categories[j],
        distance = dist_matrix[i, j],
        comparison_type = ifelse(categories[i] == categories[j], "Within", "Between"),
        category_pair = ifelse(categories[i] == categories[j], 
                               categories[i],
                               paste(sort(c(categories[i], categories[j])), collapse = " vs "))
      ))
    }
  }
  
  return(distances_df)
}

# Extract all pairwise distances
distances_df <- extract_distances(bc_matrix, metadata)
```

```{r}
# Summarize within-category distances
within_distances <- distances_df %>%
  filter(comparison_type == "Within") %>%
  group_by(category_pair) %>%
  summarise(
    mean_distance = mean(distance),
    median_distance = median(distance),
    sd_distance = sd(distance),
    n_comparisons = n()
  )

print("Within-category distances:")
print(within_distances)

# Summarize between-category distances
between_distances <- distances_df %>%
  filter(comparison_type == "Between") %>%
  group_by(category_pair) %>%
  summarise(
    mean_distance = mean(distance),
    median_distance = median(distance),
    sd_distance = sd(distance),
    n_comparisons = n()
  )

print("\nBetween-category distances:")
print(between_distances)

```

```{r}
# Extract specific comparisons for hypothesis testing
# Farmers to Older_pigs
farmers_to_older <- distances_df %>%
  filter((category1 == "Pig_farmers" & category2 == "Older_pigs") |
         (category1 == "Older_pigs" & category2 == "Pig_farmers")) %>%
  pull(distance)

# Controls to Older_pigs
controls_to_older <- distances_df %>%
  filter((category1 == "Non_farmer_controls" & category2 == "Older_pigs") |
         (category1 == "Older_pigs" & category2 == "Non_farmer_controls")) %>%
  pull(distance)

# Statistical test: Wilcoxon rank-sum test (Mann-Whitney U)
# Testing if farmers are closer (smaller distances) to Older_pigs than controls
wilcox_test <- wilcox.test(farmers_to_older, controls_to_older, 
                           alternative = "less")  # one-sided: farmers < controls

print("\n=== Statistical Test Results ===")
print("H0: Farmers and Controls are equally distant from Older_pigs")
print("H1: Farmers are closer to Older_pigs than Controls (distances are smaller)")
cat("\nFarmers to Older_pigs:")
cat("\n  Mean distance:", mean(farmers_to_older))
cat("\n  Median distance:", median(farmers_to_older))
cat("\n  N comparisons:", length(farmers_to_older))

cat("\n\nControls to Older_pigs:")
cat("\n  Mean distance:", mean(controls_to_older))
cat("\n  Median distance:", median(controls_to_older))
cat("\n  N comparisons:", length(controls_to_older))

cat("\n\nWilcoxon rank-sum test:")
cat("\n  W statistic:", wilcox_test$statistic)
cat("\n  P-value:", wilcox_test$p.value)
cat("\n  Significant at α=0.05:", wilcox_test$p.value < 0.05)

# Alternative: permutation test for more robust results
set.seed(123)
n_perm <- 9999

observed_diff <- mean(farmers_to_older) - mean(controls_to_older)
combined <- c(farmers_to_older, controls_to_older)
n_farmers <- length(farmers_to_older)
n_total <- length(combined)

perm_diffs <- replicate(n_perm, {
  shuffled <- sample(combined)
  mean(shuffled[1:n_farmers]) - mean(shuffled[(n_farmers+1):n_total])
})

perm_pvalue <- mean(perm_diffs <= observed_diff)

cat("\n\nPermutation test (", n_perm, " permutations):")
cat("\n  Observed difference:", observed_diff)
cat("\n  P-value:", perm_pvalue)
cat("\n  Significant at α=0.05:", perm_pvalue < 0.05)

# helper for readable p-values in plot subtitles
fmt_p <- function(x) {
  ifelse(x < 0.001, "<0.001", sprintf("%.3f", x))
}

# Visualization 1: Boxplot comparison
plot_data <- data.frame(
  distance = c(farmers_to_older, controls_to_older),
  group = c(rep("Farmers → Older_pigs", length(farmers_to_older)),
            rep("Controls → Older_pigs", length(controls_to_older)))
)

p1 <- ggplot(plot_data, aes(x = group, y = distance, fill = group)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 2) +
  theme_bw(base_family = "DejaVu Sans") +
  labs(
    title = "Bray-Curtis Distance to Older Pigs",
    subtitle = paste0(
      "Wilcoxon p = ", fmt_p(wilcox_test$p.value),
      " | Permutation p = ", fmt_p(perm_pvalue),
      " | Δmean = ", sprintf("%.3f", observed_diff)
    ),
    x = "",
    y = "Bray-Curtis Distance"
  ) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

print(p1)
ggsave("../figures/farmers_vs_controls_to_older_pigs.svg", p1, width = 6, height = 5)

# Visualization 2: All between-category comparisons
between_plot_data <- distances_df %>%
  filter(comparison_type == "Between") %>%
  mutate(category_pair = factor(category_pair))

# Omnibus test across all between-category pairs
kw_between <- kruskal.test(distance ~ category_pair, data = between_plot_data)

p2 <- ggplot(between_plot_data, aes(x = category_pair, y = distance)) +
  geom_boxplot(alpha = 0.7, fill = "steelblue") +
  theme_bw(base_family = "DejaVu Sans") +
  labs(
    title = "Between-Category Bray-Curtis Distances",
    subtitle = paste0("Kruskal-Wallis p = ", fmt_p(kw_between$p.value)),
    x = "Category Comparison",
    y = "Bray-Curtis Distance"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p2)
ggsave("../figures/between_category_distances.svg", p2, width = 10, height = 5)

# Visualization 3: Within vs Between comparison
all_distances_plot <- distances_df %>%
  mutate(category_pair = factor(category_pair))

# Global within-vs-between test
wilcox_within_between <- wilcox.test(
  distance ~ comparison_type,
  data = all_distances_plot,
  alternative = "two.sided"
)

p3 <- ggplot(all_distances_plot, aes(x = category_pair, y = distance, fill = comparison_type)) +
  geom_boxplot(alpha = 0.7) +
  theme_bw(base_family = "DejaVu Sans") +
  labs(
    title = "Within- and Between-Category Bray-Curtis Distances",
    subtitle = paste0("Within vs Between Wilcoxon p = ", fmt_p(wilcox_within_between$p.value)),
    x = "Category",
    y = "Bray-Curtis Distance",
    fill = "Comparison Type"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Within" = "#2E7D32", "Between" = "#1976D2"))

print(p3)
ggsave("../figures/within_vs_between_distances.svg", p3, width = 12, height = 6)


```



## Farm-Based Distance Analysis

```{r}
# Function to extract distances by farm pairs
extract_farm_distances <- function(dist_matrix, metadata) {
  
  distances_df <- data.frame()
  
  samples <- rownames(dist_matrix)
  farms <- metadata$farm[match(samples, rownames(metadata))]
  
  # Loop through all pairs of samples
  for (i in 1:(length(samples)-1)) {
    for (j in (i+1):length(samples)) {
      distances_df <- rbind(distances_df, data.frame(
        sample1 = samples[i],
        sample2 = samples[j],
        farm1 = farms[i],
        farm2 = farms[j],
        distance = dist_matrix[i, j],
        comparison_type = ifelse(farms[i] == farms[j], "Same Farm", "Different Farms"),
        farm_pair = ifelse(farms[i] == farms[j], 
                           as.character(farms[i]),
                           paste(sort(c(as.character(farms[i]), as.character(farms[j]))), collapse = " vs "))
      ))
    }
  }
  
  return(distances_df)
}


ps_percent_farm <- ps_percent_filt %>%
  ps_filter(category2 != "Non_farmer_controls")  # Focus on pig samples for farm analysis

metadata_filt <- data.frame(sample_data(ps_percent_farm))
rownames(metadata_filt) <- sample_names(ps_percent_farm)
bc_dist_filt <- phyloseq::distance(ps_percent_farm, method = "bray")
bc_matrix_filt <- as.matrix(bc_dist_filt)

# Extract farm-based distances (using filtered phyloseq without controls/suckling)
farm_distances_df <- extract_farm_distances(bc_matrix_filt, metadata_filt)

# Summarize same-farm distances
same_farm_distances <- farm_distances_df %>%
  filter(comparison_type == "Same Farm") %>%
  group_by(farm_pair) %>%
  summarise(
    mean_distance = mean(distance),
    median_distance = median(distance),
    sd_distance = sd(distance),
    n_comparisons = n()
  )

print("Same-farm distances:")
print(same_farm_distances)

# Summarize different-farm distances
different_farm_distances <- farm_distances_df %>%
  filter(comparison_type == "Different Farms") %>%
  summarise(
    mean_distance = mean(distance),
    median_distance = median(distance),
    sd_distance = sd(distance),
    n_comparisons = n()
  )

print("\nDifferent-farm distances:")
print(different_farm_distances)

# Statistical test: Same farm vs Different farms
same_farm_vals <- farm_distances_df %>%
  filter(comparison_type == "Same Farm") %>%
  pull(distance)

diff_farm_vals <- farm_distances_df %>%
  filter(comparison_type == "Different Farms") %>%
  pull(distance)

wilcox_farm_test <- wilcox.test(same_farm_vals, diff_farm_vals, 
                                 alternative = "less")

cat("\n=== Farm-Based Distance Test ===")
cat("\nH0: Same-farm and different-farm distances are equal")
cat("\nH1: Same-farm distances are smaller than different-farm distances")
cat("\n\nSame farm:")
cat("\n  Mean distance:", mean(same_farm_vals))
cat("\n  Median distance:", median(same_farm_vals))
cat("\n  N comparisons:", length(same_farm_vals))

cat("\n\nDifferent farms:")
cat("\n  Mean distance:", mean(diff_farm_vals))
cat("\n  Median distance:", median(diff_farm_vals))
cat("\n  N comparisons:", length(diff_farm_vals))

cat("\n\nWilcoxon rank-sum test:")
cat("\n  W statistic:", wilcox_farm_test$statistic)
cat("\n  P-value:", wilcox_farm_test$p.value)
cat("\n  Significant at α=0.05:", wilcox_farm_test$p.value < 0.05)

# Visualization: Same farm vs Different farms
farm_plot_data <- data.frame(
  distance = c(same_farm_vals, diff_farm_vals),
  comparison = c(rep("Same Farm", length(same_farm_vals)),
                 rep("Different Farms", length(diff_farm_vals)))
)

p_farm <- ggplot(farm_plot_data, aes(x = comparison, y = distance, fill = comparison)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  theme_bw(base_family = "DejaVu Sans") +
  labs(
    title = "Bray-Curtis Distances: Same vs Different Farms",
    subtitle = paste0("Wilcoxon test p-value: ", format(wilcox_farm_test$p.value, digits = 3)),
    x = "",
    y = "Bray-Curtis Distance"
  ) +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("Same Farm" = "#4CAF50", "Different Farms" = "#FF9800"))

print(p_farm)
ggsave("../figures/same_vs_different_farm_distances.pdf", p_farm, width = 6, height = 5)

# Enhanced analysis: Species type + Farm comparison
# Add species information to farm distances
farm_distances_df_species <- farm_distances_df %>%
  mutate(
    species1 = metadata_filt$species[match(sample1, rownames(metadata_filt))],
    species2 = metadata_filt$species[match(sample2, rownames(metadata_filt))],
    species_pair = case_when(
      species1 == "Human" & species2 == "Human" ~ "Human-Human",
      species1 == "Pig" & species2 == "Pig" ~ "Pig-Pig",
      TRUE ~ "Human-Pig"
    ),
    comparison_detailed = paste(species_pair, comparison_type, sep = " | ")
  )

# Summarize by species pair and farm status
species_farm_summary <- farm_distances_df_species %>%
  group_by(species_pair, comparison_type) %>%
  summarise(
    mean_distance = mean(distance),
    median_distance = median(distance),
    sd_distance = sd(distance),
    n_comparisons = n(),
    .groups = "drop"
  )

print("\nSpecies-Farm Distance Summary:")
print(species_farm_summary)

# Statistical tests for each species pair
# Human-Human same vs different farm
hh_same <- farm_distances_df_species %>%
  filter(species_pair == "Human-Human", comparison_type == "Same Farm") %>%
  pull(distance)
hh_diff <- farm_distances_df_species %>%
  filter(species_pair == "Human-Human", comparison_type == "Different Farms") %>%
  pull(distance)

# Pig-Pig same vs different farm
pp_same <- farm_distances_df_species %>%
  filter(species_pair == "Pig-Pig", comparison_type == "Same Farm") %>%
  pull(distance)
pp_diff <- farm_distances_df_species %>%
  filter(species_pair == "Pig-Pig", comparison_type == "Different Farms") %>%
  pull(distance)

# Human-Pig same vs different farm
hp_same <- farm_distances_df_species %>%
  filter(species_pair == "Human-Pig", comparison_type == "Same Farm") %>%
  pull(distance)
hp_diff <- farm_distances_df_species %>%
  filter(species_pair == "Human-Pig", comparison_type == "Different Farms") %>%
  pull(distance)

cat("\n=== Species-Farm Distance Tests ===\n")

if (length(hh_same) > 0 & length(hh_diff) > 0) {
  wilcox_hh <- wilcox.test(hh_same, hh_diff, alternative = "less")
  cat("\nHuman-Human:")
  cat("\n  Same farm (n=", length(hh_same), "): mean=", mean(hh_same), ", median=", median(hh_same))
  cat("\n  Diff farm (n=", length(hh_diff), "): mean=", mean(hh_diff), ", median=", median(hh_diff))
  cat("\n  Wilcoxon p-value:", wilcox_hh$p.value)
}

if (length(pp_same) > 0 & length(pp_diff) > 0) {
  wilcox_pp <- wilcox.test(pp_same, pp_diff, alternative = "less")
  cat("\n\nPig-Pig:")
  cat("\n  Same farm (n=", length(pp_same), "): mean=", mean(pp_same), ", median=", median(pp_same))
  cat("\n  Diff farm (n=", length(pp_diff), "): mean=", mean(pp_diff), ", median=", median(pp_diff))
  cat("\n  Wilcoxon p-value:", wilcox_pp$p.value)
}

if (length(hp_same) > 0 & length(hp_diff) > 0) {
  wilcox_hp <- wilcox.test(hp_same, hp_diff, alternative = "less")
  cat("\n\nHuman-Pig:")
  cat("\n  Same farm (n=", length(hp_same), "): mean=", mean(hp_same), ", median=", median(hp_same))
  cat("\n  Diff farm (n=", length(hp_diff), "): mean=", mean(hp_diff), ", median=", median(hp_diff))
  cat("\n  Wilcoxon p-value:", wilcox_hp$p.value)
}

# Visualization: Species-specific farm comparisons
p_species_farm <- ggplot(farm_distances_df_species, 
                         aes(x = species_pair, y = distance, fill = comparison_type)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  theme_bw(base_family = "DejaVu Sans") +
  labs(
    title = "Bray-Curtis Distances by Species Pair and Farm Status",
    subtitle = "Comparing same vs different farm within species pairs",
    x = "Species Comparison",
    y = "Bray-Curtis Distance",
    fill = "Farm Status"
  ) +
  scale_fill_manual(values = c("Same Farm" = "#4CAF50", "Different Farms" = "#FF9800"))

print(p_species_farm)
ggsave("../figures/species_farm_distances.svg", p_species_farm, width = 8, height = 6)

# Alternative visualization: Faceted by species pair
p_species_farm_facet <- ggplot(farm_distances_df_species, 
                                aes(x = comparison_type, y = distance, fill = comparison_type)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  facet_wrap(~species_pair, scales = "free_y") +
  theme_bw(base_family = "DejaVu Sans") +
  labs(
    title = "Bray-Curtis Distances: Same vs Different Farm by Species Pair",
    x = "",
    y = "Bray-Curtis Distance"
  ) +
  theme(legend.position = "none",
        #axis.text.x = element_text(angle = 45, hjust = 1)
        ) +
  scale_fill_manual(values = c("Same Farm" = "#4CAF50", "Different Farms" = "#FF9800"))

print(p_species_farm_facet)
ggsave("../figures/species_farm_distances_facet.svg", p_species_farm_facet, width = 7, height = 4, web_font = "DejaVu Sans")
```




 
#### Taxa associations with PCoA axes (Hellinger+envfit & Spearman correlations)


``` 


# takes very long


# Extract ordination object and sample scores
ord_obj <- ord_get(pcoa_ps)
sample_scores <- scores(ord_obj, display = "sites") %>% as.data.frame()

# Ensure we have at least two axes
n_axes <- min(ncol(sample_scores), 3)
print(paste("Number of PCoA axes available:", ncol(sample_scores)))

# Convert otu_table to sample x taxa matrix
otu_mat <- as(otu_table(ps_percent_filt), "matrix")
if (taxa_are_rows(ps_percent_filt)) otu_mat <- t(otu_mat)

# Hellinger-transform taxa (recommended before envfit/correlation)
X_hell <- decostand(otu_mat, method = "hellinger")

## envfit-style taxa fitting using vegan
fit_taxa <- vegan::envfit(as.matrix(sample_scores[, 1:n_axes]), X_hell, permutations = 999)

# Extract vectors and statistics
vecs <- as.data.frame(fit_taxa$vectors$arrows)
colnames(vecs) <- paste0("Axis", 1:ncol(vecs))
vecs$Taxa <- rownames(fit_taxa$vectors$arrows)
vecs$r <- fit_taxa$vectors$r
vecs$pval <- fit_taxa$vectors$pvals

# Filter significant taxa (p < 0.05)
sig_vecs <- vecs %>% filter(pval < 0.05) %>% arrange(desc(abs(r)))
print("Top significant taxa from envfit (r, p):")
print(head(sig_vecs, 20))

# Prepare to plot arrows on the PCoA
if (nrow(sig_vecs) > 0) {
    if (n_axes >= 2) {
        # Determine scale to make arrows visible relative to site scores
        site_range <- apply(sample_scores[, 1:2, drop = FALSE], 2, function(v) diff(range(v, na.rm = TRUE)))
        vec_range <- apply(sig_vecs[, 1:2, drop = FALSE], 2, function(v) diff(range(v, na.rm = TRUE)))
        scale_factor <- min(site_range / (vec_range + 1e-8), na.rm = TRUE) * 0.5
        sig_vecs$x_end <- sig_vecs$Axis1 * scale_factor
        sig_vecs$y_end <- sig_vecs$Axis2 * scale_factor
    } else {
        # single-axis ordination: map Axis1 into X-axis
        site_range <- diff(range(sample_scores[, 1], na.rm = TRUE))
        vec_range <- diff(range(sig_vecs$Axis1, na.rm = TRUE))
        scale_factor <- (site_range / (vec_range + 1e-8)) * 0.5
        sig_vecs$x_end <- sig_vecs$Axis1 * scale_factor
        sig_vecs$y_end <- 0
    }

    # Add arrows and labels to the PCoA plot
    pcoa_envfit_plot <- pcoa_plot +
        geom_segment(data = sig_vecs, aes(x = 0, y = 0, xend = x_end, yend = y_end),
                                 arrow = arrow(length = unit(0.2, "cm")), colour = "black") +
        geom_text(data = sig_vecs, aes(x = x_end, y = y_end, label = Taxa), size = 2, hjust = 0.5, vjust = -0.5)

    plot(pcoa_envfit_plot)
    ggsave("../figures/pcoa_bray_sgb_envfit.svg", plot = pcoa_envfit_plot, height = 6, width = 8)
}

## Spearman correlation approach (Hellinger taxa vs axes)
cor_results <- data.frame(Taxa = colnames(X_hell))
cor_results$Cor_Axis1 <- apply(X_hell, 2, function(z) cor(z, sample_scores[, 1], method = "spearman", use = "complete.obs"))
cor_results$Pval_Axis1 <- apply(X_hell, 2, function(z) cor.test(z, sample_scores[, 1], method = "spearman", exact = FALSE)$p.value)
cor_results$Padj_Axis1 <- p.adjust(cor_results$Pval_Axis1, method = "fdr")

cor_results$Cor_Axis2 <- apply(X_hell, 2, function(z) cor(z, sample_scores[, 2], method = "spearman", use = "complete.obs"))
cor_results$Pval_Axis2 <- apply(X_hell, 2, function(z) cor.test(z, sample_scores[, 2], method = "spearman", exact = FALSE)$p.value)
cor_results$Padj_Axis2 <- p.adjust(cor_results$Pval_Axis2, method = "fdr")

if (n_axes >= 3) {
  cor_results$Cor_Axis3 <- apply(X_hell, 2, function(z) cor(z, sample_scores[, 3], method = "spearman", use = "complete.obs"))
  cor_results$Pval_Axis3 <- apply(X_hell, 2, function(z) cor.test(z, sample_scores[, 3], method = "spearman", exact = FALSE)$p.value)
  cor_results$Padj_Axis3 <- p.adjust(cor_results$Pval_Axis3, method = "fdr")
}

# Top taxa for axis 1/2/3 by abs(correlation), FDR < 0.05
top_cor_axis1 <- cor_results %>% filter(Padj_Axis1 < 0.05) %>% arrange(desc(abs(Cor_Axis1)))
top_cor_axis2 <- cor_results %>% filter(Padj_Axis2 < 0.05) %>% arrange(desc(abs(Cor_Axis2)))
if (n_axes >= 3) top_cor_axis3 <- cor_results %>% filter(Padj_Axis3 < 0.05) %>% arrange(desc(abs(Cor_Axis3)))

print("Top taxa (Spearman) Axis 1:")
print(head(top_cor_axis1, 10))
print("Top taxa (Spearman) Axis 2:")
print(head(top_cor_axis2, 10))
if (n_axes >= 3) print("Top taxa (Spearman) Axis 3:")
if (n_axes >= 3) print(head(top_cor_axis3, 10))

# Save the correlation results
write.csv(cor_results, "../data/processed/pcoa_taxa_correlations_spearman_hellinger.csv", row.names = FALSE)

## Biplot showing top taxa (union of envfit and correlation results)
top_taxa <- unique(c(
    if (exists("sig_vecs") && nrow(sig_vecs) > 0) head(sig_vecs$Taxa, 10) else NULL,
    if (exists("top_cor_axis1") && nrow(top_cor_axis1) > 0) head(top_cor_axis1$Taxa, 10) else NULL,
    if (exists("top_cor_axis2") && nrow(top_cor_axis2) > 0) head(top_cor_axis2$Taxa, 10) else NULL
))
if (length(top_taxa) > 0) {
    # Convert taxa names to indices for plot_taxa (robust to microViz expectations)
    top_taxa_idx <- which(taxa_names(ps_percent_filt) %in% top_taxa)
    # Use microViz plotting to overlay top taxa labels/arrows
    pcoa_taxa_plot <- ord_plot(pcoa_ps, color = "category2", size = 2, plot_taxa = top_taxa_idx) +
    stat_ellipse(aes(group = category2, color = category2)) +
    ggtitle("PCoA (Bray-Curtis, SGB): top taxa overlay") +
    labs(color = "Category") +
    scale_color_manual(values = unlist(color_codes$category2), labels = color_codes$label_mapping)
    plot(pcoa_taxa_plot)
  ggsave("../figures/pcoa_bray_sgb_top_taxa.svg", plot = pcoa_taxa_plot, height = 6, width = 8)
}
```
```{r}
# load  cor_results

cor_results <- read.csv("../data/processed/pcoa_taxa_correlations_spearman_hellinger.csv")


```

```{r}

cor_results[cor_results$Padj_Axis1<0.05,] |> arrange(desc(abs(Cor_Axis1))) |> head(50) |> left_join(tax_tibble(pcoa_ps), by = c("Taxa"="FeatureID")) |> select(Taxa,Phylum, Species,Cor_Axis1,Cor_Axis2)




```
```{r}

tax_of_interest="t__SGB4837" # "g__Blautia" #"t__SGB4820" #"t__SGB4837"

plot_data <- pcoa_ps %>%
  #tax_agg(rank = "Genus") %>%
  tax_transform("log2", zero_replace = "halfmin", chain = TRUE) %>%
  ps_get() %>%
  ps_otu2samdat(tax_of_interest) %>% # adds tax as sample data!
  samdat_tbl()

ggplot(plot_data, aes(x = category2, y = .data[[tax_of_interest]],color=category2)) +
  geom_boxplot(width = 0.5) +
  #geom_jitter(width = 0.2, alpha = 0.5) +
  scale_color_manual(values = unlist(color_codes$category2), labels = color_codes$label_mapping)+
  labs(y="Log2 abundace", title=tax_table(pcoa_ps)[tax_of_interest,"Species"])+
  theme_bw(base_family = "DejaVu Sans")

ggsave("../figures/abundance_boxplot_"+tax_of_interest+".svg", height = 6, width = 6)

```

```{r}

cor_results[cor_results$Padj_Axis2<0.05,] |> arrange(desc(abs(Cor_Axis2))) |> head(50) |> left_join(tax_tibble(pcoa_ps), by = c("Taxa"="FeatureID")) |> select(Phylum, Species,Cor_Axis1,Cor_Axis2)




```

### PCoA Jaccard (presence absence based)



```{r}

jaccard_ps <- ps_percent_filt %>%
    tax_transform("identity", rank = "SGB") %>%
    dist_calc("jaccard") %>%
    ord_calc(method = "PCoA")


```

```{r}
jaccard_plot <- ord_plot(jaccard_ps, color = "category2", size = 2) +
    stat_ellipse(aes(group = category2, color = category2)) +
    ggtitle("PCoA (Jaccard, SGB)") +
    labs(color = "Category") +
    scale_color_manual(values = unlist(color_codes$category2), labels = color_codes$label_mapping)

plot(jaccard_plot)
ggsave("../figures/pcoa_jaccard_sgb.svg", plot = jaccard_plot, height = 6, width = 8)
```






