# NMDS - realtive abundance data {#sec-NMDS-rel-abundance-data}

## Dependencies

`utils.R` **-script**

```{r}
#setwd("/Users/simeonstreit/Library/CloudStorage/OneDrive-UniversitaetBern/20_Pig_Project/Simeon")
source("code/utils.R")

# Load color codes from YAML file
color_codes <- yaml::read_yaml("../Metadata/color_codes.yaml")

# Extract specific color palettes
sample_type_colors <- unlist(color_codes$sample_types)
pig_stage_colors <- unlist(color_codes$pig_stages)
category2_colors <- unlist(color_codes$category2)
```

**Packages**

```{r}
#| echo: true
#| message: false
#| warning: false
#| results: hide


# devtools::install_github("gmteunisse/fantaxtic")
# devtools::install_github("gmteunisse/ggnested")
library(tidyverse)
library(BiocManager)
library(phyloseq)
library(ggnested)
library(vegan)
library(matrixStats) # for 'colAnys'-function in'filter_species_by_group_prevalence'-function.
library(microViz)
library(fantaxtic) # for making barplots
library(gplots) # for enhanced heatmaps

NPERMUTATIONS <- 999
```

## Data: Loading and Filtering (relative-Abundance data)

```{r}
# load metaphlan percentage file (relative abundance file)
metaphlanData_percent <- read.table("data/raw/MF_SGB_percent.txt", header = TRUE, sep = "", row.names = 1)
# convert rownames into first column
metaphlanData_percent <- rownames_to_column(metaphlanData_percent, var = "clade_name")
metaphlanData_percent$clade_name <- as.character(metaphlanData_percent$clade_name)

# load metadata file
metaData <- read.table("../metadata/metadata_silas.csv", header = TRUE, sep = ",") %>%
    column_to_rownames(var = "id")


metaData <- metaData[!grepl("ppB", rownames(metaData)), ]

```

check metaphlan percentage file

```{r}
metaphlanData_percent_check <- metaphlanData_percent %>%
    select(-clade_name)

summary(colSums(metaphlanData_percent_check))
```

filter metaphlan data (all ppB) samples

```{r}
# metaphlanData percentage


```

## Phyloseq-Object

```{r}
meta_physeq_percent <- metaphlan_to_phyloseq(metaphlanData_percent, metaData, version = 4, verbose = TRUE, tax_lvl = "SGB")
ps_percent <- meta_physeq_percent
ps_percent <- subset_taxa(ps_percent, Kingdom != "k__Eukaryota") # remove eukaryotes


# check the ps-percent - object
ps_kingdom <- tax_glom(ps_percent, taxrank = "Kingdom")
otu_mat <- as(otu_table(ps_kingdom), "matrix")
# print(otu_mat)


# look if we have species without t__
tax_names <- taxa_names(ps_percent)
not_t_taxa <- tax_names[!grepl("t__", tax_names)]
print(not_t_taxa)


ps_taxa <- ps_percent

```


# Load AR genes

```{r}
# Load AR gene abundance data


ph_arg = readRDS("../Genecatalog/results/physeq_normalized.rds")

# filter AR gene abundance data to include only samples present in the metadata

ph_arg <- prune_samples(rownames(metaData), ph_arg)

# add fake taxonomy

tax_table(ph_arg) <- matrix(paste0("ARG_", seq_len(ntaxa(ph_arg))), ncol = 1, dimnames = list(taxa_names(ph_arg), "Variant"))

```

### Filter Phyloseq-Object based on minimal relative abundance

retain taxa whose average relative abundance across all samples exceeds $10^{-5}$.

```{r}

# percentage data
# 10^-5 = 0.001% relative abundance
ps_taxa_filt <- filter_taxa(
    ps_taxa,
    function(x) mean(x) > 1e-5,
    prune = TRUE
)

# counts per million
# 10 CPM = 0.001% relative abundance
ps_arg_filt <- filter_taxa(
    ph_arg,
    function(x) mean(x) > 10,
    prune = TRUE
)


print(ps_taxa_filt)

print(ps_arg_filt)

```


## CLR transformation

```{r}

# CLR transformation for taxa data
ps_taxa_clr <- tax_transform(ps_taxa_filt, trans = "clr",zero_replace="halfmin")

# CLR transformation for AR gene abundance data
ps_arg_clr <- tax_transform(ps_arg_filt, trans = "clr",zero_replace="halfmin",rank = "unique")

```








## Ordination analysis using microViz

### PCA (CLR, SGB)
```{r}
ord_taxa <- ord_calc(ps_taxa_clr, method = "PCA")
ord_arg <- ord_calc(ps_arg_clr, method = "PCA")



```

```{r}
tax_ord_plot <- ord_plot(ord_taxa, color = "category2", size = 2) +
  stat_ellipse(aes(color = category2)) +
  #    ggtitle("PCA (CLR, SGB)") +
  labs(color = "Category") +
  scale_color_manual(
    values = unlist(color_codes$category2),
    labels = color_codes$label_mapping
  )
plot(tax_ord_plot)

```


```{r}
arg_ord_plot <- ord_plot(ord_arg, color = "category2", size = 2) +
  stat_ellipse(aes(color = category2)) +
  #    ggtitle("PCA (CLR, AR)") +
  labs(color = "Category") +
  scale_color_manual(
    values = unlist(color_codes$category2),
    labels = color_codes$label_mapping
  )
plot(arg_ord_plot)



```



# Procrustes analysis

```{r}
Procrustes_Results = list()


run_procrustes <- function(ps_taxa_clr, ps_arg_clr, sample_index) {

    samples_in_both_datasets <- intersect(sample_names(ps_taxa_clr), sample_names(ps_arg_clr))

    sample_index_internal <- intersect(samples_in_both_datasets, sample_index)

    if (length(sample_index_internal) != length(sample_index)) {
        dropped_samples <- setdiff(sample_index, samples_in_both_datasets)


        warning(paste("Dropped samples in Procrustes analysis:", paste(dropped_samples, collapse = ", ")))
    }

    ps_taxa_clr |> ps_reorder(sample_index_internal) -> ps_taxa_clr
    ps_arg_clr |> ps_reorder(sample_index_internal) -> ps_arg_clr

    taxa_dist <- ps_taxa_clr |> otu_table() |> t() |> vegdist(method = "euclidean")
    arg_dist <- ps_arg_clr |> otu_table() |> t() |> vegdist(method = "euclidean")

    results <- vegan::protest(taxa_dist, arg_dist, permutations = NPERMUTATIONS)

    return(list(
        p_value = results$signif,
        correlation = results$t0
    ))
}

```


```{r}

setdiff(sample_names(ps_arg_clr), sample_names(ps_taxa_clr))

setdiff(sample_names(ps_taxa_clr), sample_names(ps_arg_clr)) #s24H1 missing in arg

```

## all samples taxa vs AR genes

```{r}

sample_index <- intersect(sample_names(ps_taxa_clr), sample_names(ps_arg_clr))



Procrustes_Results[["all_samples"]] <- run_procrustes(ps_taxa_clr, ps_arg_clr, sample_index)
```


## all samples, excluding Non_farmer_controls

```{r}
sample_index <- setdiff(intersect(sample_names(ps_taxa_clr), sample_names(ps_arg_clr)), rownames(metaData)[metaData$category2 == "Non_farmer_controls"])

Procrustes_Results[["excluding_non_farmer_controls"]] <- run_procrustes(ps_taxa_clr, ps_arg_clr, sample_index)


```


## subgroups

```{r}

# Run species/stage specific Procrustes analyses

for (species in unique(metaData$species)) {
    for (stage in c( "Fattening", "Suckling",  "Weaning")) {
        sample_index <- rownames(metaData)[metaData$species == species & metaData$growth_stage == stage]

        name= paste0(species, "_", stage)

        if (length(sample_index) > 1) {

            
            print(paste("Running Procrustes for:", name))

            Procrustes_Results[[name]] <- run_procrustes(ps_taxa_clr, ps_arg_clr, sample_index)
        } else {
            print(paste("Skipping Procrustes for:", name, "- not enough samples"))
        }
    }
}

```



## Procrustes results to dataframe

```{r}
procrustes_df <- do.call(rbind, lapply(names(Procrustes_Results), function(name) {
    data.frame(
        Group = name,
        P_value = Procrustes_Results[[name]]$p_value,
        Correlation = Procrustes_Results[[name]]$correlation
    )
}))
print(procrustes_df)


```


# Correlation between ARG and taxa


```{r}


# Get the CLR-transformed abundance matrices
common_samples <- setdiff(intersect(sample_names(ps_taxa_clr), sample_names(ps_arg_clr)), rownames(metaData)[metaData$category2 == "Non_farmer_controls"])

taxa_matrix <- ps_taxa_clr |> ps_reorder(common_samples) |> otu_table() |> t()
arg_matrix <- ps_arg_clr |> ps_reorder(common_samples) |> otu_table() |> t()

# Calculate pearson correlation, p value and adjust p values for multiple testing
# Pre-compute all pairwise Pearson correlations at once
cor_matrix <- cor(taxa_matrix, arg_matrix, method = "pearson")

# Compute p-values using vectorized correlation test
# Uses the t-statistic: t = r * sqrt((n-2)/(1-r^2))
n <- nrow(taxa_matrix)
t_stat <- cor_matrix * sqrt((n - 2) / (1 - cor_matrix^2))
p_matrix <- 2 * pt(abs(t_stat), df = n - 2, lower.tail = FALSE)

# Reshape both matrices to long format in one step
cor_results <- data.frame(
    Taxa             = rep(colnames(taxa_matrix), times = ncol(arg_matrix)),
    ARG              = rep(colnames(arg_matrix),  each  = ncol(taxa_matrix)),
    Correlation      = as.vector(cor_matrix),
    P_value          = as.vector(p_matrix),
    Adjusted_P_value = NA_real_
)

# Adjust p-values for multiple testing using Benjamini-Hochberg method
cor_results$Adjusted_P_value <- p.adjust(cor_results$P_value, method = "BH")

cor_results[cor_results$Adjusted_P_value < 0.1 & abs(cor_results$Correlation) > 0.5, ]



```





## Prcurstes, human vs pig,  species per farm

```{r}



Procrustes_Results2 = list()


for (stage in c( "Fattening", "Suckling",  "Weaning")) {


pig_samples <- rownames(metaData)[metaData$species == "Pig" & metaData$growth_stage == stage]

if (stage== "Weaning") {
    human_samples <- rownames(metaData)[metaData$species == "Human" & metaData$growth_stage == "Suckling"]
}else{
human_samples <- rownames(metaData)[metaData$species == "Human" & metaData$growth_stage == stage]
}


ps_taxa_clr  |>  ps_reorder(human_samples) |> otu_table() |> t() -> human_matrix
ps_taxa_clr  |> ps_reorder(pig_samples) |> otu_table() |> t() -> pig_matrix


# name by farm
rownames(human_matrix) <- metaData[human_samples, "farm"]
rownames(pig_matrix) <- metaData[pig_samples, "farm"]

farm_samples_intersect <- intersect(rownames(human_matrix), rownames(pig_matrix))


results <- vegan::protest(vegdist(human_matrix[farm_samples_intersect,], method = "euclidean"), vegdist(pig_matrix[farm_samples_intersect,], method = "euclidean"), permutations = NPERMUTATIONS)


Procrustes_Results2[[stage]] <- list(
    p_value = results$signif,
    correlation = results$t0
)

}

Procrustes_Results2
```


# pigs stage 1 per stage2


```{r}


for (stage1 in c( "Fattening", "Suckling",  "Weaning")) {
    for (stage2 in c( "Fattening", "Suckling",  "Weaning")) {
        pig_samples_stage1 <- rownames(metaData)[metaData$species == "Pig" & metaData$growth_stage == stage1]
        pig_samples_stage2 <- rownames(metaData)[metaData$species == "Pig" & metaData$growth_stage == stage2]

        pig_matrix_stage1 <- ps_taxa_clr |> ps_reorder(pig_samples_stage1) |> otu_table() |> t()
        pig_matrix_stage2 <- ps_taxa_clr |> ps_reorder(pig_samples_stage2) |> otu_table() |> t()


        # Rename rows to farm
        rownames(pig_matrix_stage1) <- metaData[pig_samples_stage1, "farm"]
        rownames(pig_matrix_stage2) <- metaData[pig_samples_stage2, "farm"]


        samples_intersect <- intersect(rownames(pig_matrix_stage1), rownames(pig_matrix_stage2))

        
        results <- vegan::protest(vegdist(pig_matrix_stage1[samples_intersect,], method = "euclidean"), vegdist(pig_matrix_stage2[samples_intersect,], method = "euclidean"), permutations = NPERMUTATIONS)

        Procrustes_Results2[[paste0("Pig_",stage1, "_vs_", stage2)]] <- list(
            p_value = results$signif,
            correlation = results$t0
        )

    }
}


```