---
title: "strain_analysis_6"
format: html
---

# Strain Analylsis 6

**clean scripts for detecting shared strains from Permanova analysis (2 groups; pig - farmers) suckling pigs exluded, **mrpp** analysis **\

## setup chunk

```{r}
#| include = FALSE


my_theme <- function() {
  theme_bw(base_family = "DejaVu Sans") +
    theme(
      plot.title = element_text(hjust = 0.5),
    )
}

library(ggplot2)
theme_set(my_theme())



```

## dependencies

```{r}
library(tidyverse)
library(utils)
library(archive)
library(R.utils) # to define and work with paths
library(purrr)
library(vegan) # betadispr
library(dplyr)
library(stringr)
library(tidyr) # for separate function
library(readr) # for reading and writing .csv-files
library(svglite) # to save plots as svg)

# fonts

library(systemfonts)
subset(system_fonts(), family == "DejaVu Sans")

```

**define file pathways**

```{r}
# define paths
dir_path <- "../data/raw/GD"
out_path <- "../data/processed/GD"
```

## load metadata

```{r}
metaData <- read.table("../../Metadata/metadata_silas.csv", header = TRUE, sep = ",") %>%
  column_to_rownames(var = "id")


sgb_annotation <- read.table("../../Metadata/sgb_annotation.csv", header = T, sep = ",")
```

## extract all distance matrices

```{r}

# make sure that out_path exists
if(!dir.exists(out_path)) dir.create(out_path)

# extract all files into processed folder
files_gz <- list.files(dir_path, pattern = "\\.tsv\\.gz$", full.names = TRUE)

for (f in files_gz) {
  out_file <- file.path(out_path, basename(sub("\\.gz$", "", f)))
  R.utils::gunzip(f, destname = out_file, overwrite = TRUE)

}

# load metadata

metaData <- read.table("../../metadata/metadata_silas.csv", header = TRUE, sep = ",") %>%
  column_to_rownames(var = "id")
```

**list all files the processed directory**

```{r}
files_tsv <- list.files(out_path, pattern = "\\.tsv$", full.names = TRUE)
print(length(files_tsv))
```


## mrpp-approach
```{r}
set.seed(123)
library(dplyr)
library(stringr)

# MRPP workflow 
res.mrpp_pig_worker <- map_dfr(files_tsv, function(f) {
  
  message("Processing file: ", f)

  # Matrix prep 
  m <- read.table(f, header = TRUE, sep = "\t")
  rownames(m) <- m[,1]; m <- m[,-1]
  m[] <- lapply(m, function(x) as.numeric(as.character(x))); m <- as.matrix(m)
  if (nrow(m) != ncol(m)) stop("Non-square matrix")
  colnames(m) <- rownames(m)
  
  # normalize the matrix
  # m <- m / max(m, na.rm = TRUE)
  
  # Metadata: ppB/pigs+farmers only
  metaData_clean <- metaData %>% rownames_to_column("id") %>%
    mutate(sample_strip = str_remove(id, "^C"))
  meta_sgb <- metaData_clean %>%
    filter(sample_strip %in% rownames(m)) %>%
    mutate(sample = sample_strip) %>%
    select(sample, farm, species, category, everything()) %>%
    filter(!str_detect(sample, "ppB")) %>%
    filter(category != "Non_farmer_controls") %>%
    mutate(Host = category) %>%  # Pigs vs Pig_farmers
    droplevels()
  
  Host_counts <- table(meta_sgb$Host)
  if (any(Host_counts < 5) || length(Host_counts) < 2) {
    message("⚠️ Skipping: need >=5/group")
    return(tibble(SGB = str_extract(basename(f), "SGB\\d+"), 
                  A_stat = NA, p_value = NA))
  }
  
  keep_samples <- intersect(meta_sgb$sample, rownames(m))
  m_sub <- m[keep_samples, keep_samples, drop = FALSE]
  meta <- meta_sgb[match(keep_samples, meta_sgb$sample), , drop = FALSE]
  meta$Host <- as.factor(meta$Host)
  dist_matrix <- as.dist(m_sub)
  
  # MRPP 
  mrpp_result <- mrpp(dist_matrix, grouping = meta$Host, permutations = 999)
  
  # Extract stats
  A_stat <- mrpp_result[["A"]]  # Chance-corrected within-group agreement
  p_value <- mrpp_result[["Pvalue"]]  # Permutation p-value
  
  n_sample_pigs <- sum(meta$Host == "Pigs")
  n_sample_farmers <- sum(meta$Host == "Pig_farmers")
  
  
  sgb <- str_extract(basename(f), "SGB\\d+")
  tibble(SGB = sgb, 
         A_stat = A_stat,      # >0 = stronger within-group (host-specific)
         p_value = p_value,    # <0.05 = sig within-group structure
         n_sample_pigs = n_sample_pigs,
         n_sample_farmers = n_sample_farmers)
})

# Interpretation
host_specific <- res.mrpp_pig_worker %>% 
  filter(A_stat > 0 & p_value < 0.05)  # Host-specific strains

shared_strains <- res.mrpp_pig_worker %>% 
  filter(A_stat <= 0 | p_value >= 0.05)  # Shared/no structure

```
#### barplot
```{r}
head(res.mrpp_pig_worker)

df_barplot_mrpp <- data.frame(
  host_specific = sum(res.mrpp_pig_worker$p_value < 0.05, na.rm= TRUE),
  non_host_specific = sum(res.mrpp_pig_worker$p_value >= 0.05,  na.rm = T),
  amount_NA = sum(is.na(res.mrpp_pig_worker$p_value))
  )%>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "count")

head(df_barplot_mrpp)

  barplot_mrpp <- ggplot(data = df_barplot_mrpp, aes(x = variable, y = count))+
  geom_bar(stat ="identity", width = 0.45, color = "black", fill = "lightblue")+
  theme_bw()+
  xlab("")+
  geom_text(
    aes(label = paste0("n = ",count)),
    vjust = 1.5,
    size = 5
  )+
    theme_publication()+
    theme(axis.text.x = element_text(angle = 25, hjust = 1, vjust = 1))+
    ggtitle("MRPP-Appraoch, SGB classification")


barplot_mrpp

# save:
ggsave("../figures/mrpp_n_sgb.svg", plot = barplot_mrpp, width = 6, height = 6, web_fonts = "DejaVu Sans")
```


## Individualization

### individualized with normalized distances

-   normalization by median of the matrix

-   calculation of all minimal distances to a next sample

-   plot the median of all SGBs to the nearest next strain

-   check distribution

```{r}
results_min_normalized <- map_dfr(files_tsv, function(f) {
# show which sample is processed
  message("Processing file: ", f)
  
  # extract and manipulate matrix
  m <- read.table(f, header = TRUE, sep = "\t")
  rownames(m) <- m[,1]
  # remove sample name column
  m <- m[,-1]
  
  # ensure numeric
  m[] <- lapply(m, function(x) as.numeric(as.character(x)))
  m <- as.matrix(m)
  
  # normalize matrix by maximal branch length
  m <- m/median(m, na.rm = TRUE) 
  
   # metadata: extract sample names
  metadata <- data.frame(
    sample = rownames(m)
  )

  # set row names the same as sample name
  row.names(metadata) <- metadata$sample
  
  
  # metadata: assign the group
  metadata <- metadata %>%
    mutate(group = if_else(str_detect(sample, "pp"), "pig",
                           if_else(str_detect(sample, "H"), "worker", "ctrl"))) %>%
    mutate(group_2 = if_else(str_detect(group, "pig"), "pig", "human"),
           is_ppB = str_detect(sample, "ppB"))%>%
    filter(!is_ppB)
  
  # turn group into factor
  metadata$group <- as.factor(metadata$group)
  metadata$group_2 <- as.factor(metadata$group_2)
  
  
  # filter: keep only groups with >=5 samples
  keep_groups <- names(which(table(metadata$group_2) >= 5))
  metadata <- droplevels(metadata[metadata$group_2 %in% keep_groups, , drop = FALSE])
  
  # subset the matrix based on keep_groups 
  if (nlevels(metadata$group_2) < 2) {
    message("Skipping ", f, " (not enough pig/human samples)")
    return(tibble(
      SGB = str_extract(basename(f), "SGB\\d+"),
      median_min = NA_real_
    ))
  }
  
  keep_samples <- rownames(metadata)
  
  keep_samples <- intersect(keep_samples, rownames(m))
  keep_samples <- intersect(keep_samples, colnames(m))
  
  if (length(keep_samples) < 2) {
    message("Skipping ", f, " (less than 2 samples after matching)")
    return(tibble(
      SGB = str_extract(basename(f), "SGB\\d+"),
      median_min = NA_real_
    ))
  }
  
   # final safety check (prevents subscript out of bounds)
  if (!all(keep_samples %in% rownames(m)) ||
      !all(keep_samples %in% colnames(m))) {
    message("Skipping ", f, " (mismatch between metadata and matrix names)")
    return(tibble(
      SGB      = str_extract(basename(f), "SGB\\d+"),
      median_min = NA_real_
    ))
  }

  
  m_sub <- m[keep_samples, keep_samples, drop = F]
  
  # remove diagonal, to ignore self
  diag(m_sub) <- NA
  min_per_sample <- apply(m_sub,1,min,na.rm = T)
  tibble(
      SGB = str_extract(basename(f), "SGB\\d+"),
      median_min = median(min_per_sample, na.rm = T)
    )
  
})
```

display density and save

```{r}
head(results_min_normalized)

density_min_SGB_normalized <- ggplot(data = results_min_normalized, aes(x = median_min))+
  geom_density()+
  labs(
    subtitle = "median distance to nearest other strain based on normalized distances" 
  )
print(density_min_SGB_normalized)

# ggsave
ggsave("../figures/median_minimal_SGB_distance.svg",
       plot = density_min_SGB_normalized,
       width = 6, 
       height = 6,
       device = svglite,
       web_fonts = "DejaVu Sans")
```

### pick a threshold

```{r}
quantile(results_min_normalized$median_min, probs = c(0.25,0.5, 0.75, 0.9, 0.95), na.rm = TRUE)

t <- median(results_min_normalized$median_min, na.rm = T)

t
```

classify the results in individualized vs. non_individualized

```{r}
results_min_classified <- results_min_normalized %>%
  mutate(individualized = median_min > t)

# check counts
table(results_min_classified$individualized, useNA = "ifany")
```

ggplot to see separation

```{r}
ggplot(results_min_classified, aes(x = median_min, fill = individualized)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = t, linetype = "dashed") +
  theme_bw() +
  labs(x = "median_min", title = "Mean nearest-neighbor nGD per SGB")

```


**take SGBs for inspection make four groups**

```{r}

res_4cat_mrpp <- results_min_classified %>%
  left_join(res.mrpp_pig_worker, by = "SGB") %>%
  mutate(
    host_specific = p_value < 0.05,
    category = case_when(
      host_specific &  individualized ~ "host_specific + individualized",
      host_specific & !individualized ~ "host_specific + non_individualized",
     !host_specific &  individualized ~ "non_host_specific + individualized",
     !host_specific & !individualized ~ "non_host_specific + non_individualized"
    )
  )

```

summary the amount per groups

```{r}
summary_res_4cat_mrpp <- dplyr::count(
  res_4cat_mrpp,
  category,
  name = "n"
) %>%
  filter(!is.na(category)) %>%
  mutate(
    category = factor(
      category,
      levels = c(
        "host_specific + individualized",
        "host_specific + non_individualized",
        "non_host_specific + individualized",
        "non_host_specific + non_individualized"
      )
    )
  ) %>%
  arrange(category) %>%
  mutate(pct = round(100 * n / sum(n), 1))

summary_res_4cat_mrpp


```

### Host type effect vs. individualization (boxplots)

```{r}
res_4cat_mrpp <- res_4cat_mrpp %>%
  filter(!is.na(category))


boxplots_perm_ind <- ggplot(res_4cat_mrpp, aes(x = category, y = A_stat, fill = category)) +
  geom_boxplot(outlier.alpha = 0.3) +
  geom_jitter(alpha = 0.5, width = 0.1)+
  my_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  ylab("MRPP A statistics (human vs pig)") +
  xlab("") +
  labs(title = "Host-type effect versus individualization")+
  scale_fill_manual(
    values=c("#1b85b8", "#5a5255", "#559e83", "#ae5a41"))+
  theme_publication()+
  theme(
    legend.position ="none",
    axis.text.x = element_text(angle = 37, hjust =1, size = 10)
  )

boxplots_perm_ind

# ggsave
ggsave("../figures/boxplot_mrpp_individualization.svg",
       plot = boxplots_perm_ind,
       width = 7,
       height = 6,
       device = svglite,
       web_fonts = "DejaVu Sans")
```

### R2 vs. mean min plot

```{r}
# remove NA
res_4cat_clean <- res_4cat_mrpp %>%
  na.omit()

# summary table
res_4cat_clean_summary <- res_4cat_clean %>%
  group_by(category)%>%
  summarise(
    n = n(),
    .groups = "drop"
  )

res_4cat_clean_summary

# labels vec
labels_vec <- with(res_4cat_clean_summary, setNames(
  paste0(category, " (n = ", n, ")"),
  category))

labels_vec 


#  scatterplot
scatter_perm_ind <- ggplot(data = res_4cat_clean, aes(x = median_min, y = A_stat, color = category))+
  geom_point()+
  labs(
    #title = "Scatterplot by Category"
  )+
  scale_color_manual(
    values=c("#1b85b8", "#5a5255", "#559e83", "#ae5a41"),
    labels = labels_vec)+
  geom_vline(xintercept = t, col = "red", linetype = "dashed")+
    ylab("MRPP A statistics")+
    xlab("minimal neighbour distribution")+
  theme_publication()+
  theme(
    legend.title = element_blank(),
    legend.position = c(0.73, 0.91),        # top-right inside plot
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA)
  )
  

scatter_perm_ind


```

```{r}
# ggsave
ggsave("../figures/scatterplot_mrpp_individualization.svg",
       plot = scatter_perm_ind,
       width = 10,
       height = 8,
       device = svglite,
       web_fonts = "DejaVu Sans")
```

## SGB Annotation

```{r}
df_annot <- res_4cat_clean %>%
  left_join(.,sgb_annotation, by = "SGB")

#check dataframe
amount_nonsig <- sum(res_4cat_clean$host_specific == F)
amount_nonsig

write.csv(df_annot, file = "../data/processed/sub-species_analysis_annot.csv", row.names = F)

```

### 4 categories (host specificity and individualization)

#### Phylum-Level

```{r}
df_phylum <- df_annot %>%
  group_by(category, Phylum) %>%
  summarise(
    n = n(),
    .groups = "drop"
  )


barplot_phylum <- ggplot(data = df_phylum, aes(x = Phylum, y = n, fill = category))+
  geom_col(position = "dodge", stat = "identity")+
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.9),
    vjust = -0.4,
    size = 5
  )+
  my_theme()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  scale_fill_manual(values=c("#1b85b8", "#5a5255", "#559e83", "#ae5a41", "grey"))+
  theme_publication()+
  theme(
    legend.title = element_blank(),
    legend.position = c(0.78, 0.85),        # top-right inside plot
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )+
  xlab("")

print(barplot_phylum)

#save
ggsave("../figures/4quadrant_mrpp_phylum.svg", plot = barplot_phylum, 
       width = 10,
       height = 8,
       device = svglite,
       web_fonts = "DejaVu Sans") 
```

#### Family-Level

```{r}
df_family <- df_annot %>%
  group_by(category, Family) %>%
  summarise(
    n = n(),
    .groups = "drop"
  )%>%
  filter(n>=2)


barplot_family <- ggplot(data = df_family, aes(x = Family, y = n, fill = category))+
  geom_col(position = "dodge", stat = "identity")+
  my_theme()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  scale_fill_manual(values=c("#1b85b8", "#5a5255", "#559e83", "#ae5a41", "grey"))

print(barplot_family)
```

### 2 categories (host specificity)

#### Phylum-Level

```{r}
df_phylum_hp <- df_annot %>%
  group_by(host_specific, Phylum) %>%
  summarise(
    n = n(),
    .groups = "drop"
  )%>%
  mutate(host_specific = if_else(host_specific == T, "host specific", "non-host specific"))


barplot_phylum_hp <- ggplot(data = df_phylum_hp, aes(x = Phylum, y = n, fill = host_specific))+
  geom_col(position = "dodge", stat = "identity")+
  my_theme()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  scale_fill_manual(values=c("#1b85b8", "#ae5a41"))

print(barplot_phylum_hp)
```

#### Family_Level

```{r}
df_family_hp <- df_annot %>%
  group_by(host_specific, Family) %>%
  summarise(
    n = n(),
    .groups = "drop"
  )%>%
  mutate(host_specific = if_else(host_specific == T, "host specific", "non-host specific"))%>%
  filter(n>=2)


barplot_family_hp <- ggplot(data = df_family_hp, aes(x = Family, y = n, fill = host_specific))+
  geom_col(position = "dodge", stat = "identity")+
  my_theme()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  scale_fill_manual(values=c("#1b85b8", "#ae5a41"))

print(barplot_family_hp)
```
## export data to csv
```{r}
write.csv(df_annot, "../data/processed/strain_analysis_annotations_6.csv", row.names = F)
```

## mrpp on farm level


## mrpp-approach: Within-farm pig similarity (comparing across farms)
```{r}
set.seed(123)

res.mrpp_pigs_within_farm <- map_dfr(files_tsv, function(f) {
  
  message("Processing file: ", f)
  sgb <- str_extract(basename(f), "SGB\\d+")

  # Matrix prep 
  m <- read.table(f, header = TRUE, sep = "\t")
  rownames(m) <- m[,1]; m <- m[,-1]
  m[] <- lapply(m, function(x) as.numeric(as.character(x))); m <- as.matrix(m)
  if (nrow(m) != ncol(m)) stop("Non-square matrix")
  colnames(m) <- rownames(m)
  
  # normalize the matrix
  m <- m / max(m, na.rm = TRUE)
  
  # Metadata: pigs only, grouped by farm
  metaData_clean <- metaData %>% rownames_to_column("id") %>%
    mutate(sample_strip = str_remove(id, "^C"))
  
  meta_sgb <- metaData_clean %>%
    filter(sample_strip %in% rownames(m)) %>%
    mutate(sample = sample_strip) %>%
    select(sample, farm, species, category, everything()) %>%
    filter(!str_detect(sample, "ppB")) %>%
    filter(category != "Non_farmer_controls") %>%
    filter(category == "Pigs") %>%  # PIGS ONLY
    droplevels()
  
  # Check: need >=2 pigs per farm
    farm_counts <- table(meta_sgb$farm)
    farms_with_2plus <- names(farm_counts[farm_counts >= 2])
  
    # Filter metadata to keep only farms with >=2 pigs
    meta_sgb <- meta_sgb %>% filter(farm %in% farms_with_2plus)
  
    # Now check if we have >=2 farms remaining
    if (nrow(meta_sgb) < 4 || length(unique(meta_sgb$farm)) < 2) {
      message("⚠️ Skipping: need >=2 farms with >=2 pigs each (after filtering)")
      return(tibble(SGB = sgb, A_stat = NA, p_value = NA, n_farms = NA))
    }
  
  keep_samples <- intersect(meta_sgb$sample, rownames(m))
  m_sub <- m[keep_samples, keep_samples, drop = FALSE]
  meta <- meta_sgb[match(keep_samples, meta_sgb$sample), , drop = FALSE]
  meta$farm <- as.factor(meta$farm)
  
  dist_matrix <- as.dist(m_sub)
  
  # Check for NaN/Inf/NA in distance matrix
  if (any(!is.finite(as.matrix(dist_matrix)))) {
    message("⚠️ Skipping: distance matrix contains non-finite values (NaN/Inf/NA)")
    return(tibble(SGB = str_extract(basename(f), "SGB\\d+"), 
                  A_stat = NA, p_value = NA, n_farms = NA))
  }
  
  # MRPP: pigs grouped by farm
  mrpp_result <- mrpp(dist_matrix, grouping = meta$farm, permutations = 999)
  
  # Extract stats
  A_stat <- mrpp_result[["A"]]
  p_value <- mrpp_result[["Pvalue"]]
  n_farms <- nlevels(meta$farm)
  
  tibble(SGB = sgb, 
         A_stat = A_stat,
         p_value = p_value,
         n_farms = n_farms)
})

# Significant within-farm pig clustering
pigs_farm_structured <- res.mrpp_pigs_within_farm %>% 
  filter(A_stat > 0 & p_value < 0.05)

head(res.mrpp_pigs_within_farm)
```
### p-valueplot
```{r}
mrpp_farm_level <- res.mrpp_pigs_within_farm %>%
  filter(!is.na(A_stat))%>%
  mutate(farm_specific = if_else(p_value < 0.05 & A_stat > 0, TRUE, FALSE))%>%
  rename(A_stat_farm_level = A_stat, p_value_farm_level = p_value)

ggplot(data = mrpp_farm_level, aes(y = A_stat_farm_level, x = p_value_farm_level, col = farm_specific))+
  geom_point(alpha = 0.5)
```
### barplot farm levle
```{r}
df_barplot_mrpp_farm <- data.frame(
  farm_specific = sum(res.mrpp_pigs_within_farm$p_value < 0.05, na.rm= TRUE),
  non_farm_specific = sum(res.mrpp_pigs_within_farm$p_value >= 0.05,  na.rm = T),
  amount_NA = sum(is.na(res.mrpp_pigs_within_farm$p_value))
  )%>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "count")

head(df_barplot_mrpp_farm)

barplot_mrpp_farm <- ggplot(data = df_barplot_mrpp_farm, aes(x = variable, y = count))+
geom_bar(stat ="identity", width = 0.3, color = "black", fill = "lightblue")+
theme_bw()+
xlab("")+
geom_text(
  aes(label = paste0("n = ",count)),
  vjust = 1.5,
  size = 3
)+
  theme_publication()+
  labs(
    title = "MRPP-Approach farm specificity"
  )+
  theme(
    axis.text.x = element_text(angle = 25, vjust = 1, hjust = 1)
  )

barplot_mrpp_farm

# ggsave
ggsave("../figures/barplot_mrpp_farm.svg", plot = barplot_mrpp_farm, 
       width = 6, height = 6, web_fonts = "DejaVu Sans")
```

## Questions for manuscript
merge with mrpp-result
```{r}
df_annot_final <- df_annot %>%
  left_join(mrpp_farm_level, by = "SGB")

## GB classification checks

# add derived flags used for the questions
df_gb <- df_annot_final %>%
  mutate(
    host_ambivalent = !host_specific,
    phylum_simple = if_else(is.na(Phylum) | Phylum == "", "Unknown", Phylum)
  )
```


### Q1: Bacteroidota show more host-ambivalent strains
```{r}
df_bact_host <- df_gb %>%
  filter(!is.na(host_ambivalent)) %>%
  mutate(is_bacteroidota = phylum_simple == "Bacteroidota")

tab_bact_host <- table(df_bact_host$is_bacteroidota, df_bact_host$host_ambivalent)
fisher_bact_host <- tryCatch(fisher.test(tab_bact_host), error = function(e) NA)
fisher_bact_host

df_plot <- df_bact_host %>%
  dplyr::count(is_bacteroidota, host_ambivalent, name = "n") %>%
  group_by(is_bacteroidota) %>%
  mutate(prop = n / sum(n)) %>%  # Proportions within each phylum
  ungroup()

plot_bact_host <- df_plot %>%
  ggplot(aes(x = is_bacteroidota, y = prop, fill = host_ambivalent)) +
  geom_col(position = "fill") +  # "fill" = 100% stacked
  geom_text(aes(label = paste0("n=", n)), 
            position = position_fill(vjust = 0.5), size = 3, 
            color = "white", fontface = "bold") +  # n-labels on proportions
  scale_x_discrete(labels = c(`TRUE` = "Bacteroidota", `FALSE` = "Other")) +
  scale_y_continuous(labels = scales::percent_format()) +  # % y-axis
  scale_fill_brewer(palette = "Set2", name = "Host-ambivalent") +
  labs(
    x = "Phylum", 
    y = "Proportion", 
    title = "Host Specificity Comparison: Bacteroidota vs Other",
    subtitle = paste("Fisher's exact test p =", round(fisher_bact_host$p.value, 3))
  ) +
  my_theme() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 12, face = "bold"))

print(plot_bact_host)
ggsave("../figures/bact_host_specificity_stacked.svg", plot_bact_host, 
       width = 8, height = 6, dpi = 300, web_fonts = "DeJaVu Sans")
```

### Q2: Bacteroidota are more farm-specific
```{r}
df_bact_farm <- df_gb %>%
  filter(!is.na(farm_specific)) %>%
  mutate(is_bacteroidota = phylum_simple == "Bacteroidota")

#fisher
tab_bact_farm <- table(df_bact_farm$is_bacteroidota, df_bact_farm$farm_specific)
fisher_bact_farm <- tryCatch(fisher.test(tab_bact_farm), error = function(e) NA)
fisher_bact_farm


df_plot_farm <- df_bact_farm %>%
  dplyr::count(is_bacteroidota, farm_specific, name = "n")%>%
  group_by(is_bacteroidota)%>%
  mutate(prop = n/sum(n)) %>%
  ungroup()

plot_bact_farm <- ggplot(data = df_plot_farm, aes(x = is_bacteroidota, y = prop, fill = farm_specific))+
  geom_col(position = "fill")+
  geom_text(aes(label = paste0("n= ", n)),
            position = position_fill(vjust = 0.5), size = 3,
            color = "black", fontface = "bold")+
  scale_x_discrete(labels = c(`TRUE`= "Bacteriodota", `FALSE` = "Other"))+
  scale_y_continuous(labels = scales::percent_format())+
  scale_fill_brewer(palette = "Set3", name = "Farm-specific")+
  labs(
    y = "Proportion of SGBs",
    x = "Phylum",
    title ="Farm Specific Comparison: Bacteroidota vs. Other", 
    subtitle = paste("Fisher's exact test p =", round(fisher_bact_farm$p.value,3))
  )+
  my_theme()+
  theme(legend.position = "bottom",
        plot.title = element_text(size = 12, face = "bold"))

print(plot_bact_farm)

# save this plot
ggsave("../figures/bact_farm_specific_stacked.svg", plot_bact_farm,
       width = 8, height = 6, dpi = 300, web_fonts = "DeJaVu Sans")
```

### Q2b: Farm specificity across phyla (absolute counts)
```{r}
# Absolute stacked barplots across multiple phyla
df_farm_base <- df_gb %>%
  filter(!is.na(farm_specific), !is.na(phylum_simple)) %>%
  mutate(
    farm_specific_label = if_else(farm_specific, "Farm-specific", "Non-farm-specific")
  )

# Keep top 4 phyla by total count; collapse remaining to "Others" (5 groups total)
top4_phyla_farm <- df_farm_base %>%
  dplyr::count(phylum_simple, name = "total_n") %>%
  arrange(desc(total_n)) %>%
  slice_head(n = 4) %>%
  pull(phylum_simple)

df_farm_phylum_abs <- df_farm_base %>%
  mutate(
    phylum_group = if_else(phylum_simple %in% top4_phyla_farm, phylum_simple, "Others")
  ) %>%
  dplyr::count(phylum_group, farm_specific_label, name = "n")

# Keep order: top 4 by abundance, then Others
phylum_order <- c(top4_phyla_farm, "Others")
df_farm_phylum_abs <- df_farm_phylum_abs %>%
  mutate(phylum_group = factor(phylum_group, levels = phylum_order))

# Precompute label positions for outside labels with connector lines
df_farm_phylum_abs <- df_farm_phylum_abs %>%
  mutate(farm_specific_label = factor(farm_specific_label,
                                      levels = c("Non-farm-specific", "Farm-specific")))

label_df_farm <- df_farm_phylum_abs %>%
  group_by(phylum_group) %>%
  arrange(desc(farm_specific_label), .by_group = TRUE) %>%
  mutate(
    y_center = cumsum(n) - n / 2,
    x_num = as.numeric(phylum_group)
  ) %>%
  ungroup()

# Labeling rule: low-count segments outside with connector lines
label_threshold <- 6
label_df_farm_low <- label_df_farm %>%
  filter(n > 0, n < label_threshold)
label_df_farm_high <- label_df_farm %>%
  filter(n >= label_threshold)

plot_farm_phylum_abs <- ggplot(
  df_farm_phylum_abs,
  aes(x = phylum_group, y = n, fill = farm_specific_label)
) +
  geom_col(position = "stack", color = "black", width = 0.55) +
  geom_segment(
    data = label_df_farm_low,
    aes(x = x_num + 0.27, xend = x_num + 0.36, y = y_center, yend = y_center),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 0.3
  ) +
  geom_text(
    data = label_df_farm_low,
    aes(x = x_num + 0.39, y = y_center, label = if_else(n > 0, as.character(n), "")),
    inherit.aes = FALSE,
    size = 3,
    color = "black",
    fontface = "bold",
    hjust = 0
  ) +
  geom_text(
    data = label_df_farm_high,
    aes(x = phylum_group, y = y_center, label = as.character(n)),
    inherit.aes = FALSE,
    size = 3,
    color = "black",
    fontface = "bold"
  ) +
  scale_fill_manual(
    values = c("Non-farm-specific" = "#2E86C1", "Farm-specific" = "#F28E2B"),
    name = "Farm Specificity"
  ) +
  labs(
    x = "Phylum",
    y = "Number of SGBs",
    #title = "Farm Specificity Across Phyla (Absolute Counts)"
  ) +
  my_theme() +
  coord_cartesian(clip = "off") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    plot.margin = margin(5.5, 35, 5.5, 5.5)
  )+
  theme_publication()+
  theme(
    legend.title = element_blank(),
    legend.position = c(0.75, 0.85),        # top-right inside plot
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )+
  xlab("")

print(plot_farm_phylum_abs)

ggsave(
  "../figures/farm_specificity_phyla_absolute_stacked.svg",
  plot = plot_farm_phylum_abs,
  width = 5,
  height = 6,
  dpi = 300,
  web_fonts = "DeJaVu Sans"
)
```

### Q2c: Host specificity across phyla (absolute counts)
```{r}
# Absolute stacked barplots across multiple phyla
df_host_base <- df_gb %>%
  filter(!is.na(host_specific), !is.na(phylum_simple)) %>%
  mutate(
    host_specific_label = if_else(host_specific, "Host-specific", "Non-host-specific")
  )

# Keep top 4 phyla by total count; collapse remaining to "Others" (5 groups total)
top4_phyla_host <- df_host_base %>%
  dplyr::count(phylum_simple, name = "total_n") %>%
  arrange(desc(total_n)) %>%
  slice_head(n = 4) %>%
  pull(phylum_simple)

df_host_phylum_abs <- df_host_base %>%
  mutate(
    phylum_group = if_else(phylum_simple %in% top4_phyla_host, phylum_simple, "Others")
  ) %>%
  dplyr::count(phylum_group, host_specific_label, name = "n")

# Keep order: top 4 by abundance, then Others
phylum_order_host <- c(top4_phyla_host, "Others")
df_host_phylum_abs <- df_host_phylum_abs %>%
  mutate(phylum_group = factor(phylum_group, levels = phylum_order_host))

# Precompute label positions for outside labels with connector lines
df_host_phylum_abs <- df_host_phylum_abs %>%
  mutate(host_specific_label = factor(host_specific_label,
                                      levels = c("Non-host-specific", "Host-specific")))

label_df_host <- df_host_phylum_abs %>%
  group_by(phylum_group) %>%
  # Match ggplot stacking order used in geom_col for correct label y-positions
  arrange(desc(host_specific_label), .by_group = TRUE) %>%
  mutate(
    y_center = cumsum(n) - n / 2,
    x_num = as.numeric(phylum_group)
  ) %>%
  ungroup()

# Labeling rule: low-count segments outside with connector lines
label_threshold <- 6
label_df_host_low <- label_df_host %>%
  filter(n > 0, n < label_threshold)
label_df_host_high <- label_df_host %>%
  filter(n >= label_threshold)

plot_host_phylum_abs <- ggplot(
  df_host_phylum_abs,
  aes(x = phylum_group, y = n, fill = host_specific_label)
) +
  geom_col(position = "stack", color = "black", width = 0.55) +
  geom_segment(
    data = label_df_host_low,
    aes(x = x_num + 0.27, xend = x_num + 0.36, y = y_center, yend = y_center),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 0.3
  ) +
  geom_text(
    data = label_df_host_low,
    aes(x = x_num + 0.39, y = y_center, label = if_else(n > 0, as.character(n), "")),
    inherit.aes = FALSE,
    size = 3,
    color = "black",
    fontface = "bold",
    hjust = 0
  ) +
  geom_text(
    data = label_df_host_high,
    aes(x = phylum_group, y = y_center, label = as.character(n)),
    inherit.aes = FALSE,
    size = 3,
    color = "black",
    fontface = "bold"
  ) +
  scale_fill_manual(
    values = c("Non-host-specific" = "#2E86C1", "Host-specific" = "#E31A1C"),
    name = "Host Specificity"
  ) +
  labs(
    x = "Phylum",
    y = "Number of SGBs",
    #title = "Host Specificity Across Phyla"
  ) +
  my_theme() +
  coord_cartesian(clip = "off") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    plot.margin = margin(5.5, 35, 5.5, 5.5)
  )+
  theme_publication()+
   theme(
    legend.title = element_blank(),
    legend.position = c(0.75, 0.85),        # top-right inside plot
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )+
  xlab("")
  

print(plot_host_phylum_abs)

ggsave(
  "../figures/host_specificity_phyla_absolute_stacked.svg",
  plot = plot_host_phylum_abs,
  width = 5,
  height = 6,
  dpi = 300,
  web_fonts = "DeJaVu Sans"
)
```

### Q3: Firmicutes are more individualized than Bacteroidota (excluding host-specific)
```{r}
df_indiv_compare <- df_gb %>%
  filter(host_specific == FALSE) %>%
  filter(!is.na(median_min)) %>%
  filter(phylum_simple %in% c("Firmicutes", "Bacteroidota"))

wilcox_indiv_phylum <- tryCatch(
  wilcox.test(median_min ~ phylum_simple, data = df_indiv_compare),
  error = function(e) NA
)
wilcox_indiv_phylum

p.val <- wilcox_indiv_phylum$p.value

plot_indiv_phylum <- ggplot(df_indiv_compare, aes(x = phylum_simple, y = median_min, fill = phylum_simple)) +
  geom_boxplot(outlier.alpha = 0.3, width = 0.3) +
  geom_jitter(width = 0.1, alpha = 0.5 )+
  my_theme() +
  annotate("text", x = 1.5, y = max(df_indiv_compare$median_min)*1.05,
           label = paste("Wilcoxon p=", round(p.val, 3)), size = 6)+
  labs(x = "", y = "Individualization score (median_min)",
       #title = "Individualization by phylum (non-host-specific)"
       ) +
  theme(legend.position = "none")+
  scale_fill_brewer()+
  theme_publication()+
  theme(
    legend.position = "none"
  )


print(plot_indiv_phylum)

# ggsave
ggsave("../figures/individualization_non_host_specific.svg", plot = plot_indiv_phylum, width = 6, height = 6, web_fonts = "DeJaVu Sans")
```

### Q4: Prevotella vs other Bacteroidota families
```{r}
df_prevotella <- df_gb %>%
  filter(phylum_simple == "Bacteroidota") %>%
  mutate(
    prevotella_group = case_when(
      !is.na(Genus) & str_detect(tolower(Genus), "prevotella") ~ "Prevotella",
      !is.na(Family) & str_detect(tolower(Family), "prevotell") ~ "Prevotella",
      TRUE ~ "Other_Bacteroidota"
    )
  )

tab_prev_host <- table(df_prevotella$prevotella_group, df_prevotella$host_ambivalent)
fisher_prev_host <- tryCatch(fisher.test(tab_prev_host), error = function(e) NA)
fisher_prev_host

tab_prev_farm <- table(df_prevotella$prevotella_group, df_prevotella$farm_specific)
fisher_prev_farm <- tryCatch(fisher.test(tab_prev_farm), error = function(e) NA)
fisher_prev_farm

wilcox_prev_indiv <- tryCatch(
  wilcox.test(median_min ~ prevotella_group, data = df_prevotella),
  error = function(e) NA
)
wilcox_prev_indiv

plot_prev_indiv <- ggplot(df_prevotella, aes(x = prevotella_group, y = median_min, fill = prevotella_group)) +
  geom_boxplot(outlier.alpha = 0.3, width = 0.3) +
  geom_jitter(alpha = 0.5, width = 0.1)+
  my_theme() +
  annotate("text", x = 1.5, y = max(df_indiv_compare$median_min)*1.05,
           label = paste("Wilcoxon p=", round(wilcox_prev_indiv$p.value, 3)), size = 6)+
  labs(x = "", y = "Individualization score (median_min)",
       #title = "Prevotella vs other Bacteroidota"
       ) +
  theme_publication()+
  theme(legend.position = "none")+
  scale_fill_brewer(palette = 4)

print(plot_prev_indiv)

# ggsave
ggsave("../figures/individualization_prevotella_vs_other_Bact.svg",
       plot = plot_prev_indiv,
       width = 6, height = 6,
       web_fonts = "DeJaVu Sans")
```



